### 2022-10-04 main.tex
Added figs, tex file, bib file

\documentclass[aps,superscriptaddress, notitlepage,longbibliography]{revtex4-1}

\usepackage{amsmath}
\usepackage{physics}
\usepackage[pdftex]{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{todonotes}
\hypersetup{
    colorlinks,
    citecolor=blue,
    filecolor=black,
    linkcolor=red,
    urlcolor=blue
}

\newcommand{\comment}[1]{{\color{blue}#1}}
\newcommand{\edit}[1]{{\color{purple}#1}}
\newcommand{\del}[1]{{\color{red}\st{#1}}}
\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

\begin{document}

\title{Single-cell Type Order Parameters (scTOP): A parameterless, non-stochastic algorithm for tracking cell identity using single-cell atlases}

\author{Maria Yampolskaya}
\email{mariay@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}

\author{Pankaj Mehta}
\email{pankajm@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\affiliation{Faculty of Computing and Data Science, Boston University, Boston, MA 02215, USA}
\affiliation{Biological Design Center, Boston University, Boston, MA 02215, USA}

\begin{abstract}
To elucidate the processes of cell-level phenotypic development and reprogramming, scientists are generating an abundance of single-cell RNA-sequencing (scRNA-seq) tissue atlases, lineage tracing studies, and other measurements. Using this high-dimensional, noisy data to identify cell type and understand the dynamics of cell fate transitions remains an ongoing problem. We introduce single-cell Type Order Parameters (scTOP), an epigenetic-landscape-inspired algorithm to track individual cells in the coordinate space of known cell types. Using existing human and mouse datasets that span across the body, we show scTOP reliably identifies cell type and traces the developmental progress of cells. scTOP visualizes the trajectory of lineage tracing experiments by directly comparing each timepoint state to the final fates. By tracking cells in cell type space, scTOP provides the technical foundation to understand in vivo developmental paths and possible bifurcations in the phenotypic landscape.
\end{abstract}

\maketitle

\section{Introduction}
Many adult animals exhibit hundreds of specialized phenotypic cell states, all of which arose from a small number of embryonic cells. In the early stages of development, embryonic cells possess the transcriptomic flexibility to become virtually any of the eventual cell fates. The questions of how and why cells end up in one fate or another is of interest both in biology and the physics of self-organizing systems. Understanding which cell fate transitions are possible is necessary to establish a theoretical framework for development. Understanding transitions is also important for developing directed differentiation protocols to engineer cells in vitro.

An indispensable way to probe cell type is by measuring the transcriptome. Capturing and quantifying the amounts of mRNA in a cell provides a snapshot of its phenotype. With the advent of single-cell RNA-sequencing (scRNA-seq), gene expression data of individual cells are being produced at unprecedented rates. Typical scRNA-seq measurements involve thousands of genes and dozens to millions of cells, resulting in high-dimensional data. Due to the technical challenges of sequencing RNA, the resulting data is noisy and sparse [Lahnemann]. Genes with relatively low expression are often recorded as not being expressed at all, resulting in zero-counts called “dropouts”. Experimental-to-experiment differences in sequencing method, platform, and other difficult-to-standardize conditions result in technical variations in scRNA-seq measurements called “batch effects.” These challenges merit careful consideration of how to normalize and analyze the data in a way that allows the study of biological variation without obfuscation by technical variation.

Three major computational approaches to examining cell fate transitions in scRNA-seq data are clustering, trajectory inference, and automatic cell identification. Clustering involves applying dimensional reduction (such as UMAP or SPRING [citations]) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells. Groups are then manually annotated according to expression levels of marker genes. Trajectory inference analysis arranges cells on a manifold and assigns pseudo-times according to transcriptomic similarity. There are also methods like partition-based graph abstraction [Wolf] that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions. Machine learning classifiers such as support vector machines are able to automatically identify cells with high accuracy [Abdelaal].

Although there are many useful computational tools, the common methods share certain limitations. Dimensional reduction is often one of the first steps to analyze scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships [Chari]. Distorted embeddings inconsistently represent distances between individual data points and between cell types, which makes it difficult to interpret cell fate transitions. This issue carries over to trajectory inference since nearest neighbors can vary substantially according to the embedding. Another limitation is the lack of reproducibility inherent in stochastic machine learning methods [Wattenberg] like t-SNE and UMAP, which produce different results whenever they are executed. Depending on exactly which data points are included, the output distribution may change significantly, even of data points that are retained from one iteration to another. The algorithm results also widely vary depending on input parameters such as perplexity. These parameters insert implicit assumptions about the locality of the data features. Since different papers use different parameters, each use of these algorithms applies different assumptions, rendering quantitative comparison between outputs impossible.

Ideally, an algorithm to explore cell fate transitions using scRNA-seq data would overcome these limitations by having explicit assumptions that can be kept consistent between analyses. Reproducibility would be further enhanced by a deterministic program. Currently, the axes of dimensional reduction outputs are difficult to interpret since they represent a complicated calculation of high-dimensional distance. Analysis would be improved if output axes were intuitive and remained meaningful from one algorithm iteration to another. This would require distances between points and clusters to remain undistorted after application. Meaningful axes could quantify cell type instead of assigning discrete identity labels, enabling numerical examination of transitions.

Our algorithm, single-cell Type Order Parameters (scTOP), reproducibly projects scRNA-seq data onto the dimensions of known cell types. Order parameters are variables whose values change as the system’s phase changes. Inspired by the epigenetic landscape and spin glass physics, scTOP treats cell types as phases and calculates the order parameters for transitions between them. The inputs into scTOP are the scRNA-seq measurements of query data as well as a reference basis of known cell types. The reference basis contains the archetypal gene expression profiles for the known cell types. scTOP outputs the coordinates of the query data in cell type space; each output dimension corresponds to a measure of similarity between the query data and one of the known cell types in the reference basis. Instead of reducing the data to two arbitrary dimensions, scTOP reduces the data to the number of provided cell types. With scTOP, it’s possible to visualize the path of a cell on axes measuring similarity to the most relevant cell types. The assumptions of the algorithm are explicit: that the known cell types are dynamical system attractors, and that the provided reference basis accurately represents the gene expression profiles of the known cell types. scTOP allows direct comparison between datasets, since the coordinates of and distances between data points do not change from application to application unless the reference basis is changed. scTOP is a robust, deterministic algorithm that creates consistent, intuitive axes on which to observe cell fate transitions.

In this paper, we show that scTOP reliably quantifies cell identity. It classifies cell type of individual cells with high accuracy, as demonstrated using mouse and human samples from lungs, brains, and organs throughout the body. By analyzing lineage tracing data and embryonic development data, we also establish scTOP’s capacity to quantify transitions between cell types.

We present scTOP in the easy-to-use forms of a Python package, available on the Python Package Index [link] and Github [link], and an R package, available on Github [link].

\section{Methods}
\begin{figure}
	\centering
		\includegraphics[scale=0.8]{figs/fig1.pdf}
	\caption{Steps involved in the scTOP algorithm. (a) Each sample in the query data is pre-processed independently of the other samples in the dataset. “Sample” may refer to either an individual cell sample or the average of cell samples corresponding to a tissue or population of interest. The former leads to individual cell analysis, and step 0 would be a matrix of single-cell RNA counts. The latter, which we call aggregate scoring, leads to tissue- or population- level analysis, and step 0 would involve a matrix where each “sample” column corresponds to an average of single-cell RNA counts. (b) Single-cell RNA-seq atlases, such as the Mouse Cell Atlas, are used to define the reference basis in the algorithm. (c) scTOP scores are the projections of query data onto the non-orthogonal subspace of cell types. Since there is no statistical fitting, no tuning parameters are involved. (d, e) show the process of finding scTOP scores, and the illustrated cell-type space has as many dimensions as there are cell types in the reference basis. We can visualize this concept in 1, 2, or 3 dimensions, depending on which cell types are most relevant. In the 3-dimensional case, the shadows on the bounds of the scatter plot are included to better illustrate the position of points in 3-dimensional space. (d) The process of finding scTOP scores for aggregate samples, which are averaged RNA counts for a population of interest. Note that the aggregate score tends to be higher than the average score of the individual cells (see 1e), since averaging over the RNA counts *before* scoring reduces the impact of noise. (e) The process of finding scTOP scores for individual cell samples.}
	\label{FIG:1}
\end{figure}

Explain inspiration: waddington landscape, attractor neural networks, cite previous papers (Dame et al., 2017; Ikonomou et al., 2020a; Lang et al., 2014; Pusuluri et al., 2018)

New part: application to single cells, allows analysis of individual variation, lineage tracing, cell-level fate transitions!

In pre-processing, the raw RNA-seq count data are first normalized, then z-scored. An explanation of the process and how it affects the data distributions is provided in Appendix section \ref{preprocessing}.

The way the data is pre-processed before applying the projection method is a vital component of the algorithm, as shown in figure \ref{FIG:1} (a). Each sample is pre-processed independently, unlike other algorithms which process data across samples to find, for example, the most variable genes. Depending on the particular application of scTOP, "sample" may refer to either an individual cell or the average expression levels of a subpopulation of cells. By processing each sample independently, the algorithm output for one sample is independent of which samples are analyzed alongside that one sample. scTOP involves one-shot learning, and there is no stochasticity involved; the algorithm will produce the same output regardless of the particular instance. There are no hyper-parameters involved in the training, which bypasses another source of bias. The only assumptions involved in the algorithm are explicit assumptions: the assumptions that the gene expression profiles included in the reference basis are truly representative of the relevant cell types, and that the cell types included are indeed distinct cell phenotypes.

Constructing a representative, accurate reference basis that is relevant to the data being analyzed is vital to the scTOP algorithm. This process is shown in figure \ref{FIG:1} (b). First, relevant single-cell RNA-seq atlases are gathered to be processed. For example, in analyzing mouse lung cells, we used the Mouse Cell Atlas \todo[inline]{Add reference to MCA}, which contains single-cell samples of mouse tissues across \todo[inline]{add \# of organs} organs. Once the relevant atlases are identified, we take an average across each cell population that corresponds to a distinct cell type. It is important that enough cells are included in each average, in order to mitigate the potent effects of noise. The minimum sample size to achieve reasonable results is 200 cells. Noise affects the algorithm most strongly when it appears in the basis, since the de-correlation operation involved in the projection separates cells according to the canonical expression levels included in the reference basis. \todo[inline]{include hill function discussion in appendix to explain how reference sample size affects scTOP score?} After each cell population is averaged to create the gene expression profiles for the desired cell types, each cell type profile is pre-processed separately. This creates a modular reference basis where cell types can easily be replaced, removed, or added at will.

Although they were originally inspired by attractor neural networks, the de-correlated magnetizations can also be understood as non-orthogonal projections onto the subspace of cell type space. This idea is shown in figure \ref{FIG:1} (c). A single-cell RNA-seq measurement of a sample produces a list of values corresponding to the expression levels of different genes, and thus can be conceptualized as a vector in $G$-dimensional gene expression space, where $G$ is the number of genes. Single-cell RNA-seq atlases provide reference gene expression profiles for hundreds of known cell types; these reference profiles also exist as vectors in gene expression space. If we compile a reference basis consisting of reference profiles corresponding to cell types, these vectors span a $C$-dimensional subspace of gene expression space, where $C$ is the number of cell types in the basis. There are far more genes than cell types: $C < G$. We can project the measured sample vector onto the cell type subspace in order to find the coordinates in cell type space, which enables identification of individual cells and tissues. Since the cell type profiles are largely correlated, they form a non-orthogonal subspace, and the projection onto this space is a non-orthogonal projection.

We can write the sample vector as a sum of the projected components and the component perpendicular to the cell type subspace: $S_i = \sum_\mu a^{\mu} \xi^{\mu}_i + S^{\perp}_i$. $S_i$ is the expression level of the $i$th gene in the measured sample. $a^{\mu}$ is the projection of the sample vector onto cell type $\mu$. $\xi^{\mu}_i$ is the expression level of the $i$th gene in the reference profile of cell type $\mu$. $S^{\perp}_i$ is the expression level of the $i$th gene of the component of the sample vector that is perpendicular to the cell type subspace. $S^{\perp}$ is the part of the sample vector that is not explained by the known cell types. It results from biological processes such as cell cycle variation and non-biological sources such as scRNA-seq noise.

There are two types of scTOP scores: aggregate and individual. The aggregate scoring method is useful for analyzing identified sub-populations of data at once. The scores tend to be higher, since aggregating gene expression data decreases the effects of noise. In aggregate scoring, the raw RNA count data is averaged across each sub-population, and then these average gene expression profiles are pre-processed and projected onto the reference basis. See figure \ref{FIG:1} (d). The resulting scTOP score provides the location of the sub-population in the cell type dimensions specified by the reference basis, which can be visualized in one, two, or three dimensions by selecting one, two, or three relevant cell types. In individual scoring, as shown in figure \ref{FIG:1} (e), each individual cell is pre-processed and projected onto the reference basis separately. Due to the well-documented effects of scRNA-seq noise, the resulting scores tend to be lower. Nonetheless, the resulting scores are able to accurately and reproducibly identify and track individual cells through cell type space, in whatever number of dimensions/cell types one chooses.

\section{Results}

\subsection{Benchmarking and validation}

\subsubsection{Classification of cell fate}
scTOP scores reliably and accurately identify the cell types of individual cells as well as of clusters of cells. To show this, we applied the algorithm to several scRNA-seq datasets across species and across laboratories. Table \ref{table:1} shows the data sets analyzed with the scTOP score accuracies of each. "TopN" refers to the percent of cells whose true cell types were in the top "N" scTOP scores, with a score greater than 0.1. "Unspecified" refers to the percent of cells for which the highest scTOP score was less than 0.1; this represents a failure of the algorithm to confidently identify the cell. The accuracies across species and tissues are high and the unspecified percentages are very low, even in the dataset exceeding a million cells. \todo[inline]{is it worth discussing sequencing method/effect of UMIs?}

\begin{table}[h!]
\centering
\begin{tabular}{| c | c | c | c | c | c | c |}
\hline
Query Data & Species & Reference Basis & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Total cells \\ 
\hline
Kotton lab & Mouse & Mouse Cell Atlas & 97.19 & 99.77 & 2.69 & 4,805 \\ 
\hline
Brain atlas & Mouse & Self & 86.28 & 96.89 & 1.78 & 1,161,041 \\
\hline
CellBench & Human & Self & 100 & 100 & 0 & 2,822 \\
\hline
Lung atlas & Human & Self & 79.78 & 85.16 & 3.32 & 2,952 \\  
\hline
\hline
Data set & Species & Reference Basis & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Number of cell types \\
\hline
Mouse Cell Atlas & Mouse & Tabula Muris & 79.17 & 87.5 & 0 & 48 \\
\hline

\end{tabular}
\caption{Accuracy scores of included sample types for each the described data sets.}
\label{table:1}
\end{table}

scTOP works across labs/different sequencing conditions.

The Kotton lab dataset \todo[inline]{add reference to paper, GEO accession number, etc.} measures gene expression levels of mouse lung tissue samples for \todo[inline]{add \# of different cell types}. See figure \ref{FIG:2} (a). The reference basis used to find the individual scTOP scores for this data set was derived from the Mouse Cell Atlas. Despite the fact that the Kotton lab dataset and the Mouse Cell Atlas were created in different laboratories and different sequencing methods, the resulting scTOP scores distinctly separate the cells from the Kotton laboratory. The resulting clusters agree with the annotations of the true cell type provided by the Kotton laboratory.

The mouse brain atlas identified \todo[inline]{add \# of cell types} cell types in over one million cells. For a reference basis, we reserved a subset of 200 cells from each of the identified types to use as training data. The test data, from which the accuracy score was calculated, was the remaining cells of each type not included in the training data. Ten different cell types are visualized in a grid in figure \ref{FIG:2} (b). The color of each grid square corresponds to the average scTOP score of cells annotated as the type on the y-axis, with the cell type corresponding to the scTOP dimension on the x-axis. As shown by the distinct blue diagonal, the average scTOP score correctly matches the corresponding annotated cell type.

scTOP correctly identifies individual cells, and it can also identify the tissue to which a population of cells belongs. As shown in \ref{FIG:2} (c), we found the aggregate scTOP scores for cells from the Mouse Cell Atlas projected onto the reference basis of Tabula Muris organs \todo[inline]{add ref for Tabula Muris}. Most cell types were correctly and strongly identified with the annotated tissue of origin. Parenchymal cell types were the most associated with the correct organ, while stromal cells were sometimes misidentified -- for example, the mammary gland endothelial cells. This is consistent with the fact that stromal cells are less specified to each organ. \todo[inline]{add analysis of stromal vs parenchymal? parenchyma alone?}

\begin{figure}
	\centering
		\includegraphics[scale=0.8]{figs/fig2.pdf}
	\caption{scTOP identifies mouse cell types. (a) Using the Mouse Cell Atlas as the reference basis, we show that mouse lung types separate clearly on scTOP score axes. The data points correspond to individual cells, and the marker shape and color indicate the "true type" as determined by annotations from [citation]. The axes used are scTOP scores for similarity with lung ciliated cells and alveolar types I and II. We also plotted the scTOP scores for the same data compared to random, irrelevant cell types, showing that the scores are tightly distributed around zero, as expected. (b) A subset of cells from the Mouse Brain Atlas are used as the reference basis to query other cells from the same data set. The shade of each bin corresponds to the average scTOP score for individual cells of the type indicated on the y-axis, compared to the reference types on the x-axis. The diagonal indicates that scTOP accurately matches query cells to the true reference type. (c) Aggregate samples from the Mouse Cell Atlas are compared to a reference basis consisting of organs from Tabula Muris. The scTOP scores are significantly higher when comparing a cell type with the organ of origin. This coarse-grain identification shows scTOP is able to classify tissues at different resolutions.}
	\label{FIG:2}
\end{figure}

The CellBench dataset \todo[inline]{add more info abt cellbench, and reference} consists of five separate cancer lines, designed to benchmark scRNA-seq analysis methods. As for the mouse brain atlas data, 200 cells from each type were used as training data to create the reference basis. As shown in figure \ref{FIG:3} (a) and table \ref{table:1}, scTOP identifies these five phenotypes with 100 percent accuracy.

The human lung atlas dataset ... \todo[inline]{add more info + reference} The reference basis was also created from a subset of cells from the same dataset.

\begin{figure}
	\centering
		\includegraphics[scale=0.8]{figs/fig3.pdf}
	\caption{scTOP correctly identifies human tissues. (a) The CellBench data set contains human cell cancer lines specifically processed to serve as benchmarking for scRNA-seq algorithms. We take a subset of cells to use as the reference basis in analyzing the rest of the CellBench data. The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis. The y-axis lists the true types of the queried cells. The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types. (b) Similar to (a), cells from the Human Lung Atlas are compared to a reference basis constructed from cells from the Human Lung Atlas. The colors of the bins correspond to average scTOP scores of individual cells. Again, the diagonal is the most prominent section, showing that cells of the y-axis types are correctly matched to the reference x-axis identities.}
	\label{FIG:3}
\end{figure}

\subsubsection{Tracking cell type transitions}
The previous section showed that scTOP accurately tracks the developed cell type of individual cells and clusters of cells. Now that we have shown that it correctly identifies the coordinates in cell type space, we can apply to developing or transitioning cells to see dynamics in cell type space.

The atlas of lung development contains cells from \todo{add ref} embryonic mice at days 12 through 18 post-conception. They are annotated by cell type, so we can track their differentiation in cell type space using scTOP scores. Figure \ref{FIG:4} (a) shows the aggregate scores of alveolar type I (AT1) and alveolar type II cells (AT2) as the days progress. The x-axis shows the scTOP score for AT1, the y-axis shows AT2, and each point is colored by the scTOP score for early epithelial cell type. The AT1 cells are marked with circles and the AT2 cells are marked with "X"s. It is clear to see that the AT1 cells separate from the AT2 cells early in development, and each one progresses on its separate track. Because scTOP provides us with reference axes that don't change according to whatever data is included, it is possible to track different data points across time on the same axes. Also, these axes give an exact interpretation: they measure the similarity to a canonical known cell type.

The hematopoietic dataset ... \todo{add ref} scRNA-seq destroys the cell it measures, but by barcoding sister cells, this dataset allows tracing of the lineage of each clone family. Figure \ref{FIG:4} (c) shows 2D histograms of every single cloned cell, colored according to the most likely time for each of the bins (see appendix for details \todo{add info in appendix}). From these histograms, one can see that some cell types are mutually suppressive while others form a continuous spectrum. For example, megakaryocytes and monocytes form a distinct L-shape that indicates the development program for each of these types prevents the program of the other; cells are either megakaryocytes, monocytes, or neither, there is no in-between stage. However, the plot for monocytes and neutrophils tells a different story: instead of having a distinct L-shape with narrow distribution along the axes, this plot shows a continuous spectrum of cells that are monocyte-like, neutrophil-like, and every step in-between. \todo{separate undiff plots}

Day 2 undifferentiated cells without clones and day 6 differentiated cells without clones were used to form the reference basis.

(move to appendix: we excluded erythrocytes and (?) because only one marker gene was used to annotate the original data, which caused a lot of noise since using only one marker gene does not sufficiently isolate the red blood cell population)

\begin{figure}
	\centering
		\includegraphics[scale=0.8]{figs/fig4.pdf}
	\caption{Beyond bench-marking: tracing development in cell-type space with scTOP. (a) scTOP illustrates specification of alveolar type I and type II cells in murine embryonic development. Alveolar cells from the Atlas of Lung Development are compared to a reference basis constructed from adult lung cells from [Kotton lab citation] and early epithelial cells from the Atlas of Lung Development. The labels next to each data point indicate the age of the mouse embryo from which lungs were extracted, from post-conception day 12 (E12) to day of birth (P0). The points with O-shaped (X-shaped) markers are alveolar type I (II) cells. The x-axis (y-axis) corresponds to the scTOP score for alveolar type I (II) identities. The shade of each point corresponds to the scTOP score for "early epithelium" identity. (b), (c) scTOP shows cell fate specification of hematopoietic cells using lineage tracing data from [citation]. }
	\label{FIG:4}
\end{figure}

\subsubsection{Finding genes important for determining cell types (marker genes?)}
using predictivity to find which cell types are highly correlated with which genes

\section{Discussion}

\section*{Acknowledgments}

Boston University Kilachand Multicellular Design Program NIH NIGMS 1R35GM119461

Simons Investigator in the Mathematical Modeling of Living Systems (MMLS) award to P.M.

\section*{Appendix}
\beginsupplement
\subsection{Pre-processing} \label{preprocessing}

\begin{figure}
	\centering
		\includegraphics[scale=0.6]{figs/fig1a supplement.pdf}
	\caption{Data distributions at each pre-processing step}
	\label{FIG:supp1}
\end{figure}

\subsection{Details on how the reference bases were constructed}
Describe how MCA, TM, and subsets of data were used. number of cells taken.

\subsection{Interpreting scores}
What scores are "high enough" or low, according to aggregate or individual

What a negative score means

How accuracies were calculated (what "unspecified" means)

\subsubsection{scTOP scores for irrelevant cell types}
Using same data as some of the figures in the main paper (ones that don't use the heatmap, since that already shows the irrelevant type scores are low), show that the false positive rate is very low

\subsection{Robustness of reference basis}
Show that the scores don't change significantly unless the reference basis is very bad. Show what happens when cell types are combined

\subsubsection{Cases where it doesn't work}
All the cases where scTOP fails are expected; they are where the underlying assumption of "reference basis = attractor basins" is false

Stromal vs parenchyma (different levels of specialization)

When the reference basis is ill-defined (e.g. early in embryonic development, when cells aren't well-differentiated) (or when not enough cells are used for the archetypal gene expression profiles)

\bibliography{export-data.bib}

\end{document}

### 2022-10-13 main.tex
updated to first draft

@@
 \usepackage{hyperref}
+\setlength {\marginparwidth }{2cm}
 \usepackage{todonotes}
@@
 Inspired by the epigenetic landscape and spin glass physics, scTOP treats cell types as phases and calculates the order parameters for transitions between them.
+It builds on previous work applying attractor-based order parameters to bulk RNA-sequencing [citations].
 The inputs into scTOP are the scRNA-seq measurements of query data as well as a reference basis of known cell types.
@@
 
-We present scTOP in the easy-to-use forms of a Python package, available on the Python Package Index [link] and Github [link], and an R package, available on Github [link].
+We present scTOP in the easy-to-use forms of a Python package, available on the Python Package Index [link] and Github [link], and an R package, available on Bioconductor [link] and Github [link].
 
@@
 
-Explain inspiration: waddington landscape, attractor neural networks, cite previous papers (Dame et al., 2017; Ikonomou et al., 2020a; Lang et al., 2014; Pusuluri et al., 2018)
-
-New part: application to single cells, allows analysis of individual variation, lineage tracing, cell-level fate transitions!
+To calculate cell type similarity scores, scTOP uses a linear algebra projection method inspired by the concept of cell types as attractors in a dynamical system [cite Huang et al.].
+This projection method has been applied previously to bulk RNA-seq samples [citations: Lang et al., 2014; Dame et al., 2017; Pusuluri et al., 2018; Ikonomou et al., 2020a], and we will summarize the theoretical background here.
+A cell type is a state of the gene regulatory network, so a natural way to represent attractors in this system is with an attractor neural network.
+If we represent each gene as a node in a network with an associated value measuring gene expression (positive value representing high gene expression, low value indicating low gene expression), then a cell type may be denoted by a vector of gene expression values.
+Cell types are often highly similar in their patterns of gene regulation.
+Kanter and Sompolinsky [citation] defined a pattern storage method for spin-glass-like neural networks where even correlated patterns robustly act as attractors.
+For the same system, they also defined order parameters $a^{\mu}$ which are de-correlated versions of the conventional spin glass order parameter, magnetization.
+$a^{\mu}$ measures the similarity between a given network state and the network state $\mu$, which is a stored pattern (in our case, a gene regulatory pattern that corresponds to a cell type).
+As [Lang et al.] have shown, the order parameters $a^{\mu}$ provided an excellent similarity score for analyzing bulk RNA-seq data.
+However, the cells digested by bulk RNA-seq may be in different stages of differentiation, and bulk processing renders it impossible to measure cell fate transitions beyond an average tissue state.
+scTOP uses the same order parameters, $a^{\mu}$, to track cell type at the resolution of individual cells, making it possible to directly observe cells in different stages of differentiation.
 
@@
 
-Although they were originally inspired by attractor neural networks, the de-correlated magnetizations can also be understood as non-orthogonal projections onto the subspace of cell type space.
+Although they were originally inspired by attractor neural networks and spin glass physics, the order parameters $a^{\mu}$ can also be understood as non-orthogonal projections onto the subspace of cell type space.
 This idea is shown in figure \ref{FIG:1} (c).
@@
 Due to the well-documented effects of scRNA-seq noise, the resulting scores tend to be lower.
-Nonetheless, the resulting scores are able to accurately and reproducibly identify and track individual cells through cell type space, in whatever number of dimensions/cell types one chooses.
+Nonetheless, the resulting scores are able to accurately and reproducibly identify and track individual cells through cell type space, in whatever number of dimensions (i.e.
+cell types) one chooses.
+
+The crux of scTOP is choosing a good reference basis.
+scTOP scores are reliable when the input reference basis incorporates many cells and represents a wide, relevant set of specialized cell types.
+Since scRNA-seq data is noisy and dropout is prevalent, many cells of each reference cell type must be included in order to create representative gene expression profiles.
+The choice of cell types in the basis is vital, too.
+The main underlying assumption of scTOP is that the reference cell types accurately represent attractors of the gene regulatory network.
+If the cell types aren’t well-differentiated from one another, as in early embryonic development or certain stromal cells when compared between organs, the resulting scTOP scores may be unreliable; they may change drastically when the basis is only slightly changed, or they may be incoherent.
+More information on the limitations of scTOP may be found in appendix [insert appendix num.].
+
+The accuracy of cell type information that scTOP can extract from scRNA-seq is limited by the low signal-to-noise ratio.
+As shown in appendix [insert appendix num.], the scTOP scores for false identities form a Gaussian distribution around zero.
+These distributions give an estimate of error.
+In aggregate scTOP scoring, RNA counts are averaged across a population and the resulting score is generally higher for relevant cell types than the average individual scTOP score for the same population.
+This is because the averaging step in aggregate scoring mitigates the effects of sparsity and noise in the RNA count matrices.
 
@@
 
-\subsection{Benchmarking and validation}
-
-\subsubsection{Classification of cell fate}
+\subsection{Benchmarking and validation: Classification of cell fate}
 scTOP scores reliably and accurately identify the cell types of individual cells as well as of clusters of cells.
 To show this, we applied the algorithm to several scRNA-seq datasets across species and across laboratories.
-Table \ref{table:1} shows the data sets analyzed with the scTOP score accuracies of each.
+scTOP is organism-agnostic as long as an appropriate reference basis is used.
+Mice and humans are among the most common subjects of scRNA-seq measurements.
+Many atlases across organs and stages of development have been created for both species.
+We used some of these atlases, such as the Mouse Cell Atlas [citation] and the Atlas of Lung Development [citation], as reference bases to analyze datasets from various sources.
+
+Table \ref{table:1} lists each of the datasets examined in this paper with their respective scTOP score accuracies.
 "TopN" refers to the percent of cells whose true cell types were in the top "N" scTOP scores, with a score greater than 0.1.
@@
 The accuracies across species and tissues are high and the unspecified percentages are very low, even in the dataset exceeding a million cells.
-\todo[inline]{is it worth discussing sequencing method/effect of UMIs?}
 
@@
 
-scTOP works across labs/different sequencing conditions.
-
-The Kotton lab dataset \todo[inline]{add reference to paper, GEO accession number, etc.} measures gene expression levels of mouse lung tissue samples for \todo[inline]{add \# of different cell types}.
+To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from Herriges et al.
+[citation] onto the mouse cell atlas.
+The Herriges et al.
+b dataset  measures gene expression levels of six different specialized lung types from mice used as control samples.
 See figure \ref{FIG:2} (a).
@@
 
-The mouse brain atlas identified \todo[inline]{add \# of cell types} cell types in over one million cells.
-For a reference basis, we reserved a subset of 200 cells from each of the identified types to use as training data.
+The mouse brain atlas\cite{yao_taxonomy_2021} sequenced over one million cells and clustered them into 42 subclasses and 101 supertypes.
+For a reference basis, we reserved a subset of 200 cells from each of the identified subclasses to use as training data.
 The test data, from which the accuracy score was calculated, was the remaining cells of each type not included in the training data.
@@
 scTOP correctly identifies individual cells, and it can also identify the tissue to which a population of cells belongs.
-As shown in \ref{FIG:2} (c), we found the aggregate scTOP scores for cells from the Mouse Cell Atlas projected onto the reference basis of Tabula Muris organs \todo[inline]{add ref for Tabula Muris}.
+As shown in \ref{FIG:2} (c), we found the aggregate scTOP scores for cells from the Mouse Cell Atlas projected onto the reference basis of Tabula Muris organs \cite{schaum_single-cell_2018}.
 Most cell types were correctly and strongly identified with the annotated tissue of origin.
@@
 This is consistent with the fact that stromal cells are less specified to each organ.
-\todo[inline]{add analysis of stromal vs parenchymal?
-parenchyma alone?}
+A discussion of the accuracy of scTOP when applied to stromal types versus parenchymal types may be found in appendix \ref{failure_cases}.
 
@@
 	\centering
-		\includegraphics[scale=0.8]{figs/fig2.pdf}
+		\includegraphics[scale=0.7]{figs/fig2.pdf}
 	\caption{scTOP identifies mouse cell types.
 (a) Using the Mouse Cell Atlas as the reference basis, we show that mouse lung types separate clearly on scTOP score axes.
-The data points correspond to individual cells, and the marker shape and color indicate the "true type" as determined by annotations from [citation].
+The data points correspond to individual cells, and the marker shape and color indicate the "true type" as determined by annotations from Herriges et al.
 The axes used are scTOP scores for similarity with lung ciliated cells and alveolar types I and II.
@@
 
-The CellBench dataset \todo[inline]{add more info abt cellbench, and reference} consists of five separate cancer lines, designed to benchmark scRNA-seq analysis methods.
-As for the mouse brain atlas data, 200 cells from each type were used as training data to create the reference basis.
-As shown in figure \ref{FIG:3} (a) and table \ref{table:1}, scTOP identifies these five phenotypes with 100 percent accuracy.
+The CellBench dataset \cite{tian_benchmarking_2019} was designed specifically to benchmark scRNA-seq analysis methods.
+We used the subset of data pertaining to five cancer lines.
+200 cells from each line were used as training data to create the reference basis, then the rest of the cells were input as query data.
+As shown in figure \ref{FIG:3} (a) and table \ref{table:1}, scTOP classifies cell line identity of the test data with 100 percent accuracy.
 
-The human lung atlas dataset ... \todo[inline]{add more info + reference} The reference basis was also created from a subset of cells from the same dataset.
+To further prove the power of scTOP in classifying human samples, we analyzed data from the human lung atlas dataset \cite{travaglini_molecular_2020}.
+This dataset sequenced thousands of human lung cells and identified 58 phenotypic popoulations.
+For our analysis, we restricted the training and test data to epithelial and stromal types that had at least 200 cells in each type population.
+The reference basis was created using 80\% of the cells from each population, and the query data consisted of the remaining 20\% of the data.
+Figure \ref{FIG:3} (b) shows that the scTOP scores were consistently high in cases where the score type matched the true type of the query.
 
@@
 
-\subsubsection{Tracking cell type transitions}
+As demonstrated by the diverse data sources examined here, as long as a well-defined reference basis is available, scTOP works well across different measurement conditions, species, organs, and resolutions.
+
+\subsection{Tracking cell type transitions}
 The previous section showed that scTOP accurately tracks the developed cell type of individual cells and clusters of cells.
-Now that we have shown that it correctly identifies the coordinates in cell type space, we can apply to developing or transitioning cells to see dynamics in cell type space.
+Now that we have shown that it correctly identifies the coordinates in cell type space, we can apply it to developing or transitioning cells to see dynamics in cell type space.
 
-The atlas of lung development contains cells from \todo{add ref} embryonic mice at days 12 through 18 post-conception.
+The atlas of mouse lung development\cite{negretti_single-cell_2021} contains cells from day 12 embryonic mice to day 14 postnatal mice.
 They are annotated by cell type, so we can track their differentiation in cell type space using scTOP scores.
@@
 
-The hematopoietic dataset ...
-\todo{add ref} scRNA-seq destroys the cell it measures, but by barcoding sister cells, this dataset allows tracing of the lineage of each clone family.
-Figure \ref{FIG:4} (c) shows 2D histograms of every single cloned cell, colored according to the most likely time for each of the bins (see appendix for details \todo{add info in appendix}).
-From these histograms, one can see that some cell types are mutually suppressive while others form a continuous spectrum.
+The hematopoietic dataset\cite{weinreb_lineage_2020} represents thousands of hematopoietic stem cells and their cloned sister cells in various states of differentiation.
+scRNA-seq destroys the cell it measures, but by barcoding cloned cells, this dataset traces the lineage of each clone family.
+Day 2 undifferentiated cells without clones and day 6 differentiated cells without clones were used to form the reference basis.
+Then, all of the cloned data was analyzed using scTOP.
+Figure \ref{FIG:4} (c) shows 2D histograms of every single cloned cell regardless of final fate, colored according to the most likely time for each of the bins (see appendix for details \todo{add info in appendix}).
+The histograms with the day 2 undifferentiated scTOP scores on the x-axis show a clear path to differentiation, starting at high scores for the undifferentiated type at early days (which is represented by the lighter color) then decreasing undifferentiated-type score as time passes, with some portion ending up with high scores for each of the possible differentiated fates.
+
+From the histograms displaying two specialized cell fates, one can see that differentiation pathways for some cell types are mutually suppressive while others form a continuous spectrum between identities.
 For example, megakaryocytes and monocytes form a distinct L-shape that indicates the development program for each of these types prevents the program of the other; cells are either megakaryocytes, monocytes, or neither, there is no in-between stage.
 However, the plot for monocytes and neutrophils tells a different story: instead of having a distinct L-shape with narrow distribution along the axes, this plot shows a continuous spectrum of cells that are monocyte-like, neutrophil-like, and every step in-between.
+
 \todo{separate undiff plots}
-
-Day 2 undifferentiated cells without clones and day 6 differentiated cells without clones were used to form the reference basis.
-
-(move to appendix: we excluded erythrocytes and (?) because only one marker gene was used to annotate the original data, which caused a lot of noise since using only one marker gene does not sufficiently isolate the red blood cell population)
 
@@
 	\centering
-		\includegraphics[scale=0.8]{figs/fig4.pdf}
+		\includegraphics[scale=0.5]{figs/fig4.pdf}
 	\caption{Beyond bench-marking: tracing development in cell-type space with scTOP.
@@
 The shade of each point corresponds to the scTOP score for "early epithelium" identity.
-(b), (c) scTOP shows cell fate specification of hematopoietic cells using lineage tracing data from [citation].
-}
+(b), (c) scTOP shows cell fate specification of hematopoietic cells using lineage tracing data from Weinreb et al.
+(2020)}
 	\label{FIG:4}
@@
 
-\subsubsection{Finding genes important for determining cell types (marker genes?)}
-using predictivity to find which cell types are highly correlated with which genes
+\section{Discussion}
 
-\section{Discussion}
+scTOP successfully quantifies cell transitions and enables direct comparison between datasets at the level of individual cells.
+Using a wide range of data from various sources and species, we have shown that scTOP scores identify cell type in agreement with existing annotations without being limited to discrete cluster labels.
+scTOP scores provide clear and meaningful axes to observe differentiation, as shown in the cases of epithelial lung development and hematopoietic specification.
+The input assumptions of our algorithm are explicitly defined by the reference basis, and the results are deterministically determined by linear algebra; scTOP does not require parameter tuning or stochastic machine learning.
+
+By assigning a continuous value to cell identity, scTOP opens the door to numerical analysis of cell fate decisions.
+By tracking cell type, it is possible to obtain detailed information on which transitions are possible in vivo.
+This will provide evidence for mapping out cell lineage trees leading from progenitors to fully-specified fates.
+By plotting the scTOP scores of differentiating cells, it is possible to see which cell types are mutually repressive and which are close relatives in the lineage tree.
+Emerging technologies like Live-seq [citation] promise to allow scRNA-seq of live cells without destroying them, and combining this capability with scTOP would allow scientists to track the process of specialization in individual cells without having to settle for lineage tracing of clones.
+
+In addition to revealing in-vivo developmental pathways, scTOP can measure fidelity of engineered cells, as demonstrated in Herriges et al.
+By locating a cell in cell type space, it is possible to compare the results of directed differentiation protocols and determine which protocol provides the desired result.
+scTOP can guide reprogramming on a granular level in addition to probing which transitions are possible.
+
+By projecting from gene expression space to cell type space, we can directly track cells in the cell decision landscape.
+As discussed in Saez et al.\cite{saez_dynamical_nodate}, such an embedding is a crucial step in verifying theoretical models of cell decisions.
+Many landscape geometries and dynamical systems have been proposed to explain how cell type is decided, but experimental evidence is necessary to test the validity of theoretical models.
+By providing coordinates in the cell decision landscape, scTOP bridges the gap between theory and experiment.
 
@@
 
-\subsection{Details on how the reference bases were constructed}
+\subsection{Analysis details}
+details on how ref. bases were constructed, specific things done for each dataset, etc.
+
+describe accuracy scores + how only cells w. cell types represented in reference basis were analyzed. discuss how "unspecified" rate for unrepresented types (i.e. true negatives) was very high
+
 Describe how MCA, TM, and subsets of data were used. number of cells taken.
+
+hematopoiesis data: we excluded erythrocytes and (?) because only one marker gene was used to annotate the original data, which caused a lot of noise since using only one marker gene does not sufficiently isolate the red blood cell population
 
@@
 
-\subsubsection{Cases where it doesn't work}
+\subsubsection{Cases where it doesn't work}\label{failure_cases}
 All the cases where scTOP fails are expected; they are where the underlying assumption of "reference basis = attractor basins" is false
@@
 
+UMI vs non-UMI
+
 \bibliography{export-data.bib}

### 2022-10-17 main.tex
first edits

@@
 \setlength {\marginparwidth }{2cm}
-\usepackage{todonotes}
 \hypersetup{
@@
 \begin{abstract}
-To elucidate the processes of cell-level phenotypic development and reprogramming, scientists are generating an abundance of single-cell RNA-sequencing (scRNA-seq) tissue atlases, lineage tracing studies, and other measurements.
+To better understand cell-level phenotypic development and reprogramming, scientists are generating an abundance of single-cell RNA-sequencing (scRNA-seq) tissue atlases, lineage tracing studies, and other measurements.
 Using this high-dimensional, noisy data to identify cell type and understand the dynamics of cell fate transitions remains an ongoing problem.
-We introduce single-cell Type Order Parameters (scTOP), an epigenetic-landscape-inspired algorithm to track individual cells in the coordinate space of known cell types.
-Using existing human and mouse datasets that span across the body, we show scTOP reliably identifies cell type and traces the developmental progress of cells.
+We introduce single-cell Type Order Parameters (scTOP), an epigenetic-landscape-inspired algorithm to locate individual cells in the coordinate space of known cell types.
+We use existing human and mouse datasets that span across the body to show scTOP reliably identifies cell type and traces the developmental progress of cells.
 scTOP visualizes the trajectory of lineage tracing experiments by directly comparing each timepoint state to the final fates.
@@
 \section{Introduction}
-Many adult animals exhibit hundreds of specialized phenotypic cell states, all of which arose from a small number of embryonic cells.
-In the early stages of development, embryonic cells possess the transcriptomic flexibility to become virtually any of the eventual cell fates.
+Many adult animals exhibit hundreds of specialized cell states, all of which arose from a small number of embryonic cells.
+In early stages of development, embryonic cells possess the transcriptomic flexibility to become virtually any of the eventual cell fates.
 The questions of how and why cells end up in one fate or another is of interest both in biology and the physics of self-organizing systems.
@@
 Capturing and quantifying the amounts of mRNA in a cell provides a snapshot of its phenotype.
-With the advent of single-cell RNA-sequencing (scRNA-seq), gene expression data of individual cells are being produced at unprecedented rates.
+With the advancement of single-cell RNA-sequencing (scRNA-seq), gene expression data of individual cells are being produced at unprecedented rates.
 Typical scRNA-seq measurements involve thousands of genes and dozens to millions of cells, resulting in high-dimensional data.
-Due to the technical challenges of sequencing RNA, the resulting data is noisy and sparse [Lahnemann].
+Due to the technical challenges of sequencing RNA, the resulting data is noisy and sparse \cite{lahnemann_eleven_2020}.
 Genes with relatively low expression are often recorded as not being expressed at all, resulting in zero-counts called “dropouts”.
-Experimental-to-experiment differences in sequencing method, platform, and other difficult-to-standardize conditions result in technical variations in scRNA-seq measurements called “batch effects.” These challenges merit careful consideration of how to normalize and analyze the data in a way that allows the study of biological variation without obfuscation by technical variation.
+Experimental-to-experiment differences in sequencing method, platform, and other difficult-to-standardize conditions result in technical variations in scRNA-seq measurements called “batch effects.” These challenges merit careful consideration of how to normalize and analyze the data in a way that permits study of biological variation without obfuscation by technical variation.
 
-Three major computational approaches to examining cell fate transitions in scRNA-seq data are clustering, trajectory inference, and automatic cell identification.
-Clustering involves applying dimensional reduction (such as UMAP or SPRING [citations]) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells.
-Groups are then manually annotated according to expression levels of marker genes.
+clustering, trajectory inference, and automatic cell identification are three major computational approaches to examining cell fate using scRNA-seq data.
+Clustering involves applying dimensional reduction (such as UMAP \cite{mcinnes_umap_2018} or SPRING \cite{weinreb_spring_2018}) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells.
+These groups are often manually annotated according to expression levels of marker genes.
 Trajectory inference analysis arranges cells on a manifold and assigns pseudo-times according to transcriptomic similarity.
-There are also methods like partition-based graph abstraction [Wolf] that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions.
-Machine learning classifiers such as support vector machines are able to automatically identify cells with high accuracy [Abdelaal].
+There are also methods like partition-based graph abstraction \cite{wolf_paga_2019} that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions.
+Machine learning classifiers such as support vector machines are able to automatically identify cells with high accuracy \cite{abdelaal_comparison_2019}.
 
-Although there are many useful computational tools, the common methods share certain limitations.
-Dimensional reduction is often one of the first steps to analyze scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships [Chari].
+Although there are many useful computational tools to extract cell identity from scRNA-seq, the common methods share certain limitations.
+Dimensional reduction is often one of the first steps to analyze scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships \cite{chari_specious_2021}.
 Distorted embeddings inconsistently represent distances between individual data points and between cell types, which makes it difficult to interpret cell fate transitions.
-This issue carries over to trajectory inference since nearest neighbors can vary substantially according to the embedding.
-Another limitation is the lack of reproducibility inherent in stochastic machine learning methods [Wattenberg] like t-SNE and UMAP, which produce different results whenever they are executed.
+This issue carries over to trajectory inference, since it uses nearest neighbors which can vary substantially according to the embedding.
+Another limitation is the lack of reproducibility inherent in stochastic machine learning methods \cite{wattenberg_how_2016} like t-SNE and UMAP, which output different results whenever they are executed.
 Depending on exactly which data points are included, the output distribution may change significantly, even of data points that are retained from one iteration to another.
@@
 Inspired by the epigenetic landscape and spin glass physics, scTOP treats cell types as phases and calculates the order parameters for transitions between them.
-It builds on previous work applying attractor-based order parameters to bulk RNA-sequencing [citations].
+It builds on previous work applying attractor-based order parameters to bulk RNA-sequencing\cite{lang_epigenetic_2014} \cite{dame_thyroid_2017}\cite{pusuluri_cellular_2017} \cite{ikonomou_vivo_2020}.
 The inputs into scTOP are the scRNA-seq measurements of query data as well as a reference basis of known cell types.
@@
 
-To calculate cell type similarity scores, scTOP uses a linear algebra projection method inspired by the concept of cell types as attractors in a dynamical system [cite Huang et al.].
-This projection method has been applied previously to bulk RNA-seq samples [citations: Lang et al., 2014; Dame et al., 2017; Pusuluri et al., 2018; Ikonomou et al., 2020a], and we will summarize the theoretical background here.
+To calculate cell type similarity scores, scTOP uses a linear algebra projection method inspired by the concept of cell types as attractors in a dynamical system \cite{huang_cell_2005}.
+This projection method has been applied previously to bulk RNA-seq samples, and we will summarize the theoretical background here \cite{lang_epigenetic_2014}.
 A cell type is a state of the gene regulatory network, so a natural way to represent attractors in this system is with an attractor neural network.
@@
 Cell types are often highly similar in their patterns of gene regulation.
-Kanter and Sompolinsky [citation] defined a pattern storage method for spin-glass-like neural networks where even correlated patterns robustly act as attractors.
+Kanter and Sompolinsky \cite{kanter_associative_1987} defined a pattern storage method for spin-glass-like neural networks where even correlated patterns robustly act as attractors.
 For the same system, they also defined order parameters $a^{\mu}$ which are de-correlated versions of the conventional spin glass order parameter, magnetization.
 $a^{\mu}$ measures the similarity between a given network state and the network state $\mu$, which is a stored pattern (in our case, a gene regulatory pattern that corresponds to a cell type).
-As [Lang et al.] have shown, the order parameters $a^{\mu}$ provided an excellent similarity score for analyzing bulk RNA-seq data.
+As Lang et al.
+have shown, the order parameters $a^{\mu}$ provided an excellent similarity score for analyzing bulk RNA-seq data.
 However, the cells digested by bulk RNA-seq may be in different stages of differentiation, and bulk processing renders it impossible to measure cell fate transitions beyond an average tissue state.
@@
 In pre-processing, the raw RNA-seq count data are first normalized, then z-scored.
-An explanation of the process and how it affects the data distributions is provided in Appendix section \ref{preprocessing}.
+The details of the process and its effects on data distributions are described in Appendix section \ref{preprocessing}.
 
-The way the data is pre-processed before applying the projection method is a vital component of the algorithm, as shown in figure \ref{FIG:1} (a).
+The structure of pre-processing before applying the projection method is a vital component of the algorithm, as shown in figure \ref{FIG:1} (a).
 Each sample is pre-processed independently, unlike other algorithms which process data across samples to find, for example, the most variable genes.
 Depending on the particular application of scTOP, "sample" may refer to either an individual cell or the average expression levels of a subpopulation of cells.
-By processing each sample independently, the algorithm output for one sample is independent of which samples are analyzed alongside that one sample.
-scTOP involves one-shot learning, and there is no stochasticity involved; the algorithm will produce the same output regardless of the particular instance.
+By processing each sample independently, the output for one sample is independent of which samples are analyzed alongside that one sample.
+scTOP involves one-shot learning, and there is no stochasticity involved; the algorithm will produce the same output regardless of the particular instance of execution.
 There are no hyper-parameters involved in the training, which bypasses another source of bias.
-The only assumptions involved in the algorithm are explicit assumptions: the assumptions that the gene expression profiles included in the reference basis are truly representative of the relevant cell types, and that the cell types included are indeed distinct cell phenotypes.
+The only assumptions involved in the algorithm are explicit: that the gene expression profiles included in the reference basis are truly representative of the relevant cell types, and that the cell types included are indeed distinct cell phenotypes.
 
@@
 First, relevant single-cell RNA-seq atlases are gathered to be processed.
-For example, in analyzing mouse lung cells, we used the Mouse Cell Atlas \todo[inline]{Add reference to MCA}, which contains single-cell samples of mouse tissues across \todo[inline]{add \# of organs} organs.
+For example, in analyzing mouse lung cells, we used the Mouse Cell Atlas \cite{noauthor_mapping_nodate}, which contains single-cell samples of mouse tissues across all major organs.
 Once the relevant atlases are identified, we take an average across each cell population that corresponds to a distinct cell type.
@@
 Noise affects the algorithm most strongly when it appears in the basis, since the de-correlation operation involved in the projection separates cells according to the canonical expression levels included in the reference basis.
-\todo[inline]{include hill function discussion in appendix to explain how reference sample size affects scTOP score?} After each cell population is averaged to create the gene expression profiles for the desired cell types, each cell type profile is pre-processed separately.
+After each cell population is averaged to create the gene expression profiles for the desired cell types, each cell type profile is pre-processed separately.
 This creates a modular reference basis where cell types can easily be replaced, removed, or added at will.
@@
 Many atlases across organs and stages of development have been created for both species.
-We used some of these atlases, such as the Mouse Cell Atlas [citation] and the Atlas of Lung Development [citation], as reference bases to analyze datasets from various sources.
+We used some of these atlases, such as the Mouse Cell Atlas \cite{noauthor_mapping_nodate} and the Atlas of Lung Development \cite{negretti_single-cell_2021}, as reference bases to analyze datasets from various sources.
 
@@
 To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from Herriges et al.
-[citation] onto the mouse cell atlas.
+\cite{herriges_durable_2022} onto the mouse cell atlas.
 The Herriges et al.
@@
 Then, all of the cloned data was analyzed using scTOP.
-Figure \ref{FIG:4} (c) shows 2D histograms of every single cloned cell regardless of final fate, colored according to the most likely time for each of the bins (see appendix for details \todo{add info in appendix}).
+Figure \ref{FIG:4} (c) shows 2D histograms of every single cloned cell regardless of final fate, colored according to the most likely time for each of the bins (see appendix\ref{analysis} for details).
 The histograms with the day 2 undifferentiated scTOP scores on the x-axis show a clear path to differentiation, starting at high scores for the undifferentiated type at early days (which is represented by the lighter color) then decreasing undifferentiated-type score as time passes, with some portion ending up with high scores for each of the possible differentiated fates.
@@
 
-\todo{separate undiff plots}
-
 \begin{figure}
@@
 (a) scTOP illustrates specification of alveolar type I and type II cells in murine embryonic development.
-Alveolar cells from the Atlas of Lung Development are compared to a reference basis constructed from adult lung cells from [Kotton lab citation] and early epithelial cells from the Atlas of Lung Development.
+Alveolar cells from the Atlas of Lung Development are compared to a reference basis constructed from adult lung cells from Herriges et al.
+and early epithelial cells from the Atlas of Lung Development.
 The labels next to each data point indicate the age of the mouse embryo from which lungs were extracted, from post-conception day 12 (E12) to day of birth (P0).
@@
 The shade of each point corresponds to the scTOP score for "early epithelium" identity.
-(b), (c) scTOP shows cell fate specification of hematopoietic cells using lineage tracing data from Weinreb et al.
-(2020)}
+(b), (c) scTOP shows cell fate specification of hematopoietic cells using lineage tracing data from Weinreb et al.}
 	\label{FIG:4}
@@
 
-\subsection{Analysis details}
+\subsection{Analysis details \label{analysis}}
 details on how ref. bases were constructed, specific things done for each dataset, etc.
@@
 
-Describe how MCA, TM, and subsets of data were used. number of cells taken.
+Describe how MCA, TM, and subsets of data were used. number of cells taken. hill function discussion to explain how reference sample size affects scTOP score
 

### 2022-11-13 scTOPv1.tex
updated to version 1

	\documentclass[aps,superscriptaddress, notitlepage,longbibliography]{revtex4-1}

\usepackage{amsmath}
\usepackage{physics}
\usepackage[pdftex]{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\setlength {\marginparwidth }{2cm}
\hypersetup{
    colorlinks,
    citecolor=blue,
    filecolor=black,
    linkcolor=red,
    urlcolor=blue
}

\newcommand{\comment}[1]{{\color{blue}#1}}
\newcommand{\edit}[1]{{\color{purple}#1}}
\newcommand{\del}[1]{{\color{red}\st{#1}}}
\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }
% Standardize command font styles and environments
\newcommand{\doccmd}[1]{\texttt{\textbackslash#1}}% command name -- adds backslash automatically
\newcommand{\docopt}[1]{\ensuremath{\langle}\textrm{\textit{#1}}\ensuremath{\rangle}}% optional command argument
\newcommand{\docarg}[1]{\textrm{\textit{#1}}}% (required) command argument
\newcommand{\docenv}[1]{\textsf{#1}}% environment name
\newcommand{\docpkg}[1]{\texttt{#1}}% package name
\newcommand{\doccls}[1]{\texttt{#1}}% document class name
\newcommand{\docclsopt}[1]{\texttt{#1}}% document class option name
\newenvironment{docspec}{\begin{quote}\noindent}{\end{quote}}% command specification environment

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

\definecolor{pastel_green}{rgb}{0.18,0.65,0.34}
\newcommand{\CHW}[1]{\textcolor{pastel_green}{#1}}
%% OPTIONAL MACRO DEFINITIONS
\def\s{\sigma}

\def \M{\mathcal{M}}  
\def \P{\mathrm{P}}
\def \E{\mathbb{E}}

\def \mcl{\mathcal}
\def \mbb{\mathbb}
\def \mbf{\mathbf}

\def\bd{\boldsymbol}
\def\sm{\setminus}
\def\tb{\textbf}

\def\CH{\cal{H}}

\def\<{\langle}
\def\>{\rangle}

\def\thet{{\theta}^{(t)}}
\def\thetp{{\theta}^{(t+1)}}
\def\thets{{\theta}^{(t)\star}}

\def\be{\begin{equation}} 
\def\ee{\end{equation}}
\newcommand \bea {\begin{eqnarray}} 
\newcommand \eea {\end{eqnarray}} 
\newcommand \sign {\hbox{sign}} 
\newcommand{\nn} {\nonumber}
\newcommand{\mbbE} {\mathbb{E}}
\begin{document}

\title{ scTOP: physics-inspired order parameters for cellular identification and visualization of developmental trajectories}

\author{Maria Yampolskaya}
\email{mariay@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\author{Michael Herriges}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\author{Laertis Ikonomou}
\affiliation{Department of Oral Biology. University at Buffalo, The State University of New York, Buffalo, NY, USA}
\affiliation{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
\author{Darell Kotton}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}

\author{Pankaj Mehta}
\email{pankajm@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{Faculty of Computing and Data Science, Boston University, Boston, MA 02215, USA}
\affiliation{Biological Design Center, Boston University, Boston, MA 02215, USA}

\begin{abstract}
Recent advances in single-cell RNA-sequencing (scRNA-seq) and lineage tracing techniques provide an unprecedented window into the biology of cellular identity. The wealth of data calls for new theoretical and computational frameworks for understanding cell fate specification, accurately classifying cell fates from expression data, and integrating knowledge from cell atlases. Here, we present single-cell Type Order Parameters (scTOP): a statistical-physics-inspired approach for constructing “order parameters” for cell fate given a reference basis of cell types.  scTOP achieves near state-of-the-art performance for cell identification at the resolution of single cells and yields interpretable visualizations of developmental trajectories, such as bifurcations between closely related cell fates. Importantly, scTOP does this without using feature selection, statistical fitting, or dimensional reduction (e.g. UMAP, t-SNE, PCA, SPRING). We illustrate the power of scTOP on a wide variety of human and mouse datasets (both in vivo and in vitro), including existing tissue atlases and lineage tracing data. We also provide an easy-to-use Python package implementation of scTOP. Our results suggest that physics-inspired order parameters can serve as an important tool for understanding and analyzing developmental landscapes and cellular identity across biological contexts and organisms.
\end{abstract}

\maketitle

\section{Introduction}
Humans and other mammals have hundreds of specialized cells that arise through a complex developmental process  [CITE]. At early embryonic stages, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers[CITE]. However, as embryos mature, cellular identity becomes less plastic and more specified through a complex spatiotemporal process involving signaling and environmental cues. The questions of how and why cells end up in one fate or another is a fundamental question in developmental biology [CITE]. More practically, understanding the process of cell fate specification is crucial for developing directed differentiation protocols for engineer cells \emph{in vitro}, with important clinical application for human health and disease [CITE].

One powerful set of techniques for characterizing cellular phenotype are transcriptome-wide measurements using single-cell RNA sequencing (scRNA-seq). Typical scRNA-seq experiments involve thousands of genes and dozens to millions of cells. For this reason, this data is intrinsically high-dimensional and requires specialized computational tools for analysis and interpretation [CITATION DUMP]. This task is especially difficult because the technical challenges of sequencing RNA from individual cells results in data that is noisy and sparse \cite{lahnemann_eleven_2020}. Genes with relatively low expression are often recorded as not being expressed at all, resulting in zero-counts called ``dropouts'' [CITE]. Experimental-to-experiment differences in sequencing method, platform, and other difficult-to-standardize conditions result in technical variations in scRNA-seq measurements called ``batch effects'' [CITE].These challenges merit careful consideration of how to normalize and analyze the data in a way that permits study of biological variation without conflation by technical variation.

scRNA-seq is often used for cellular identification. The most frequently used workflow involves applying dimensional reduction (such as UMAP \cite{mcinnes_umap_2018} or SPRING \cite{weinreb_spring_2018}) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells. These groups are then often manually annotated according to expression levels of known marker genes [CITE]. Machine learning classifiers such as support vector machines are also commonly used identify cells after training with labeled data \cite{abdelaal_comparison_2019}. Another common analysis task is trajectory inference, where the goal is to infer developmental trajectories from static time shots [CITE]. Trajectory inference arranges cells on a low-dimensional manifold and assigns pseudo-times according to transcriptomic similarity. There are also methods like partition-based graph abstraction \cite{wolf_paga_2019} that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions.

Despite the power and utility of these computational tools, most commonly employed methods for scRNA-seq data share certain limitations. Dimensional reduction is often one of the first steps to analyze scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships \cite{chari_specious_2021}. Distorted embeddings inconsistently represent distances between individual data points and between cell types, making it difficult to interpret cell fate transitions. Similar issues arise in trajectory inference because  local neighborhoods can also vary substantially with the choice of embedding. Another limitation of these methods is the lack of reproducibility inherent to stochastic machine learning methods like t-SNE and UMAP, which output different results whenever they are executed  \cite{wattenberg_how_2016}. Many algorithms also perform feature selection (e.g., limiting there analysis to the most variable genes or highly-expressed genes) or require choosing parameters such as the perplexity, making results even more sensitive to the inclusion or exclusion of different datasets in the analysis [CITE].

An ideal analysis algorithm for scRNA-seq data should possess certain desirable properties. 
First, the algorithm should incorporate explicit assumptions that can be kept consistent between analyses. Second, the algorithm should be deterministic, further enhancing reproducibility. Third, the axes in visualization should be interpretable, intuitive, and remain fixed across analysis and datasets. Fourth, the algorithm should measure cell identity in a continuous way instead of assigning discrete identity labels, thus enabling a more subtle and fine-grained analysis of cellular identity and developmental trajectories. Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single cell atlases now being generated [].

Here, we present a new physics inspired algorithm, single-cell Type Order Parameters (scTOP), that has many of these properties. The algorithm is based on the statistical physics concept of an order parameter -- a coarse grained quantity that can distinguish between phases of matter. Arguably the most famous example of an order parameter is  magnetization in a ferromagnetic system (i.e., a system that can spontaneously magnetize). In a ferromagnet, the magnetization measures the alignment of spins and can be used to distinguish magnetic and paramagnetic phases [CITE STAT MECH]. It is straight forward to generalize the idea of magnetization to more complex settings such as spin glasses and attractor neural networks [CITE-AMIT]. In the context of epigenetic landscape models of cell identity, this takes the form of generalized magnetizations that measure how aligned a gene expression state is with each of the attractor basins for different cell fates \cite{lang_epigenetic_2014,pusuluri_cellular_2017}. These generalized magnetizations serve as natural order parameters for cellular identity and can be calculated directly from data \cite{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}. The scTOP algorithm builds on previous work by developing methods for integrating scRNA-seq data, greatly expanding the power and utility of this approach.

The inputs to scTOP are the scRNA-seq measurements of query data as well as a reference basis containing the archetypal gene expression profiles for cell types of interest. The output of scTOP are the coordinates  of the query data  in cell type order parameter space (i.e., generalized magnetizations), with each output dimension serving as a measure of similarity between the query data and a distinct cell type in the reference basis. In this way, scTOP naturally projects gene expression deterministically onto a space with dimension equal to the number of cell types in the reference basis. These generalized coordinates can be used for both cell type identification and visualization. One appealing feature of scTOP is that the assumptions of the algorithm are explicit and are contained entirely in the choice of reference basis. This allows direct comparison across datasets and settings since  the coordinates of, and distances between, data points do not change as long as one uses the same reference basis. One important limitation of scTOP worth noting  is that it cannot be easily used to identify novel cell fates. The underlying reason for this is that scTOP assumes one already know what cell types are of interest in the form of a reference basis and for this reason is poorly suited for exploratory work where the goal is to identify unknown cell types.

In this paper, we show that scTOP can be used to reliably assess cell identity and accurately classify cell type with high accuracy at the resolution of individual cells.  We demonstrate this using mouse and human samples from lungs, brains, and other organs. By analyzing lineage tracing data and embryonic development data, we also establish scTOP's capacity to quantify transitions between cell types, including bifurcations between closely related cell fates.

We have implemented scTOP in an easy-to-use Python package, available on the Python Package Index [link] and Github [link], and an R package, available on Bioconductor [link] and Github [link]. All code, tutorials, and examples are available on the corresponding Github site [].

\section{Methods}
\begin{figure}
	\centering
		\includegraphics[scale=0.8]{figs/scTOP manuscript fig1 - no aggregate.pdf}
	\caption{Steps involved in the scTOP algorithm. (a) Each cell in the query data is pre-processed independently of the other cells in the dataset. (b) Single-cell RNA-seq atlases, such as the Mouse Cell Atlas, are used to define the reference basis in the algorithm. (c) scTOP scores are the projections of query data onto the non-orthogonal subspace of cell types. Since there is no statistical fitting, no tuning parameters are involved. (d) The process of finding scTOP scores is shown in the grey section, and the blue section illustrates cell-type space. This space has as many dimensions as there are cell types in the reference basis. We can visualize this concept in 1, 2, or 3 dimensions, depending on which cell types are most relevant. In the 3-dimensional case, the shadows on the bounds of the scatter plot are included to better illustrate the position of points in 3-dimensional space.}
	\label{FIG:1}
\end{figure}

To calculate cell type similarity scores, scTOP uses a linear algebra projection method inspired by the concept of cell types as attractors in a dynamical system \cite{huang_cell_2005}. This projection method has been applied previously to bulk RNA-seq samples, and we will summarize the theoretical background here \cite{lang_epigenetic_2014}. A cell type is a state of the gene regulatory network. A natural way to represent attractors in this system is with an attractor neural network, where attractors correspond to different cell types. If we represent each gene as a node in a network with an associated value measuring gene expression (positive value representing high gene expression, low value indicating low gene expression), then a cell type may be denoted by a vector of gene expression values $S_i$, where $i=1,\ldots, G$ spans the $G$ genes being measured.  The gene expression profiles
corresponding to the $C$ cell type attractors are also $G$ dimensional vectors $\xi_i^\mu$, where $\mu=1,\ldots, C$ spans the $C$ cell types being "stored" in the network. In what follows, we assume that $G > C$, namely the number of genes being measured is greater than the number of cell types.

Cell types are often highly similar in their patterns of gene regulation. Kanter and Sompolinsky \cite{kanter_associative_1987} defined a pattern storage method for spin-glass-like neural networks where even correlated patterns robustly act as attractors. For the same system, they also defined order parameters: generalized magnetizations $a^{\mu}$, where $\mu$ iterates through each of the stored cell type attractors. The $a^\mu$ can be understood as "decorrelated" versions of the conventional spin glass magnetization order parameter that measures the similarity between a given network state and the network state $\xi^\mu$. 
Explicitly, the $C$ generalized magnetizations $a^\mu$ can be calculated for a gene expression state $S_i$ via the expression
\be \label{op_eqn}
a^\mu= \sum_{\nu=1}^C\sum_{j=1}^G [A_{\mu \nu}]^{-1}\xi_j^\nu S_j.
\ee
where
\be
A_{\mu \nu}= \sum_{j}^G  \xi_j^\mu \xi_j^\nu,
\ee
is the overlap matrix of gene expression profiles of different cell types. As Lang et al. have shown, the order parameters $a^{\mu}$ provided an excellent similarity score for analyzing bulk RNA-seq data \cite{lang_epigenetic_2014}. However, bulk RNA-seq measurements average gene expression over many cells, rendering it impossible to measure cell fate transitions beyond an average tissue state.

scTOP uses the same order parameters, $a^{\mu}$, to track cell type at the resolution of individual cells, making it possible to directly observe cells in different stages of differentiation. In both attractor neural networks and previous applications to bulk RNA-seq data, the expression vectors $S_i$ were binary variables. However, we have found that for working with scRNA-seq data it is helpful to treat $S_i$ as continuous.

The first stage of the algorithm is preprocessing and normalization of the the input data. Importantly, in our algorithm measurements from each cell are normalized independently and do not depend on the expression profiles of any other cells in the dataset. In the raw scRNA-seq data for each cell, each gene has an integer value which counts the number of corresponding RNA reads. Genes are then rank-ordered. The rank order is then converted to a z-score assuming a normal distribution, representing the percentile rank of the expression of a gene within the cell relative to all other genes being measured. For example, a gene that has higher expression than $97.8\%$ of genes being measured is assigned a score of $z=2$, while a gene that has higher expression than half of genes is assigned a score of $z=0$. The details of the process and its effects on data distributions are described in Appendix section \ref{preprocessing}.

The structure of pre-processing before applying the projection method is a vital component of the algorithm, as shown in figure \ref{FIG:1} (a). Each cell is pre-processed independently, unlike other algorithms which process data across samples to find, for example, the most variable genes. By processing each cell independently, the output for one cell is independent of which other cells are being analyzed. Although the strength of scTOP is in its ability to analyze single cells, it's also possible to analyze gene expression levels averaged across cell populations. This can be useful in cases where one wants to probe selected subsets of cells at low resolution; for example, if one wants to identify pre-defined clusters of cells.

scTOP involves one-shot learning, and there is no stochasticity involved; the algorithm will produce the same output regardless of the particular instance of execution. There are no hyper-parameters involved in the training, which bypasses another source of bias. The only assumptions involved in the algorithm are explicit: that the gene expression profiles included in the reference basis are truly representative of the relevant cell types, and that the cell types included are indeed distinct cell phenotypes.

Constructing a representative, accurate reference basis that is relevant to the data being analyzed is vital to the scTOP algorithm. This process is shown in figure \ref{FIG:1} (b). First, relevant single-cell RNA-seq atlases are gathered to be processed. For example, in analyzing mouse lung cells, we used the Mouse Cell Atlas \cite{noauthor_mapping_nodate}, which contains single-cell samples of mouse tissues across all major organs. Once the relevant atlases are identified, we take an average across each cell population that corresponds to a distinct cell type. It is important that enough cells are included in each average, in order to mitigate the potent effects of noise. The minimum sample size to achieve reasonable results is 200 cells. Noise affects the algorithm most strongly when it appears in the basis, since the de-correlation operation involved in the projection separates cells according to the canonical expression levels included in the reference basis. After each cell population is averaged to create the gene expression profiles for the desired cell types, each cell type profile is pre-processed separately using the same procedure as that used for processing single cells: normalization followed by z-scoring . This creates a modular reference basis where cell types can easily be replaced, removed, or added at will.

Although they were originally inspired by attractor neural networks and spin glass physics, the order parameters $a^{\mu}$ can also be understood as non-orthogonal projections onto the subspace of cell type space. This idea is shown in figure \ref{FIG:1} (c). A single-cell RNA-seq measurement of a sample produces a list of values corresponding to the expression levels of different genes, and thus can be conceptualized as a vector in $G$-dimensional gene expression space, where $G$ is the number of genes. Single-cell RNA-seq atlases provide reference gene expression profiles for hundreds of known cell types; these reference profiles also exist as vectors in gene expression space. If we compile a reference basis consisting of reference profiles corresponding to cell types, these vectors span a $C$-dimensional subspace of gene expression space, where $C$ is the number of cell types in the basis. There are far more genes than cell types: $C < G$. We can project the measured sample vector onto the cell type subspace in order to find the coordinates in cell type space, which enables identification of individual cells and tissues. Since the cell type profiles are largely correlated, they form a non-orthogonal subspace, and the projection onto this space is a non-orthogonal projection.

We can write the sample vector as a sum of the projected components and the component perpendicular to the cell type subspace: 
\be
S_i = \sum_\mu a^{\mu} \xi^{\mu}_i + S^{\perp}_i,
\ee
where $S^{\perp}_i$ is the expression level of the $i$th gene of the component of the sample vector that is perpendicular to the cell type subspace. $S^{\perp}$ is the part of the sample vector that is not explained by the known cell types. It results from biological processes such as cell cycle variation and non-biological sources such as scRNA-seq noise.

There are two types of scTOP scores: aggregate and individual. The aggregate scoring method is useful for analyzing identified sub-populations of data at once. The scores tend to be higher, since aggregating gene expression data decreases the effects of noise. In aggregate scoring, the raw RNA count data is averaged across each sub-population, and then these average gene expression profiles are pre-processed and projected onto the reference basis. See figure \ref{FIG:1} (d). The resulting scTOP score provides the location of the sub-population in the cell type dimensions specified by the reference basis, which can be visualized in one, two, or three dimensions by selecting one, two, or three relevant cell types. In individual scoring, as shown in figure \ref{FIG:1} (e), each individual cell is pre-processed and projected onto the reference basis separately. Due to the well-documented effects of scRNA-seq noise [CITE], the resulting scores tend to be lower. Nonetheless, the resulting scores are able to accurately and reproducibly identify and track individual cells through cell type space, in whatever number of dimensions (i.e. cell types) one chooses.

The crux of scTOP is choosing a good reference basis. scTOP scores are reliable when the input reference basis incorporates many cells and represents a wide, relevant set of specialized cell types. Since scRNA-seq data is noisy and dropout is prevalent, many cells of each reference cell type must be included in order to create representative gene expression profiles. The choice of cell types in the basis is vital, too. The main underlying assumption of scTOP is that the reference cell types accurately represent attractors of the gene regulatory network. If the cell types aren't sufficiently different from one another, as in early embryonic development or certain stromal cells where the same cell type is compared between organs, the resulting scTOP scores may be unreliable; exhibiting extreme sensitivity to small changes in the basis or becoming incoherent. More information on the limitations of scTOP may be found in appendix [insert appendix num.]. Generally, outside these very limited edge cases, we have found that scTOP can reliably distinguish even extremely closely related cells such (see example of pulmonary alveolar type 1 and type 2 cell lineages below).

The accuracy of cell type information that scTOP can extract from scRNA-seq is limited by the low signal-to-noise ratio. As shown in appendix [insert appendix num.], the scTOP scores for false identities form a Gaussian distribution around zero. These distributions give an estimate of error. In aggregate scTOP scoring, RNA counts are averaged across a population and the resulting score is generally higher for relevant cell types than the average individual scTOP score for the same population. This is because the averaging step in aggregate scoring mitigates the effects of sparsity and noise in the RNA count matrices due to dropout.

Many other methods for classifying and tracking cell type can be computationally intensive. Dimensional reduction methods like t-SNE involve finding the distance between every pair of points in the data set. For scRNA-seq data with $G$ genes and $N$ cells, this involves computing the $G$-dimensional distance between every pair of $N$ cells. While there are optimized ways to calculate this, it can be time-consuming when you are working with thousands to millions of cells. The core of the scTOP is equation \ref{op_eqn}, which only involves matrix inversion and multiplication. The matrix inversion only involves the reference basis (which contains $C$ cell types), so it only needs to be computed once. When applying scTOP to $N$ cells, the computation that scales with the number of cells is the multiplication of a $C \cross G$ matrix with a $G \cross N$ matrix. Modern programming libraries like Numpy are highly optimized for simple linear algebra functions; the result is that scTOP takes milliseconds to run, even for very large datasets. While other methods involve time- and memory-intensive statistical fitting, scTOP is fast b.

\section{Results}

\subsection{Benchmarking and validation: Classification of cell fate}
{\color{red}: This section is already pretty good -- I think just a little terse -- we have to talk up all the results a little bit by comparing to existing methods putting things in context -- 2-3 sentences at end of each example -- remember people think exactly what you tell them -- so tell them/show them why each example is impressive}

scTOP scores reliably and accurately identify the cell types of individual cells as well as of clusters of cells. To show this, we applied the algorithm to several scRNA-seq datasets across species and across laboratories. scTOP is organism-agnostic as long as an appropriate reference basis is used. Mice and humans are among the most common subjects of scRNA-seq measurements. Many atlases across organs and stages of development have been created for both species. We used some of these atlases, such as the Mouse Cell Atlas \cite{noauthor_mapping_nodate} and the Atlas of Lung Development \cite{negretti_single-cell_2021}, as reference bases to analyze datasets from various sources.

Table \ref{table:1} lists each of the datasets examined in this paper with their respective scTOP score accuracies. ``TopN" refers to the percent of cells whose true cell types were in the top ``N" scTOP scores, with a score greater than 0.1. "Unspecified" refers to the percent of cells for which the highest scTOP score was less than 0.1; this represents a failure of the algorithm to confidently identify the cell. The accuracies across species and tissues are high and the unspecified percentages are very low, even in dataset exceeding a million cells.

\begin{table}[h!]
\centering
\begin{tabular}{| c | c | c | c | c | c | c |}
\hline
Query Data & Species & Reference Basis & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Total cells \\ 
\hline
Kotton lab & Mouse & Mouse Cell Atlas & 97.19 & 99.77 & 2.69 & 4,805 \\ 
\hline
Brain atlas & Mouse & Self & 86.28 & 96.89 & 1.78 & 1,161,041 \\
\hline
CellBench & Human & Self & 100 & 100 & 0 & 2,822 \\
\hline
Lung atlas & Human & Self & 79.78 & 85.16 & 3.32 & 2,952 \\  
\hline
\hline
Data set & Species & Reference Basis & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Number of cell types \\
\hline
Mouse Cell Atlas & Mouse & Tabula Muris & 79.17 & 87.5 & 0 & 48 \\
\hline

\end{tabular}
\caption{Accuracy scores of included sample types for each the described data sets.}
\label{table:1}
\end{table}
{\color{red}: PM: We should say how well we do compare to other methods here-- does not have to be comprehensive but should show we perform about as well as SOTA --few sentences and revisit below each example -- beat this horse until it's dead.}

\subsubsection{Example 1: lung lineages}
To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from Herriges et al. \cite{herriges_durable_2022} onto the mouse cell atlas. The Herriges et al. dataset  measures gene expression levels of six different specialized lung types from mice used as control samples. See figure \ref{FIG:2} (a). The reference basis used to find the individual scTOP scores for this data set was derived from the Mouse Cell Atlas. Despite the fact that the Kotton lab dataset and the Mouse Cell Atlas were created in different laboratories and different sequencing methods, the resulting scTOP scores distinctly separate the cells from the Kotton laboratory. The resulting clusters agree with the annotations of the true cell type provided by the Kotton laboratory.

\subsubsection{Example 2: Mouse Brain Atlas}

The mouse brain atlas\cite{yao_taxonomy_2021} sequenced over one million cells and clustered them into 42 subclasses and 101 supertypes. For a reference basis, we reserved a subset of 200 cells from each of the identified subclasses to use as training data. The test data, from which the accuracy score was calculated, was the remaining cells of each type not included in the training data. Ten different cell types are visualized in a grid in figure \ref{FIG:2} (b). The color of each grid square corresponds to the average scTOP score of cells annotated as the type on the y-axis, with the cell type corresponding to the scTOP dimension on the x-axis. As shown by the distinct blue diagonal, the average scTOP score correctly matches the corresponding annotated cell type.

\subsubsection{Example 3: Matching cell types to tissues}

scTOP correctly identifies individual cells, and it can also identify the tissue to which a population of cells belongs. As shown in \ref{FIG:2} (c), we found the aggregate scTOP scores for cells from the Mouse Cell Atlas projected onto the reference basis of Tabula Muris organs \cite{schaum_single-cell_2018}. Most cell types were correctly and strongly identified with the annotated tissue of origin. Parenchymal cell types were the most associated with the correct organ, while stromal cells were sometimes misidentified -- for example, the mammary gland endothelial cells. This is consistent with the fact that stromal cells are less specified to each organ. A discussion of the accuracy of scTOP when applied to stromal types versus parenchymal types may be found in appendix \ref{failure_cases}.

\begin{figure}
	\centering
		\includegraphics[scale=0.5]{figs/scTOP manuscript fig 2.pdf}
	\caption{scTOP identifies mouse cell types. (a) Using the Mouse Cell Atlas as the reference basis, we show that mouse lung types separate clearly on scTOP score axes. The data points correspond to individual cells, and in the first row of plots, the marker shape and color indicate the "true type" as determined by annotations from Herriges et al. The axes used are scTOP scores for similarity with lung ciliated cells and alveolar types I and II. The second, third, and fourth rows of plots have individual cells colored by the marker genes for the type indicated by the color bars at the end of each row. (b) A subset of cells from the Mouse Brain Atlas are used as the reference basis to query other cells from the same data set. The shade of each bin corresponds to the average scTOP score for individual cells of the type indicated on the y-axis, compared to the reference types on the x-axis. The diagonal indicates that scTOP accurately matches query cells to the true reference type. (c) Aggregate samples from the Mouse Cell Atlas are compared to a reference basis consisting of organs from Tabula Muris. The scTOP scores are significantly higher when comparing a cell type with the organ of origin. This coarse-grain identification shows scTOP is able to classify tissues at different resolutions.}
	\label{FIG:2}
\end{figure}

\subsubsection{Example 4:  Human data}

The CellBench dataset \cite{tian_benchmarking_2019} was designed specifically to benchmark scRNA-seq analysis methods. We used the subset of data pertaining to five cancer lines. 200 cells from each line were used as training data to create the reference basis, then the rest of the cells were input as query data. As shown in figure \ref{FIG:3} (a) and table \ref{table:1}, scTOP classifies cell line identity of the test data with 100 percent accuracy. {\color{red} PM: How did other methods do?}

To further prove the power of scTOP in classifying human samples, we analyzed data from the human lung atlas dataset \cite{travaglini_molecular_2020}. This dataset sequenced thousands of human lung cells and identified 58 phenotypic popoulations. For our analysis, we restricted the training and test data to epithelial and stromal types that had at least 200 cells in each type population. The reference basis was created using 80\% of the cells from each population, and the query data consisted of the remaining 20\% of the data. Figure \ref{FIG:3} (b) shows that the scTOP scores were consistently high in cases where the score type matched the true type of the query.

\begin{figure}
	\centering
		\includegraphics[scale=0.8]{figs/fig3.pdf}
	\caption{scTOP correctly identifies human tissues. (a) The CellBench data set contains human cell cancer lines specifically processed to serve as benchmarking for scRNA-seq algorithms. We take a subset of cells to use as the reference basis in analyzing the rest of the CellBench data. The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis. The y-axis lists the true types of the queried cells. The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types. (b) Similar to (a), cells from the Human Lung Atlas are compared to a reference basis constructed from cells from the Human Lung Atlas. The colors of the bins correspond to average scTOP scores of individual cells. Again, the diagonal is the most prominent section, showing that cells of the y-axis types are correctly matched to the reference x-axis identities.}
	\label{FIG:3}
\end{figure}

As demonstrated by the diverse data sources examined here, as long as a well-defined reference basis is available, scTOP works extremely well  for cell type identifciation across different measurement conditions, species, organs, and resolutions. This performance is especially impressive since as discussed earlier, the algorithm does not use dimensional reduction methods such as PCA, tSNE, UMAP, or SPRING, or make use of any statistical fitting procedures.

\subsection{Tracking cell type transitions}

{\color{red}: PM: This section needs to be revisited once figures are finalized}

We now illustrate how scTOP can be used to visualize developmental dynamics directly in cell type space.

\subsubsection{Lung development}
{\color{red} PM: We need to add paragraph with some biological conclusions here -- even if already known sighting relevant literature.. when we see  divergences in lung development ... etc. --- will need final analysis}

The healthy formation of lungs is vital to mammalian life. For a mammal to breathe and live independently after birth, the final stages of embryonic development must equip it with the ability to exchange gases with its environment. Zepp et al. [CITE] sought to examine the formation of pulmonary alveoli, small sacs where oxygen and carbon dioxide are transferred between the lungs and the air, in the developing mouse. They sequenced mouse lungs between post-conception day 12.5 (E12.5) and postnatal day 42 (P42). We found the scTOP scores for cells annotated by Zepp et al. as type I pneumocytes, type II pneumocytes, and epithelial progenitor cells. The reference basis consisted of adult cell types from the Mouse Cell Atlas and an E12 epithelial progenitor cell type from the Atlas of Lung Development \cite{negretti_single-cell_2021}. In \ref{FIG:4}, we show the scTOP scores for the two adult alveolar types using the Mouse Cell Atlas as the reference basis.

{\color{red} MY: Discuss the clear transition to adult cells, then speculate about the co-expressing middle cells (which exist even at P7). Figure 1I from Zepp et al. conjectures that one begins to see separate alveolar type 1 and type 2 populations between E15.5 and E17.5. According to our plot, there are populations of cells clearly headed towards the two types, but there is also a population of cells in between expressing both type 1 and type 2 signatures. Show that it doesn't express Krt8 or the other "transitional cell" genes; it looks more like a spurious attractor/combination of AT1 + AT2}

Because scTOP provides us with reference axes that don't change according to whatever data is included, it is possible to track different data points across time on the same axes. These axes are also easily interpretable: they measure the similarity to a canonical known cell type.

\begin{figure}
	\centering
		\includegraphics[scale=0.75]{figs/scTOP manuscript fig4.pdf}
	\caption{Beyond bench-marking: tracing development in cell-type space with scTOP, using developmental mouse lung data. (a) scTOP illustrates specification of alveolar type I and type II cells in murine embryonic development. Alveolar cells from Zepp et al. are compared to a reference basis constructed from adult lung cells from Herriges et al. and early epithelial cells from the Atlas of Lung Development. The labels next to each data point indicate the age of the mouse embryo from which lungs were extracted, from post-conception day 12 (E12) to day of birth (P0). }
	\label{FIG:4}
\end{figure}

\subsubsection{Visualizing hematopoiesis using lineage tracing data}

The case of a general cell type differentiating into mature types is exemplified by the case of hematopoiesis, the process of blood regeneration that takes place in bone marrow. In this process, the progenitor is the hematopoietic stem cell, which may become one of a number of blood types such as megakaryocyte, erythrocyte, monocyte, etc. It is unknown what causes a hematopoietic stem cell to end up in one fate or another; perhaps it is due to small differences in the epigenetic landscape when the stem cells are dividing, or perhaps it is due to small differences in signals the cells receive over time. Weinreb et al. \cite{weinreb_lineage_2020} sought to answer this question using a combination of scRNA-seq and lineage tracing. Since cells are destroyed when they are sequenced, it isn't possible to repeatedly sequence the same cell to watch it differentiate over time. As a substitute for watching one cell develop, there are various methods of tracing the lineage of a cell by sequencing clones and observing the gene expression levels of clones. Weinreb et al. created a tool for lineage and RNA recovery (LARRY) and applied it to hematopoiesis in mice. By tagging cells with DNA barcodes then letting them multiply, they were able to use scRNA-seq to identify clone sisters across generations of cell division.

To apply LARRY to blood cells, they extracted hematopoietic progenitor cells from mice, barcoded the cells, waited for the progenitors to multiply in primary culture, then performed scRNA-seq at days 2, 4, and 6 post-barcoding. Then, they extracted blood cells at four and six days after transplantation. The resulting dataset contains over 300,000 cells, with 2,632 clones spanning time points. Clones belonging to the same clonal family are referred to as "sisters." They then used pseudotime orderings to identify a transcriptomic map to identify lienages of the cells. This map used transcriptomic similarity to place the individual cells accoridng to "pseudotime". With scTOP, creating a pseudotime ordering is unnecessary. Instead of arranging cells according to transcriptomic similarity to one another using a dimensional reduction method (what method did they use??), we used scTOP to compare clones to day 2 undifferentiated cells and day 6 mature cell types. We used the annotations from Weinreb et al. to identify these populations. The resulting reference basis consisted of day 2 undifferentiated progenitors and day 6 neutrophils, monocytes, megakaryocytes, mast cells, eosiniphils, and basophils.

In figure \ref{FIG:5}a, we found the scTOP scores for one particular clonal family that ended up with some sisters that are neutrophils and some that are monocytes. The colors and shapes of the points and colors of the histograms indicate the day the cells were extracted. The undifferentiated scores are indicated on the y-axis of the bottom row, and by the histograms on the lower right. It's clear to see that the cells at day 2 have high undifferentiated scores, and the undifferentiated score decreases as the cells mature -- the histograms for the undifferentiated score move to the left from day 2 to day 6. As the cells mature, their scores for neutrophil and monocytes increase. The graph with neutrophils on the horizontal axis and monocytes on the vertical axis shows that one subset of the population moved towards high-monocyte, low-neutrophil score and the other subset moved towards low-monocyte, high-neutrophil. This indicates that the same progenitor cell can result in either monocytes or neutrophils.

While figure \ref{FIG:5}a shows scTOP scores for one clonal family, figure \ref{FIG:5}b shows a 2-dimensional histogram of scTOP scores for all of the multi-generational clonal families. The horizontal axes on these plots indicate the "undifferentiated" score while the vertical axes indicate all of the mature cell types included in the reference basis. Each bin of the 2-dimensional histogram is colored according to the average day of all the cells that fall within that bin; although the cells were only sampled at days 2, 4, and 6, the bin time points form a continuous range because they are averages of varying quantities of day 2, 4, and 6 cells. These plots show that, as expected, the day 2 cells generally have high "undifferentiated" scores, which decrease over time. As time passes, the scores corresponding to a given differentiated cell type increase or remain at zero, since some clones end up at that particular type and some end up at another type.

Figure \ref{FIG:5}c shows 2D histograms for differentiated type scores compared to other differentiated type scores. One can see that differentiation pathways for some cell types are mutually suppressive while others form a continuous spectrum between identities. For example, megakaryocytes and monocytes form a distinct L-shape that indicates the development program for each of these types prevents the program of the other; cells are either megakaryocytes, monocytes, or neither, there is no in-between stage. However, the plot for monocytes and neutrophils tells a different story: instead of having a distinct L-shape with narrow distribution along the axes, this plot shows a continuous spectrum of cells that are monocyte-like, neutrophil-like, and every step in-between.

{\color{red} PM: We need 3-4 paragraphs discussing our biological results. For example, what lineages to single clones end up at together, which don't they... Then, we also need to place our conclusions in context of original paper -- and ideally what other biology is known... Like they emphasize this transition cells or whatever See Figure 4 of theirs and Ref 27 they dialogue with . }

\begin{figure}
	\centering
		\includegraphics[scale=0.75]{figs/scTOP manuscript fig5.pdf}
	\caption{Beyond bench-marking: tracing development in cell-type space with scTOP, using lineage tracing hematopoietic mouse data. (a) {\color{red}} (Describe plots) }
	\label{FIG:4}
\end{figure}

\section{Discussion}
{\color{red}: We will revisit after finalizing above... here is just some preliminary comments}

scTOP successfully quantifies cell transitions and enables direct comparison between datasets at the level of individual cells. Using a wide range of data from various sources and species, we have shown that scTOP scores identify cell type in agreement with existing annotations without being limited to discrete cluster labels. scTOP scores provide clear and meaningful axes to visualize differentiation, as shown in the cases of epithelial lung development and hematopoietic specification. The input assumptions of our algorithm are explicitly defined by the reference basis, and the results are deterministically determined by linear algebra; scTOP does not require parameter tuning or stochastic machine learning. {\color{red}: PM should add it is fast}

By assigning a continuous value to cell identity, scTOP opens the door to numerical analysis of cell fate decisions. By tracking cell type, it is possible to obtain detailed information on which transitions are possible in vivo. This will provide evidence for mapping out cell lineage trees leading from progenitors to fully-specified fates. By plotting the scTOP scores of differentiating cells, it is possible to see which cell types are mutually repressive and which are close relatives in the lineage tree. Emerging technologies like Live-seq [citation] promise to allow scRNA-seq of live cells without destroying them, and combining this capability with scTOP would allow scientists to track the process of specialization in individual cells without having to settle for lineage tracing of clones.

In addition to revealing in-vivo developmental pathways, scTOP can measure fidelity of engineered cells, as demonstrated in Herriges et al. By locating a cell in cell type space, it is possible to compare the results of directed differentiation protocols and determine which protocol provides the desired result. scTOP can guide reprogramming on a granular level in addition to probing which transitions are possible.

By projecting from gene expression space to cell type space, we can directly track cells in the cell decision landscape. As discussed in Saez et al.\cite{saez_dynamical_nodate}, such an embedding is a crucial step in verifying theoretical models of cell decisions. Many landscape geometries and dynamical systems have been proposed to explain how cell type is decided, but experimental evidence is necessary to test the validity of theoretical models. By providing coordinates in the cell decision landscape, scTOP bridges the gap between theory and experiment.

\section*{Acknowledgments}
We acknowledge useful discussions with Jason Rocks, Robert Marsland, and Jason Rocks. The work was funded by a grant to the Boston University Kilachand Multicellular Design Program (to DK and PM) and NIH NIGMS 1R35GM119461 to PM. We also acknowledge programming advice from Alena Yampolskaya.

\section*{Appendix}
\beginsupplement
\subsection{Pre-processing} \label{preprocessing}
The raw scRNA-seq count data for each dataset was downloaded as a matrix of genes and cells. An example of such a matrix is shown in step 0 of figure \ref{FIG:supp1}, where each row is a different gene and each column is a different cell. Throughout this figure, the histogram to the right of each step shows the distribution of entry values for the first column (cell) in the matrix, and the histogram below the step shows the distribution of entry values for the second row (gene Rpl7) of the matrix. The histograms are presented as examples of how the distributions (for each cell and for each gene) of gene expressions change as the data is preprocessed.

In the first step of preprocessing, each cell is normalized independently. This step is taken because we generally assume that every cell will have approximately the same amount of mRNA at any given moment. In other words, we expect the sum gene expression levels to be roughly the same across all cells. This step adjusts for differences in sum gene expression that may be caused by technical errors in the RNA sequencing process, such as some genes being amplified more than others and resulting in a cell appearing to have far more RNA.

We want scTOP to be able to compare cells across sequencing conditions, so we want to transform the normalized RNA counts into quantities that are centered around the mean expression level. Z-scores are exactly such a quantity; they measure the number of standard deviations away from the mean. To calculate the z-scores after step 1, we convert the normalized gene expressions into percentiles by ranking each gene in a given cell from lowest to highest, then dividing by the number of genes (plus one, so that the highest-rank isn't assigned the 100th percentile). For example, if a cell has 1000 genes sampled and Rpl7 is the highest-expressing gene, it will be given rank 1000 and labeled as the 99th percentile. If two or more genes are tied for rank, as is the case for zero-values, they are assigned the average of the ranks they would have been assigned if they were not tied (see "rankdata" function from the Python package Scipy). To convert from percentiles to z-scores, we assumed a normal distribution with mean 0 and standard deviation 1, and applied the corresponding quantile function.

\begin{figure}
	\centering
		\includegraphics[scale=0.6]{figs/fig1a supplement.pdf}
	\caption{Data distributions at each preprocessing step. Step 0 shows the original RNA count matrix before preprocessing, where each entry corresponds to the number of RNA detected for a particular gene in a particular cell. The rows of each matrix correspond to genes and the columns correspond to cells. The histograms to the right of each matrix show the distribution of an example cell, while the histograms below each matrix show the distribution of an example gene. Step 1 of preprocessing is to normalize each cell. Steps 2 and 3 convert the normalized gene expressions to z-scores. Step 2 ranks the genes in each cell according to magnitude and converts the rank into a percentile. Step 3 uses a quantile function, assuming a normal distribution with mean zero and standard deviation one, to convert from percentiles to z-scores.}
	\label{FIG:supp1}
\end{figure}

\subsection{Reference basis construction}
Creating an appropriate reference basis is vital to the accuracy of scTOP. There are many existing scRNA-seq atlases, such as the Mouse Cell Atlas; using them with scTOP requires some amount of curation. For each reference dataset, we took an average over each population corresponding to a given cell type. The average gene expression profiles were preprocessed, and then became the reference profiles for each cell type. Certain cell types from each atlas were dropped for various reasons, such as not including enough cells, being poorly defined (e.g. not enough marker genes to distinguish that cell type from others), etc.

Ensuring that each reference gene expression profile sampled enough cells was essential. This is because the number of cells determines how well-sampled the reference cell type is. Although cell types have distinct transcriptomic patterns, they are not all transcriptomically identical. There are variations in gene expression levels between cells of the same type due to differences in cell-level processes such as cell division or environmental stimuli. Averaging over a population of cells which are some cell type X creates an approximate gene expression profile of the archetypal X cell.

\subsubsection{Mouse Cell Atlas}
The Mouse Cell Atlas (MCA) contains hundreds of thousands of cells from tissues across the mouse body at various stages of development. For the reference basis created from the Mouse Cell Atlas (MCA), we dropped cell types that had fewer than 100 cells. For cell types that had greater than or equal to 100 cells, we averaged across the entire population for each one, then preprocessed. For cell types that were divided according to high expressions of various genes, we combined them into one cell type. For example, the original MCA contained dozens of clusters of mammary gland secretory alveoli cells such as "secretory alveoli cell, Hes1 high" or "secretory alveoli cell, Gpx3 high." These were all combined into one cell type, "mammary gland secretory alveoli." This is because we wanted each reference basis type to act as an archetype rather than a perfect representation of every possible iteration of a cell type; each reference type represents a basin of attraction, and the basins were defined by the overall cell type, not whether individual genes were particularly high within that cell type. We cleaned each of the MCA type labels and ended up with cell types defined by the organ of origin, specific cell type, and time point of collection (e.g. "lung ciliated cell week 6-10"). The time points of cell types in the MCA were either E14.5 or somewhere between week 6 and week 10 post-birth.   The resulting reference basis contained 221 cell types.

\subsubsection{Hematopoietic reference data}
For the hematopoietic dataset, we excluded erythrocytes because the authors used only one marker gene was used to annotate cells of that type. This resulted in many cells having a high erythrocyte score because selecting cells according to the expression of one gene does not sufficiently define the red blood cell population.

{\color{red} MY: Discuss the rest of the reference bases}

\subsection{Accuracy measures}
For table \ref{table:1}, the top1 accuracy was determined by applying scTOP to a cell, finding the type with the highest score, and checking whether that type matched the cell's "true type." Top3 was similarly determined, except instead of just checking the highest-scoring type, we checked whether the true type was included in the top 3 highest-scoring types.

By looking at the scTOP scores for cells whose true type did not match the reference type, we set a score of 0.1 as the lower bound for identifying a type. See figure [REF] for examples of scores where the query type does not match the reference type. In table \ref{table:1}, we labeled cells as "unspecified" if the highest scTOP score was less than 0.1. This provides a useful metric for determining whether a query cell belongs to one of the cell types in the reference basis. If the highest scTOP score is lower than 0.1 for a query cell, this indicates that the reference basis lacks the relevant types or was poorly constructed (see appendix [REF] for a discussion of ill-defined bases).

{\color{red}MY: Using same data as some of the figures in the main paper (ones that don't use the heatmap, since that already shows the irrelevant type scores are low), show that the false positive rate is very low}

scTOP can only identify cells that exist within the reference basis. As such, the accuracy scores were calculated only for the cells whose true type (as determined by the annotations from the authors) was included in the reference basis for the corresponding analysis. For cells whose true types were not represented in the basis, the unspecified rate was very high. This indicates that the rate of false positives is very low.

\subsection{Robustness of scTOP}
scTOP is consistent between iterations of the algorithm, and it is robust to changes in the reference basis. {\color{red}Discussion and figures showing dropping reference types + see how score distributions change}

\subsubsection{Cases where it doesn't work}\label{failure_cases}
The performance of scTOP relies on the quality of the reference basis. scTOP is expected to give inaccurate results in cases where the reference basis is ill-defined. The underlying assumption of the algorithm is that the reference basis consists of accurate gene expression profiles of cell types that are attractor basins. In other words, the reference types are assumed to represent stable cell types rather than cell states that are on the path to the true cell type basins. For example, scTOP performs poorly when using a reference basis made up of embryonic types that are too early in development, because the cell types are not differentiated from one another. Early embryonic cells are not specified yet, so they are not reliable references for cell type.

The reference cell types should refer to attractor basins. The other part of the assumption is that the reference types should be accurate profiles of those attractor basins. As discussed in appendix [REF: section where \# of cells used is discussed], reference types are created from averaging over cells, and the more cells included in this average the better-sampled the type is. As expected, scTOP gives unreliable results when the reference gene expression levels don't accurately reflect the true cell types' gene expression levels. An important factor in the accuracy of scRNA-seq is the amplification step: after RNA is extracted, it is amplified to improve the sensitivity of sequencing, but not all RNA is amplified at the same rate. This results in inflated counts for some genes but not others. Unique molecular identifiers (UMIs) are vital for reducing the inflating effect of amplification [CITE umi paper]. UMIs tag the RNA before amplification, and are present post-amplification; this makes it possible to see which amplified RNA came from the same original molecule, and prevents uneven double-counting of RNA. We have found that scTOP gives more accurate results when used with reference and query datasets that use UMIs, compared to scTOP results using data that did not use UMIs.

\bibliography{export-data.bib}

\end{document}

### 2022-11-30 scTOPv2.tex
manuscript version 2

--- 2022-11-13 scTOPv1.tex
@@
 \affiliation{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
-\author{Darell Kotton}
+\author{Darrell Kotton}
 \affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
 \affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
-
 \author{Pankaj Mehta}
@@
 \section{Introduction}
-Humans and other mammals have hundreds of specialized cells that arise through a complex developmental process  [CITE].
-At early embryonic stages, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers[CITE].
+Humans and other mammals have many specialized cells that arise through a complex developmental process  \cite{zeng_what_2022}.
+At early embryonic stages, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers \cite{Wolpert_2019}.
 However, as embryos mature, cellular identity becomes less plastic and more specified through a complex spatiotemporal process involving signaling and environmental cues.
-The questions of how and why cells end up in one fate or another is a fundamental question in developmental biology [CITE].
+The questions of how and why cells end up in one fate or another are fundamental to developmental biology [CITE].
 More practically, understanding the process of cell fate specification is crucial for developing directed differentiation protocols for engineer cells \emph{in vitro}, with important clinical application for human health and disease [CITE].
@@
 Fourth, the algorithm should measure cell identity in a continuous way instead of assigning discrete identity labels, thus enabling a more subtle and fine-grained analysis of cellular identity and developmental trajectories.
-Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single cell atlases now being generated [].
+Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single cell atlases now being generated [CITE].
 
@@
 
-\subsubsection{Example 1: lung lineages}
+\subsubsection{Example 1: Lung lineages}
 To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from Herriges et al.
@@
 dataset  measures gene expression levels of six different specialized lung types from mice used as control samples.
-See figure \ref{FIG:2} (a).
+See figure \ref{herriges} (a).
 The reference basis used to find the individual scTOP scores for this data set was derived from the Mouse Cell Atlas.
-Despite the fact that the Kotton lab dataset and the Mouse Cell Atlas were created in different laboratories and different sequencing methods, the resulting scTOP scores distinctly separate the cells from the Kotton laboratory.
-The resulting clusters agree with the annotations of the true cell type provided by the Kotton laboratory.
+Despite the fact that the Herriges et al.
+dataset and the Mouse Cell Atlas were created in different laboratories and different sequencing methods, the resulting scTOP scores distinctly separate the cells from the Kotton laboratory.
+The resulting clusters agree with the annotations of the true cell type from Herriges et al.
+
+\begin{figure}
+	\centering
+		\includegraphics[scale=0.45]{figs/herriges colored by type.png}
+        \includegraphics[scale=0.5]{figs/herriges colored by signature.png}
+	\caption{scTOP identifies mouse lung cell types.
+Using the Mouse Cell Atlas as the reference basis, we show that mouse lung types separate clearly on scTOP score axes.
+The data points correspond to individual cells, and in the first row of plots, the marker shape and color indicate the "true type" as determined by annotations from Herriges et al.
+The axes used are scTOP scores for similarity with lung ciliated cells and alveolar types I and II.
+The second, third, and fourth rows of plots have individual cells colored by the marker genes for the type indicated by the color bars at the end of each row.}
+	\label{herriges}
+\end{figure}
 
@@
 The test data, from which the accuracy score was calculated, was the remaining cells of each type not included in the training data.
-Ten different cell types are visualized in a grid in figure \ref{FIG:2} (b).
+Ten different cell types are visualized in a grid in figure \ref{mouse_brain} (a).
 The color of each grid square corresponds to the average scTOP score of cells annotated as the type on the y-axis, with the cell type corresponding to the scTOP dimension on the x-axis.
 As shown by the distinct blue diagonal, the average scTOP score correctly matches the corresponding annotated cell type.
+
+\begin{figure}
+	\centering
+    \includegraphics[scale=0.75]{figs/mouse brain.png}
+	\caption{scTOP identifies mouse brain cell types.
+(a) Heatmap showing the average scTOP score for individual cells of the type indicated on the y-axis, compared to the reference types on the x-axis.
+A subset of cells from the Mouse Brain Atlas are used as the reference basis to query other cells from the same data set.
+The diagonal indicates that scTOP accurately matches query cells to the true reference type.
+(b) Histograms showing the distribution of scTOP scores for individual cells of the type labelling the x-axis.
+The scores shown are of the same type as the sample data.
+The far left histogram shows the distribtion of Astro scores for Astro cells, and the average of this distribution is indicated by the color of the upper left corner box in heatmap (a).}
+	\label{mouse_brain}
+\end{figure}
 
@@
 scTOP correctly identifies individual cells, and it can also identify the tissue to which a population of cells belongs.
-As shown in \ref{FIG:2} (c), we found the aggregate scTOP scores for cells from the Mouse Cell Atlas projected onto the reference basis of Tabula Muris organs \cite{schaum_single-cell_2018}.
+Instead of finding scTOP scores for individual cells, it's possible to group cells and average across them, then find the score for the averaged type.
+These aggregate pseudo-bulk scores tend to be higher than individual cell scores, since averaging over many cells compensates for the dropout effect.
+As shown in \ref{tabula_muris}, we found the aggregate scTOP scores for cells from the Mouse Cell Atlas projected onto the reference basis of Tabula Muris organs \cite{schaum_single-cell_2018}.
 Most cell types were correctly and strongly identified with the annotated tissue of origin.
@@
 	\centering
-		\includegraphics[scale=0.5]{figs/scTOP manuscript fig 2.pdf}
-	\caption{scTOP identifies mouse cell types.
-(a) Using the Mouse Cell Atlas as the reference basis, we show that mouse lung types separate clearly on scTOP score axes.
-The data points correspond to individual cells, and in the first row of plots, the marker shape and color indicate the "true type" as determined by annotations from Herriges et al.
-The axes used are scTOP scores for similarity with lung ciliated cells and alveolar types I and II.
-The second, third, and fourth rows of plots have individual cells colored by the marker genes for the type indicated by the color bars at the end of each row.
-(b) A subset of cells from the Mouse Brain Atlas are used as the reference basis to query other cells from the same data set.
-The shade of each bin corresponds to the average scTOP score for individual cells of the type indicated on the y-axis, compared to the reference types on the x-axis.
-The diagonal indicates that scTOP accurately matches query cells to the true reference type.
-(c) Aggregate samples from the Mouse Cell Atlas are compared to a reference basis consisting of organs from Tabula Muris.
+    \includegraphics[scale=0.5]{figs/tabula muris heatmap.pdf}
+	\caption{scTOP is able to classify tissues at different resolutions.
+Aggregate pseudo-bulk samples from the Mouse Cell Atlas are compared to a reference basis consisting of organs from Tabula Muris.
 The scTOP scores are significantly higher when comparing a cell type with the organ of origin.
-This coarse-grain identification shows scTOP is able to classify tissues at different resolutions.}
-	\label{FIG:2}
+The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis.
+The y-axis lists the true types of the queried cells.
+The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types.}
+	\label{tabula_muris}
 \end{figure}
@@
 200 cells from each line were used as training data to create the reference basis, then the rest of the cells were input as query data.
-As shown in figure \ref{FIG:3} (a) and table \ref{table:1}, scTOP classifies cell line identity of the test data with 100 percent accuracy.
-{\color{red} PM: How did other methods do?}
+As shown in figure \ref{cellbench} and table \ref{table:1}, scTOP classifies cell line identity of the test data with 100 percent accuracy.
+
+\begin{figure}
+	\centering
+		\includegraphics[scale=0.5]{figs/cellbench heatmap.pdf} 
+	\caption{scTOP correctly identifies human tissues.
+The CellBench data set contains human cell cancer lines specifically processed to serve as benchmarking for scRNA-seq algorithms.
+We take a subset of cells to use as the reference basis in analyzing the rest of the CellBench data.
+The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis.
+The y-axis lists the true types of the queried cells.
+The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types.}
+	\label{cellbench}
+\end{figure}
 
@@
 The reference basis was created using 80\% of the cells from each population, and the query data consisted of the remaining 20\% of the data.
-Figure \ref{FIG:3} (b) shows that the scTOP scores were consistently high in cases where the score type matched the true type of the query.
+Figure \ref{human lung} shows that the scTOP scores were consistently high in cases where the score type matched the true type of the query.
 
@@
 	\centering
-		\includegraphics[scale=0.8]{figs/fig3.pdf}
+		\includegraphics[scale=0.5]{figs/human lung heatmap.pdf}
 	\caption{scTOP correctly identifies human tissues.
-(a) The CellBench data set contains human cell cancer lines specifically processed to serve as benchmarking for scRNA-seq algorithms.
-We take a subset of cells to use as the reference basis in analyzing the rest of the CellBench data.
-The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis.
-The y-axis lists the true types of the queried cells.
-The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types.
-(b) Similar to (a), cells from the Human Lung Atlas are compared to a reference basis constructed from cells from the Human Lung Atlas.
+Similar to figure \ref{cellbench}, cells from the Human Lung Atlas are compared to a reference basis constructed from cells from the Human Lung Atlas.
 The colors of the bins correspond to average scTOP scores of individual cells.
 Again, the diagonal is the most prominent section, showing that cells of the y-axis types are correctly matched to the reference x-axis identities.}
-	\label{FIG:3}
+	\label{human lung}
 \end{figure}
@@
 \subsubsection{Lung development}
-{\color{red} PM: We need to add paragraph with some biological conclusions here -- even if already known sighting relevant literature..
-when we see  divergences in lung development ...
-etc.
---- will need final analysis}
-
 The healthy formation of lungs is vital to mammalian life.
@@
 They sequenced mouse lungs between post-conception day 12.5 (E12.5) and postnatal day 42 (P42).
-We found the scTOP scores for cells annotated by Zepp et al.
+As shonw in figure \ref{LungMAP}, we found the scTOP scores for cells annotated by Zepp et al.
 as type I pneumocytes, type II pneumocytes, and epithelial progenitor cells.
 The reference basis consisted of adult cell types from the Mouse Cell Atlas and an E12 epithelial progenitor cell type from the Atlas of Lung Development \cite{negretti_single-cell_2021}.
-In \ref{FIG:4}, we show the scTOP scores for the two adult alveolar types using the Mouse Cell Atlas as the reference basis.
 
-{\color{red} MY: Discuss the clear transition to adult cells, then speculate about the co-expressing middle cells (which exist even at P7).
 Figure 1I from Zepp et al.
 conjectures that one begins to see separate alveolar type 1 and type 2 populations between E15.5 and E17.5.
-According to our plot, there are populations of cells clearly headed towards the two types, but there is also a population of cells in between expressing both type 1 and type 2 signatures.
-Show that it doesn't express Krt8 or the other "transitional cell" genes; it looks more like a spurious attractor/combination of AT1 + AT2}
+According to our plot, at those days there are populations of cells clearly headed towards the two types.
+There is also a population of cells in between expressing both type 1 and type 2 signatures.
+Zepp et al.
+propose these cells are a transitional state similar to the Axin2+ state described by Frank et al.
+[CITE].
+However, as shown in figure \ref{LungMAP genes} (a), the AT1/AT2 cells are not significantly expressing Axin2+.
+Verheyden and Sun [CITE] summarize the variety of previously-described transitional AT1/AT2 states: alveolar differentiation intermediate (ADI) [CITE Strunz], pre-alveolar type-1 transitional cell state (PATS) [CITE Kobayashi], and damage-associated transient progenitors (DATPs) [CITE Choi].
+None of the relevant marker genes for those states were significantly more expressed in the AT1/AT2 cells than either or both of the AT1 or AT2 clusters (figure \ref{LungMAP genes} (b) - (d)).
+Performing differential gene analysis on the clusters also failed to identify any genes that were uniquely expressed in the AT1/AT2 cluster, indicating the cells in this population are truly in a mixed state.
 
 Because scTOP provides us with reference axes that don't change according to whatever data is included, it is possible to track different data points across time on the same axes.
-These axes are also easily interpretable: they measure the similarity to a canonical known cell type.
+These axes are also easily interpretable: they measure the similarity to a known cell type.
 
@@
 	\centering
-		\includegraphics[scale=0.75]{figs/scTOP manuscript fig4.pdf}
-	\caption{Beyond bench-marking: tracing development in cell-type space with scTOP, using developmental mouse lung data.
+		\includegraphics[scale=0.7]{figs/LungMAP.png}
+	\caption{Visualizing the developing mouse lung with scTOP.
 (a) scTOP illustrates specification of alveolar type I and type II cells in murine embryonic development.
@@
 and early epithelial cells from the Atlas of Lung Development.
-The labels next to each data point indicate the age of the mouse embryo from which lungs were extracted, from post-conception day 12 (E12) to day of birth (P0).
-}
-	\label{FIG:4}
+The color and shape of each data point indicates the age of the mouse embryo from which the sample was extracted, from post-conception day 12,5 (E12.5) to 42 days post-birth (P42).
+(b) The same data as in (a) is displayed as a 2D histogram, with each bin colored corresponding to the average day of the data point(s) falling within that bin.
+The color trend shows that early cells generally have very low AT1 and AT2 scores, then the AT1 and AT2 scores increase over time.  (c) Scatter plots of the same data as in (a) and (b), now colored by the average expression of AT1 markers genes (left) and AT2 marker genes (right).
+The scTOP scores of AT1 (AT2) cells are high when the average AT1 (AT2) marker gene expression is high.
+(d) Scatter plots showing the same data as in the above plots, separated by day to show the combination AT1/AT2 population is most apparent at days P3 and P7.}
+	\label{LungMAP}
 \end{figure}
@@
 To apply LARRY to blood cells, they extracted hematopoietic progenitor cells from mice, barcoded the cells, waited for the progenitors to multiply in primary culture, then performed scRNA-seq at days 2, 4, and 6 post-barcoding.
-Then, they extracted blood cells at four and six days after transplantation.
+Then, they extracted blood cells at four and six days in cultre.
 The resulting dataset contains over 300,000 cells, with 2,632 clones spanning time points.
@@
 
-In figure \ref{FIG:5}a, we found the scTOP scores for one particular clonal family that ended up with some sisters that are neutrophils and some that are monocytes.
-The colors and shapes of the points and colors of the histograms indicate the day the cells were extracted.
-The undifferentiated scores are indicated on the y-axis of the bottom row, and by the histograms on the lower right.
-It's clear to see that the cells at day 2 have high undifferentiated scores, and the undifferentiated score decreases as the cells mature -- the histograms for the undifferentiated score move to the left from day 2 to day 6.
-As the cells mature, their scores for neutrophil and monocytes increase.
+We found scTOP scores for clonal families with sisters that spanned multiple days.
+The scores for the undifferentiated hematopoietic progenitor stem cell decreased over time, while the scores for the mature types increased.
+Figure \ref{hem clones}, shows these changes through time for three clonal families.
+
+The clonal family in figure \ref{hem clones} (a) ended up with some sisters that are neutrophils and some that are monocytes.
+As time passes and the cells mature, their scores for neutrophil and monocytes increase.
 The graph with neutrophils on the horizontal axis and monocytes on the vertical axis shows that one subset of the population moved towards high-monocyte, low-neutrophil score and the other subset moved towards low-monocyte, high-neutrophil.
 This indicates that the same progenitor cell can result in either monocytes or neutrophils.
+
+scTOP can reveal differentiation pathways without the need for pseudotime inference.
+Pseudotime inference and other dimensional reduction methods place cells next to one another, which can bias results.
+It's difficult to draw conclusions about certain genes being expressed at different points in pseudotime; patterns may arise that suggest a continuous progression of differentiation, but the progression may be the result of an algorithm inferring that cells which express similar genes should be next to one another in pseudotime, creating a progression by assumption.
+scTOP avoids this issue by comparing cells to a reference instead of grouping them by similarity to one another.
+Weinreb et al.
+claim that the dimensional reduction shown in figure 4 (a) indicates there are two pathways of monocyte differentiation, one that expresses dendritic cell markers and one that expresses neutrophil markers.
+We used scTOP to visualize all the in vitro clones in terms of similarity to neutrophils, monocytes, and progenitors (\ref{hem genes}).
+This allows us to see which cells are more likely to be neutrophils that exhibit some monocyte-like genes rather than neutrophil-like monocytes.
+However, one limitation of scTOP is that it requires sufficient cells to create a reliable reference from; since there were too few dendritic cells to serve as reference, it was not possible to calculate dendritic cell scores.
 
@@
 	\centering
-		\includegraphics[scale=0.75]{figs/scTOP manuscript fig5.pdf}
-	\caption{Beyond bench-marking: tracing development in cell-type space with scTOP, using lineage tracing hematopoietic mouse data. (a) {\color{red}} (Describe plots) }
-	\label{FIG:4}
+		\includegraphics[scale=0.75]{figs/hem. clone families.png}
+	\caption{Tracing hematopoietic cells differentiation with scTOP.
+(a), (b), (c) display scatter plots and distributions of scTOP scores for individual in vitro clone families from Weinreb et al.
+The scatter plots show the x-axis scTOP score compared to the y-axis scTOP score for individual cells, while the distributions show the scTOP scores of the type indicated on the x-axis of the corresponding column.
+The color and shape of each marker (and color of each distribution) indicate the days in culture at which the cells were extracted.
+(a) A clone family where some clones end up becoming neutrophils and some become monocytes.
+At day 2, the 3 sister cells have low neutrophil and monocyte scores, and high progenitor scores.
+At days 4 and 6, the progenitor scores decrease and a portion of the cells increase in monocyte scores while others increase in neutrophil scores.
+(b) A clone family where all of the cells end up becoming neutrophils.
+(c) A clone family where the descendant cells become neutrophils, eosiniphils, and basophils.}
+	\label{hem clones}
+\end{figure}
+
+\begin{figure}
+	\centering
+		\includegraphics[scale=0.75]{figs/hem progenitor hists.png}
+	\caption{Tracing differentiation of hematopoietic cells from progenitor to mature type.
+The plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al., with the x-axis of each one indicating the scTOP score for hematopoietic progenitor type and the y-axis showing the score for a mature type.
+Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin.
+Early cells have high progenitor scores and low mature type scores.
+As time passes cells end up with high scores in mature types.}
+	\label{hem progenitor}
+\end{figure}
+
+\begin{figure}
+	\centering
+		\includegraphics[scale=0.59]{figs/hem mature hists.png}
+	\caption{Comparing differentiation of hematopoietic cells between mature types.
+As in figure \ref{hem progenitor}, the plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al.
+The x-axis and y-axis of each histogram indicate the scTOP score of the labelled mature type.
+Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin.  Some pairs of types have cells with high scores of both types.
+This is apparent in the neutrophil-monocyte histogram, which presents a continuous spectrum of scores from the horizontal axis to the vertical.
+Other pairs of types only have cells expressing one type or the other, as in the L-shaped neutrophil-megakaryocyte histogram.}
+	\label{hem mature}
 \end{figure}
@@
 \section*{Acknowledgments}
-We acknowledge useful discussions with Jason Rocks, Robert Marsland, and Jason Rocks.
+We acknowledge useful discussions with Jason Rocks and Robert Marsland.
 The work was funded by a grant to the Boston University Kilachand Multicellular Design Program (to DK and PM) and NIH NIGMS 1R35GM119461 to PM.
-We also acknowledge programming advice from Alena Yampolskaya.
+We also acknowledge programming support from Alena Yampolskaya.
 
@@
 
-\subsection{Reference basis construction}
+\subsection{Reference basis construction}\label{basis construction}
 Creating an appropriate reference basis is vital to the accuracy of scTOP.
@@
 By looking at the scTOP scores for cells whose true type did not match the reference type, we set a score of 0.1 as the lower bound for identifying a type.
-See figure [REF] for examples of scores where the query type does not match the reference type.
+As shown in figure \ref{robustness hists} (c) and (f), when the reference type does not match the sample type the scores for individual cells fall well within [-0.1, 0.1], even when most of the reference basis is missing.
+
 In table \ref{table:1}, we labeled cells as "unspecified" if the highest scTOP score was less than 0.1.
 This provides a useful metric for determining whether a query cell belongs to one of the cell types in the reference basis.
-If the highest scTOP score is lower than 0.1 for a query cell, this indicates that the reference basis lacks the relevant types or was poorly constructed (see appendix [REF] for a discussion of ill-defined bases).
-
-{\color{red}MY: Using same data as some of the figures in the main paper (ones that don't use the heatmap, since that already shows the irrelevant type scores are low), show that the false positive rate is very low}
+If the highest scTOP score is lower than 0.1 for a query cell, this indicates that the reference basis lacks the relevant types or was poorly constructed (see appendix \ref{failure_cases} for a discussion of ill-defined bases).
 
@@
 scTOP is consistent between iterations of the algorithm, and it is robust to changes in the reference basis.
-{\color{red}Discussion and figures showing dropping reference types + see how score distributions change}
+As shown in figure \ref{robustness hists}, many scTOP scores do not change significantly even when most of the reference basis is removed.
+Generally, scores for cells that match the reference type do not change much even when only 25\% of the original basis is retained (figure \ref{robustness hists} (d), (e)).
+The largest difference occurs for cells that are similar to the reference type, such as the AT1 score for AT2 cells (figure \ref{robustness hists} (g)).
+These scores tend to increase, because removing other types from the basis reduces the ability to de-correlate effectively.
+In other words, scTOP is more likely to confuse similar types when it has fewer reference types to compare.
+This effect is most apparent when relevant types are removed from the basis.
+Figures \ref{robustness hists} (j) and (k) show that all lung types score higher for AT1 and AT2 when only lung alveolar types are included in the basis.
+Figures \ref{robustness hists} (c) and (f) show that irrelevant scores, such as bladder and thymus cells, do not change much even when large portions of the basis are removed.
+scTOP does not confuse sample cells for types that are dissimilar even with a small reference basis because there is no need to de-correlate between types that are naturally not correlated.
 
@@
 The other part of the assumption is that the reference types should be accurate profiles of those attractor basins.
-As discussed in appendix [REF: section where \# of cells used is discussed], reference types are created from averaging over cells, and the more cells included in this average the better-sampled the type is.
+As discussed in appendix \ref{basis construction}, reference types are created from averaging over cells, and the more cells included in this average the better-sampled the type is.
 As expected, scTOP gives unreliable results when the reference gene expression levels don't accurately reflect the true cell types' gene expression levels.
@@
 
+\begin{figure}
+	\centering
+		\includegraphics[scale=0.8]{figs/LungMAP genes.png}
+	\caption{The AT1/AT2 combination state (most apparent in days P3 and P7) do not uniquely express genes of AT1/AT2 transitional states described by previous papers.
+Alveolar cells from Zepp et al.
+are plotted on AT1 and AT2 axes, separated by day.
+Each cell is colored by the expression z-score of a particular gene, and the particular gene is different in each row.
+The gene used to color the cells is indicated by the color bar on the far right of each row.
+(a) Alveolar cells colored by Axin2+, showing that the AT1/AT2 combination cells do not appear to belong to the Axin2+ sublineage described by [CITE].
+(b) Alveolar cells colored by Krt8 and Trp53 expression, which are genes corresponding to alveolar differentiation intermediate cells (ADI) described by [CITE].
+AT1, AT2, and the combination AT1/AT2 state all express Krt8 and Trp53 at similar levels.
+(c) Alveolar cells colored by expression of pre-alveolar type-1 transitional cell state (PATS) genes [CITE].
+AT1, AT2 and the combination AT1/AT2 state all express Cldn4 at similar levels.
+AT1 and AT1/AT2 combination cells both express Krt19 at similar levels, although AT1 cells express the gene at higher levels.
+(d) Alveolar cells colored by expression of damage-associated transient progenitor (DATP) genes [CITE].
+AT1, AT2, and the combination AT1/AT2 state all express Il1r1 and HIF1a at similar levels.
+}
+	\label{LungMAP genes}
+\end{figure}
+
+\begin{figure}
+	\centering
+		\includegraphics[scale=0.72]{figs/hem genes.png}
+	\caption{Scatter plots of in vitro hematopoietic clone families from Weinreb et al., with scTOP scores for progenitor, neutrophil, and monocyte types, and colored by normalized gene expression.
+In Weinreb et al.
+figure 4 (a), cells visualized using pseudotime inference were separated into branches.
+The monocyte branch appeared to have half of its cells expressing dendritic cell genes and the other half expressing neutrophil genes.
+This figure plots those same cells on progenitor, neutrophil, and monocyte axes, and colors according to neutrophil, dendritic cell, and monocyte marker genes.
+(a) Hematopoietic cells colored according to expression of neutrophil marker genes S100a8 and Elane.
+(b) Hematopoietic cells colored according to expression of dendritic cell marker genes Cd74 and H2-Aa.
+(c) Hematopoietic cells colored according to expression of monocyte marker genes Csf1r and Klf4. }
+	\label{hem genes}
+\end{figure}
+
+\begin{figure}
+	\centering
+		\includegraphics[scale=0.82]{figs/robustness hists.png}
+	\caption{Distributions of scTOP scores for lung cells from Herriges et al., using varying portions of the Mouse Cell Atlas as the basis.
+the label above each plot indicates the Herriges et al.
+sample subset, while the x-axis label indicates the scTOP reference type.
+"All lung types" means that all of the lung types from the data set were included.
+(a) - (i) show how scTOP scores change when 25-100\% of the basis is used.
+(j) - (k) show how scTOP scores change when the basis is changed from the full basis (which includes many tissues and organs), to just the lung types, to just the two alveolar types.
+(c) and (f) show scTOP scores for random irrelevant types, Bladder Epithelium and Thymus B Cell, to show that the scTOP scores remain low for types that don't match the sample.}
+	\label{robustness hists}
+\end{figure}
+
 \bibliography{export-data.bib}

### 2023-01-25 manuscript/scTOPv5.tex
added notebooks and manuscript v5

	\documentclass[aps,superscriptaddress, notitlepage,longbibliography]{revtex4-1}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{physics}
\usepackage[pdftex]{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\setlength {\marginparwidth }{2cm}
\hypersetup{
    colorlinks,
    citecolor=blue,
    filecolor=black,
    linkcolor=red,
    urlcolor=blue
}

\newcommand{\comment}[1]{{\color{blue}#1}}
\newcommand{\edit}[1]{{\color{purple}#1}}
\newcommand{\del}[1]{{\color{red}\st{#1}}}
\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }
% Standardize command font styles and environments
\newcommand{\doccmd}[1]{\texttt{\textbackslash#1}}% command name -- adds backslash automatically
\newcommand{\docopt}[1]{\ensuremath{\langle}\textrm{\textit{#1}}\ensuremath{\rangle}}% optional command argument
\newcommand{\docarg}[1]{\textrm{\textit{#1}}}% (required) command argument
\newcommand{\docenv}[1]{\textsf{#1}}% environment name
\newcommand{\docpkg}[1]{\texttt{#1}}% package name
\newcommand{\doccls}[1]{\texttt{#1}}% document class name
\newcommand{\docclsopt}[1]{\texttt{#1}}% document class option name
\newenvironment{docspec}{\begin{quote}\noindent}{\end{quote}}% command specification environment

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

\definecolor{pastel_green}{rgb}{0.18,0.65,0.34}
\newcommand{\CHW}[1]{\textcolor{pastel_green}{#1}}
%% OPTIONAL MACRO DEFINITIONS
\def\s{\sigma}

\def \M{\mathcal{M}}  
\def \P{\mathrm{P}}
\def \E{\mathbb{E}}

\def \mcl{\mathcal}
\def \mbb{\mathbb}
\def \mbf{\mathbf}

\def\bd{\boldsymbol}
\def\sm{\setminus}
\def\tb{\textbf}

\def\CH{\cal{H}}

\def\<{\langle}
\def\>{\rangle}

\def\thet{{\theta}^{(t)}}
\def\thetp{{\theta}^{(t+1)}}
\def\thets{{\theta}^{(t)\star}}

\def\be{\begin{equation}} 
\def\ee{\end{equation}}
\newcommand \bea {\begin{eqnarray}} 
\newcommand \eea {\end{eqnarray}} 
\newcommand \sign {\hbox{sign}} 
\newcommand{\nn} {\nonumber}
\newcommand{\mbbE} {\mathbb{E}}
\begin{document}

\title{ scTOP: physics-inspired order parameters for cellular identification and visualization}

\author{Maria Yampolskaya}
\email{mariay@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\author{Michael Herriges}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
\author{Laertis Ikonomou}
\affiliation{Department of Oral Biology. University at Buffalo, The State University of New York, Buffalo, NY, USA}
\affiliation{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
\author{Darrell Kotton}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
\author{Pankaj Mehta}
\email{pankajm@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{Faculty of Computing and Data Science, Boston University, Boston, MA 02215, USA}
\affiliation{Biological Design Center, Boston University, Boston, MA 02215, USA}

\begin{abstract}
Advances in single-cell RNA-sequencing (scRNA-seq) provide an unprecedented window into cellular identity. The increasing abundance of data requires new theoretical and computational frameworks for understanding cell fate determination, accurately classifying cell fates from expression data, and integrating knowledge from cell atlases. Here, we present single-cell Type Order Parameters (scTOP): a statistical-physics-inspired approach for constructing ``order parameters'' for cell fate given a reference basis of cell types. scTOP can quickly and accurately classify cells at a single-cell resolution, generate interpretable visualizations of developmental trajectories, and assess the fidelity of engineered cells. Importantly, scTOP does this without using feature selection, statistical fitting, or dimensional reduction (e.g., UMAP, PCA, etc.). We illustrate the power of scTOP utilizing a wide variety of human and mouse datasets (both {\it in vivo} and {\it in vitro}). By reanalyzing mouse lung alveolar development data, we characterize a transient perinatal hybrid alveolar type 1/alveolar type 2 (AT1/AT2) cell population that disappears by 15 days post-birth and show that it is transcriptionally distinct from previously identified adult AT2-to-AT1 transitional cell types. Visualizations of lineage tracing data on hematopoiesis using scTOP confirm that a single clone can give rise to as many as three distinct differentiated cell types. We also show how scTOP can quantitatively assess the transcriptional similarity between endogenous and transplanted cells in the context of murine pulmonary cell transplantation. Finally, we provide an easy-to-use Python implementation of scTOP. Our results suggest that physics-inspired order parameters can be an important tool for understanding development and characterizing engineered cells.
\end{abstract}

\maketitle

\section{Introduction}
Humans and other mammals have many specialized cells that arise through a complex developmental process  \cite{zeng_what_2022}. At the blastocyst/epiblast stages of embroynic development, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers \cite{Wolpert_2019}. However, as embryos mature, cellular identity becomes less plastic and more specified through a complex spatiotemporal process involving signaling and other environmental cues. The questions of how and why cells end up in one fate or another are fundamental to developmental biology \cite{stephenson_intercellular_2012, zhu2020principles, rand2021geometry}. More practically, understanding the process of cell fate specification is crucial for developing directed differentiation protocols for engineering cells \emph{in vitro}, with important clinical applications for human health and disease \cite{rossant_impact_2011, rowe2019induced, basil2020cellular}.

One powerful set of techniques for characterizing cellular phenotype is transcriptome-wide measurement using single-cell RNA sequencing (scRNA-seq). Typical scRNA-seq experiments involve thousands of genes and dozens to millions of cells. For this reason, this data is intrinsically high-dimensional and requires specialized computational tools for analysis and interpretation \cite{zhu2020principles, setty2019characterization, schiebinger2019optimal}. This task is especially difficult because the technical challenges of sequencing RNA from individual cells result in data that is noisy and sparse \cite{hicks2018missing, lahnemann_eleven_2020, argelaguet2021computational}. Genes with relatively low expression are often recorded as not being expressed at all, resulting in zero counts called dropouts. Experiment-to-experiment differences in sequencing methods, platforms, and other difficult-to-standardize conditions result in technical variations in scRNA-seq measurements called batch effects. These challenges merit careful consideration of how to normalize and analyze the data in a way that permits the study of biological variation without conflation by technical variation.

scRNA-seq is often used for cellular identification. The most frequently used workflow involves applying dimensional reduction (such as Uniform Manifold Approximation and Projection (UMAP) \cite{mcinnes_umap_2018} or SPRING \cite{weinreb_spring_2018}) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells. These groups are then often manually annotated according to expression levels of known marker genes. Machine learning classifiers such as support vector machines are also commonly used to identify cells after training with labeled data \cite{abdelaal_comparison_2019}. Another common analysis task is trajectory inference, where the goal is to infer developmental trajectories from static time shots (reviewed in \cite{saelens2019comparison,wagner2020lineage}). Trajectory inference arranges cells on a low-dimensional manifold and assigns pseudo-times according to transcriptomic similarity. There are also methods such as partition-based graph abstraction \cite{wolf_paga_2019} that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions. RNA-velocity-based methods like Dynamo \cite{qiu2022mapping} use splicing information or time-resolved data to predict paths of differentiation from RNA data.

Despite the power and utility of these computational tools, most commonly employed methods for analyzing scRNA-seq data share certain limitations. Dimensional reduction is often one of the first steps in visualizing scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships \cite{chari_specious_2021}. Distorted embeddings inconsistently represent distances between individual data points and between cell types, making it difficult to interpret cell fate transitions. Similar issues arise in trajectory inference and RNA-velocity methods because local neighborhoods can also vary substantially with the choice of embedding. Another limitation of these methods is the lack of reproducibility inherent to stochastic machine learning methods like t-SNE and UMAP, which output different results on separate executions even when hyperparameters remain the same \cite{wattenberg_how_2016}. Many algorithms also perform feature selection (e.g., limiting their analysis to the most variable genes or highly-expressed genes) or require choosing parameters such as perplexity, making results even more sensitive to the inclusion or exclusion of different datasets in the analysis.

An ideal analysis algorithm for scRNA-seq data should possess certain desirable properties. First, the algorithm should incorporate explicit assumptions that can be kept consistent between analyses. Second, the algorithm should be deterministic, further enhancing reproducibility. Third, the axes in visualization should be interpretable, intuitive, and remain fixed across analysis and datasets. Fourth, the algorithm should assign a continuous value to cell identity instead of assigning discrete identity labels, thus enabling a more subtle and fine-grained analysis of cellular identity and developmental trajectories that reflect the continuous nature of biology. Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single-cell atlases now being generated \cite{quake2022decade}.

Here, we present a new physics-inspired algorithm, single-cell Type Order Parameters (scTOP), that has all of these desired properties. The algorithm is based on the statistical physics concept of an order parameter -- a coarse-grained quantity that can distinguish between phases of matter. Arguably the most famous example of an order parameter is magnetization in a ferromagnetic system (i.e., a system that can spontaneously magnetize). In a ferromagnet, the magnetization measures the alignment of spins and can be used to distinguish magnetic and paramagnetic phases \cite{landau2013statistical,sethna2021statistical}. It is straightforward to generalize the idea of magnetization to more complex settings such as spin glasses and attractor neural networks \cite{amit1985spin}. In the context of epigenetic landscape models of cell identity, this takes the form of generalized magnetizations that measure how aligned a gene expression state is with each of the attractor basins for different cell fates \cite{lang_epigenetic_2014,pusuluri_cellular_2017}. These generalized magnetizations serve as natural order parameters for cellular identity and can be calculated directly from data \cite{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}. The scTOP algorithm builds on previous work by developing methods for integrating scRNA-seq data, greatly expanding the power and utility of this approach.

The inputs to scTOP are the scRNA-seq measurements of query data as well as a reference basis containing the archetypal gene expression profiles for cell types of interest. The outputs of scTOP are the coordinates of the query data in cell type order parameter space (i.e., generalized magnetizations), with each output dimension serving as a measure of similarity between the query data and a distinct cell type in the reference basis. In this way, scTOP naturally projects gene expression deterministically onto a space with dimension equal to the number of cell types in the reference basis. These generalized coordinates can be used for both cell type identification and visualization. One appealing feature of scTOP is that the assumptions of the algorithm are explicit and are contained entirely in the choice of reference basis. This allows direct comparison across datasets and settings since the coordinates of and distances between data points do not change as long as one uses the same reference basis. One important limitation of scTOP worth noting is that it cannot be easily used to identify novel cell fates. The underlying reason for this is that scTOP assumes one already knows what cell types are of interest in the form of a reference basis and, for this reason, is less suited for exploratory work with undocumented cell types.

In this paper, we show that scTOP can be used to reliably assess cell identity and accurately classify cell type with high accuracy at the resolution of individual cells. We demonstrate this using mouse and human samples from lungs, brains, and other organs, including transplanted cells that were generated through the directed differentiation of pluripotent stem cells (PSCs). By analyzing lineage tracing data and embryonic development data, we also establish scTOP's capacity to quantify transitions between cell types, including bifurcations between closely related cell fates. We also show that scTOP can be used to quantitatively assess transcriptional similarity of engineered and natural cells by analyzing murine pulmonary transplantation experiments.

We have implemented scTOP in an easy-to-use Python package, available on the \href{https://pypi.org/project/scTOP/}{Python Package Index} and \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP}{Github}. The code written for the analyses in this paper is available on \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP-manuscript}{Github}.

\section{Methods}
\begin{figure}
	\centering
		\includegraphics[scale=0.8]{figs/scTOP manuscript fig1 - no aggregate.pdf}
	\caption{Steps involved in the scTOP algorithm. (a) Each cell in the query data is preprocessed independently of the other cells in the dataset. (b) scRNA-seq atlases, such as the Mouse Cell Atlas, are used to define the reference basis in the algorithm. (c) scTOP scores are the projections of query data onto the non-orthogonal subspace of cell types. Since there is no statistical fitting, no tuning parameters are involved. (d) The process of finding scTOP scores is shown in the grey section, and the blue section illustrates cell-type space. This space has as many dimensions as there are cell types in the reference basis. We can visualize this concept in 1, 2, or 3 dimensions, depending on which cell types are most relevant. In the 3-dimensional case, the shadows on the bounds of the scatter plot are included to better illustrate the position of points in 3-dimensional space.}
	\label{FIG:1}
\end{figure}

\subsection{Mathematical Details}
To calculate cell type similarity scores, scTOP uses a linear algebra projection method inspired by the concept of cell types as attractors in a dynamical system \cite{huang_cell_2005}. This projection method has been applied previously to bulk RNA-seq samples, and we will summarize the theoretical background here \cite{lang_epigenetic_2014}. A cell type is a state of the gene regulatory network. A natural way to represent attractors in this system is with an attractor neural network, where attractors correspond to different cell types. If we represent each gene as a node in a network with an associated value measuring gene expression (positive value representing high gene expression, negative value indicating low gene expression), then a cell type may be denoted by a vector of gene expression values $S_i$, where $i=1,\ldots, G$ spans the $G$ genes being measured. The gene expression profiles
corresponding to the $C$ cell type attractors are also $G$ dimensional vectors $\xi_i^\mu$, where $\mu=1,\ldots, C$ spans the $C$ cell types being stored in the network. In what follows, we assume that $G > C$; in other words, the number of genes being measured is greater than the number of cell types.

Cell types are often highly similar in their patterns of gene regulation. Kanter and Sompolinsky \cite{kanter_associative_1987} defined a pattern storage method for spin-glass-like neural networks where even correlated patterns robustly act as attractors. For the same system, they also defined order parameters: generalized magnetizations $a^{\mu}$, where $\mu$ iterates through each of the stored cell type attractors. The $a^\mu$ can be understood as de-correlated versions of the conventional spin glass magnetization order parameter that measures the similarity between a given network state and the network state $\xi^\mu$. 
Explicitly, the $C$ generalized magnetizations $a^\mu$ can be calculated for a gene expression state $S_i$ via the expression
\be \label{op_eqn}
a^\mu= \sum_{\nu=1}^C\sum_{j=1}^G [A_{\mu \nu}]^{-1}\xi_j^\nu S_j.
\ee
where
\be
A_{\mu \nu}= \sum_{j}^G  \xi_j^\mu \xi_j^\nu,
\ee
is the overlap matrix of gene expression profiles of different cell types. Previous works have shown the order parameters $a^{\mu}$ provided an excellent similarity score for analyzing bulk RNA-seq data \cite{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}. However, bulk RNA-seq measurements average gene expression over many cells, rendering it impossible to measure cell fate transitions beyond an average tissue state.

scTOP uses the same order parameters, $a^{\mu}$, to track cell type at the resolution of individual cells, making it possible to directly observe cells in different stages of differentiation. In both attractor neural networks and previous applications to bulk RNA-seq data, the expression vectors $S_i$ were binary variables. However, we have found that for working with scRNA-seq data, it is helpful to treat $S_i$ as continuous.

\subsection{Preprocessing}

As shown in figure \ref{FIG:1} (a), the first stage of the algorithm is preprocessing and normalization of the input data. In our algorithm, measurements from each cell are normalized independently and do not depend on the expression profiles of any other cells in the dataset. We assume that the scRNA-seq data has been processed, resulting in a gene-wise count matrix: for each cell, each gene has an integer value that counts the number of assigned RNA reads. Genes are then rank-ordered. The rank order is subsequently converted to a z-score representing the percentile rank of the expression of a gene within the cell relative to all other genes being measured (assuming a normal distribution with mean zero and standard deviation one, for the sake of simplicity). For example, a gene that has higher expression than $97.8\%$ of genes being measured is assigned a score of $z=2$, while a gene that has higher expression than half of the genes is assigned a score of $z=0$. The details of the process and its effects on data distributions are described in Appendix section \ref{preprocessing}. The preprocessing step is a key component of the algorithm. By processing each cell independently, the output for one cell is independent of other cells included in the analysis. This is in contrast with other algorithms which normalize gene expression across cells by, for example, selecting the most variable genes.

\subsection{Construction of reference basis}

Constructing a representative, accurate reference basis for the data being analyzed is vital to the scTOP algorithm. This process is shown in figure \ref{FIG:1} (b). First, relevant single-cell RNA-seq atlases are gathered to be processed. For example, in analyzing mouse lung cells, we used the Mouse Cell Atlas \cite{noauthor_mapping_nodate}, which contains single-cell samples of mouse tissues across all major organs. Once the relevant atlases are identified, we take an average across each cell population that corresponds to a distinct cell type. To mitigate the potent effects of noise and avoid training data imbalances, a sufficient number of cells must be included in each population. We have empirically found that the minimum sample size to achieve reasonable results is 100-200 cells. Noise affects the algorithm most strongly when it appears in the basis since the decorrelation operation involved in the projection separates cells according to the canonical expression levels included in the reference basis. After each cell population is averaged to create the gene expression profiles for the desired cell types, each cell type profile is preprocessed separately using the same procedure as that used for processing single cells: normalization followed by z-scoring. This creates a modular reference basis where cell types can easily be replaced, removed, or added at will.

\subsection{Calculation of scTOP scores}

Given a reference basis, we can calculate the order parameters $a^\mu$ using Eq.~\ref{op_eqn}. Although these order parameters were originally inspired by attractor neural networks and spin glass physics, the $a^{\mu}$ can also be understood as non-orthogonal projections onto the subspace spanned by the reference basis (i.e., cell type space). This is illustrated in figure \ref{FIG:1} (c). A single-cell RNA-seq measurement of a sample can be thought of as a vector in $G$-dimensional gene expression space, where $G$ is the number of genes. The $C$-cell types in the reference basis are also each represented by a $G$-dimensional vector and form a non-orthogonal basis for the $C$-dimensional cell type space (we assume $C < G$). To find the coordinates $a^\mu$ in cell type space, we project the sample vector onto the reference basis using Eq.~\ref{op_eqn}.

Mathematically, we can always write the sample vector as a sum of the projected components and a component $S^{\perp}_i$  perpendicular to the cell type subspace: 
\be
S_i = \sum_\mu a^{\mu} \xi^{\mu}_i + S^{\perp}_i.
\ee
Biologically, $S^{\perp}$ contains information about biological processes that introduce variation in transcriptional profile such as cell cycle as well as expression resulting from technical and biological noise. We refer to the $a^\mu$ as the scTOP scores. Each component of this vector measures the projection of the cell with gene expression profile $S_i$ onto the $\mu$-the cell type in the reference basis. These scores can be used to accurately classify cell identity and provide a natural visualization of gene expression in the space of possible cell fates (see figure \ref{FIG:1} (e)).

In general, we are interested in calculating scTOP scores for individual cells. However, it is also sometimes useful to compute ``aggregate'' scTOP scores for cellular populations by averaging the mRNA counts \textit{first}, then preprocessing the data. This produces a gene expression profile and a set of scTOP aggregate scores for each population. Aggregate scores tend to be higher because averaging over the RNA counts mitigates the well-documented effects of scRNA-seq noise \cite{hicks2018missing, lahnemann_eleven_2020}; the averaged gene expression profile for the populations are not as sparse as individual profiles and thus are more similar to the averaged profiles that form the reference basis.

 In the appendix, we give a detailed discussion of scTOP, including the effects of basis choice, score distributions for correct and incorrect cell types, and other technical details.

\subsection{Assumptions and computational complexity}

scTOP does not have any stochasticity, require tuning of hyper-parameters, or statistical fitting procedures. The assumptions of the algorithm are explicit: that the gene expression profiles included in the reference basis are truly representative of the relevant cell types and that the reference cell types correspond to distinct cellular populations. We have found that constructing a reference basis for a cell type typically requires at least 100 cells. Additionally, if the cell types are not sufficiently different from one another, as in early embryonic development or certain stromal cells where the same cell type is compared between organs, the resulting scTOP scores may be unreliable and exhibit extreme sensitivity to small changes in the choice of reference basis. More information on the limitations of scTOP may be found in the appendix. Generally, outside these very limited edge cases, we have found that scTOP can reliably distinguish even extremely closely related cells (as shown in the example of pulmonary alveolar epithelial type 1 (AT1) and type 2 (AT2) cell lineages below).

Finally, we note that scTOP is extremely computationally efficient. The core of the scTOP is equation \ref{op_eqn}. This requires a single matrix inversion involving the reference basis that can be precomputed. For this reason, the computational complexity of scTOP scales \emph{linearly} with the number of cells. The end result is that scTOP takes milliseconds to run, even for very large datasets. 
\section{Results}

\subsection{Benchmarking and validation: Classification of cell fate}

scTOP scores can be used to reliably and accurately predict cell identity. To show this, we applied the algorithm to several scRNA-seq datasets across species and laboratories. scTOP is organism-agnostic as long as an appropriate reference basis is used. Mice and humans are among the most common subjects of scRNA-seq measurements, and we restrict our analysis in this manuscript to data from these organisms. To do so, we make use of single-cell atlases, such as the Mouse Cell Atlas \cite{noauthor_mapping_nodate} and the Atlas of Lung Development \cite{negretti_single-cell_2021}.

Table \ref{table:1} lists each of the datasets examined in this paper with their respective scTOP score accuracies. TopN refers to the percent of cells whose true cell types were in the top N scTOP scores, with a score greater than 0.1. Unspecified refers to the percent of cells for which the highest scTOP score was less than 0.1; this represents a failure of the algorithm to confidently identify the cell. The accuracies across species and tissues are high, and the unspecified percentages are very low, even in datasets exceeding a million cells.

\begin{table}[h!]
\centering
\begin{tabular}{| c | c | c | c | c | c | c |}
\hline
Query Data & Species & Reference Basis & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Total cells \\ 
\hline
Kotton lab & Mouse & Mouse Cell Atlas & 97.19 & 99.77 & 2.69 & 4,805 \\ 
\hline
Brain atlas & Mouse & Self & 86.28 & 96.89 & 1.78 & 1,161,041 \\
\hline
CellBench & Human & Self & 100 & 100 & 0 & 2,822 \\
\hline
Lung atlas & Human & Self & 79.78 & 85.16 & 3.32 & 2,952 \\  
\hline
\hline
Data set & Species & Reference Basis & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Number of cell types \\
\hline
Mouse Cell Atlas & Mouse & Tabula Muris & 79.17 & 87.5 & 0 & 48 \\
\hline

\end{tabular}
\caption{Accuracy scores of included sample types for each of the described data sets.}
\label{table:1}
\end{table}

scTOP performs comparably to state-of-the-art methods for cell identification. Abdelaal et al. compared the performance of automatic cell identification algorithms for various datasets, including the CellBench dataset and an older version of the Allen Mouse Brain atlas \cite{abdelaal_comparison_2019}. They found that the median F1-score for the CellBench dataset ranged from 0.9 to 1; scTOP has a median F1-score of 1 for the same data. The version of the Allen Mouse Brain atlas they analyzed contained three levels of cell population annotations, with 3, 16, and 92 populations for each annotation level. The median F1-score ranged from 0.64 to 1 for the 16-population case and 0 to 0.98 for the 92-population case. The version of the brain atlas we analyzed contained 40 populations, and scTOP had a median F1-score of 0.88 for this dataset.

\subsubsection{Example 1: Lung lineages}
To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from Herriges et al. \cite{herriges_durable_2022} onto the mouse cell atlas. The Herriges et al. dataset measures gene expression levels of six different specialized lung types from healthy mice: AT1, AT2, basal, ciliated, secretory, and neuroendocrine (See figure \ref{herriges} (a)). The reference basis used to find the individual scTOP scores for this data set was derived from the Mouse Cell Atlas. Even though the Herriges et al. dataset and the Mouse Cell Atlas were created in different laboratories using different sequencing methods, the resulting scTOP scores distinctly separate the Herriges et al. dataset into the expected cell types. Furthermore, the individual scTOP scores strongly correlate with the expression of lineage marker sets (figure \ref{herriges} (b)).  Altogether this suggests that scTOP provides a reliable method to annotate scRNA-seq datasets without the need for the expertise required to generate lineage marker gene sets.

\begin{figure}
	\centering
		\includegraphics[scale=0.45]{figs/herriges colored by type.png}
        \includegraphics[scale=0.5]{figs/herriges colored by signature.png}
	\caption{scTOP identifies mouse lung cell types. Using the Mouse Cell Atlas as the reference basis, we show that mouse lung types separate clearly on the scTOP score axes. The data points correspond to individual cells, and in the first row of plots, the marker shape and color indicate the true cell type as determined by annotations from Herriges et al. The axes used are scTOP scores for similarity with lung ciliated cells and alveolar types 1 and 2. The second, third, and fourth rows of plots have individual cells colored by the marker genes for the type indicated by the color bars at the end of each row (see appendix \ref{alveolar markers} for details on marker genes used).}
	\label{herriges}
\end{figure}

\subsubsection{Example 2: Mouse Brain Atlas}

The mouse brain atlas \cite{yao_taxonomy_2021} sequenced over one million cells and clustered them into 42 subclasses and 101 supertypes. For a reference basis, we reserved a subset of 200 cells from each of the identified subclasses to use as training data. The test data, from which the accuracy score was calculated, was the remaining cells of each type not included in the training data. Ten different cell types are visualized in a grid in figure \ref{mouse_brain} (a). The color of each grid square corresponds to the average scTOP score of cells annotated as the type on the y-axis, with the cell type corresponding to the scTOP dimension on the x-axis. As shown by the distinct blue diagonal, the average scTOP score correctly matches the corresponding annotated cell type. This demonstrates that scTOP is able to distinguish cell fates even from closely related neural lineages.

\begin{figure}
	\centering
    \includegraphics[scale=0.75]{figs/mouse brain.png}
	\caption{scTOP identifies mouse brain cell types. (a) Heatmap showing the average scTOP score for individual cells of the type indicated on the y-axis, compared to the reference types on the x-axis. A subset of cells from the Mouse Brain Atlas is used as the reference basis to query other cells from the same data set. The diagonal indicates that scTOP accurately matches query cells to the true reference type. (b) Histograms showing the distribution of scTOP scores for individual cells of the type labeling the x-axis. The scores shown are of the same type as the sample data. The far left histogram shows the distribution of Astro scores for Astro cells, and the average of this distribution is indicated by the color of the upper left corner box in the heatmap shown in (a).}
	\label{mouse_brain}
\end{figure}

\subsubsection{Example 3: Matching cell types to tissues}

scTOP can also be used to match cells with tissues/organs. We illustrate this by comparing cells from the Mouse Cell Atlas with a reference basis consisting of Tabula Muris organs \cite{schaum_single-cell_2018}. To do so, we averaged RNA counts across cell populations from the Mouse Cell Atlas, then preprocessed the averaged data and calculated scTOP scores. These aggregate pseudo-bulk scores tend to be higher than individual cell scores because averaging over many cells compensates for the dropout effect. As shown in figure \ref{tabula_muris}, we found the aggregate scTOP scores for cells from the Mouse Cell Atlas were correctly and strongly identified with the annotated tissue of origin in the Tabula Muris atlas. Parenchymal cell types were the most associated with the correct organ, while stromal cells were sometimes misidentified -- for example, the mammary gland endothelial cells. This suggests that stromal cells are less specified to each organ. A discussion of the accuracy of scTOP, when applied to stromal types versus parenchymal types, may be found in appendix \ref{failure_cases}.

\begin{figure}
	\centering
    \includegraphics[scale=0.5]{figs/tabula muris heatmap.pdf}
	\caption{scTOP is able to classify tissues at different resolutions. Aggregate pseudo-bulk samples from the Mouse Cell Atlas are compared to a reference basis consisting of organs from Tabula Muris. The scTOP scores are significantly higher when comparing a cell type with the organ of origin. The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis. The y-axis lists the true types of the queried cells. The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types.}
	\label{tabula_muris}
\end{figure}

\subsubsection{Example 4:  Human data}

The CellBench dataset \cite{tian_benchmarking_2019} was designed specifically to benchmark scRNA-seq analysis methods. We used the subset of data pertaining to five cancer lines. 200 cells from each line were used as training data to create the reference basis, then the rest of the cells were input as query data. As shown in figure \ref{cellbench} and table \ref{table:1}, scTOP classifies the cell line identity of the test data with 100 percent accuracy.

\begin{figure}
	\centering
		\includegraphics[scale=0.5]{figs/cellbench heatmap.pdf} 
	\caption{scTOP correctly identifies human tissues. The CellBench data set contains human cell cancer lines specifically processed to benchmark scRNA-seq algorithms. We take a subset of cells to use as the reference basis in analyzing the rest of the CellBench data. The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis. The y-axis lists the true types of the queried cells. The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types.}
	\label{cellbench}
\end{figure}

To illustrate the power of scTOP for classifying human samples, we analyzed data from the human lung atlas dataset \cite{travaglini_molecular_2020}. This dataset sequenced thousands of human lung cells and identified 58 phenotypic populations. For our analysis, we restricted the training and test data to epithelial and stromal types that had at least 200 cells in each type population. The reference basis was created using 80\% of the cells from each population, and the query data consisted of the remaining 20\% of the data. Figure \ref{human lung} shows that the scTOP scores were consistently high in cases where the score type matched the true type of the query.

\begin{figure}
	\centering
		\includegraphics[scale=0.5]{figs/human lung heatmap.pdf}
	\caption{scTOP correctly identifies human tissues. Similar to figure \ref{cellbench}, cells from the Human Lung Atlas are compared to a reference basis constructed from cells from the Human Lung Atlas. The colors of the bins correspond to the average scTOP scores of individual cells. Again, the diagonal is the most prominent section, showing that cells of the y-axis types are correctly matched to the reference x-axis identities.}
	\label{human lung}
\end{figure}

As demonstrated by the diverse data sources examined here, as long as a well-defined reference basis is available, scTOP works extremely well for cell type identification across different measurement conditions, species, organs, and resolutions. This performance is especially impressive since, as discussed earlier, the algorithm does not use any statistical fitting procedures or dimensional reduction methods such as PCA, tSNE, UMAP, or SPRING.

\subsection{Visualizing developmental dynamics using scTOP }
Having demonstrated the robust ability of scTOP to identify stable cell types, we next demonstrate how scTOP can be used in combination with scRNA-seq time series to visualize transient cell states in dynamic processes like development and differentiation.

\subsubsection{Lung development}
The alveolar epithelium of the lung undergoes dramatic maturation and morphological changes during embryonic and perinatal development, ultimately giving rise to AT1 cells and AT2 cells. To investigate this process with scTOP, we reanalyzed alveolar epithelial cells from a recent dataset by Zepp et al. \cite{zepp2021genomic}, which includes seven time points between embryonic day 12.5 (E12.5) and postnatal day 42 (P42). These cells were then compared against a reference basis developed from adult cell types in the Mouse Cell Atlas and an E12.5 epithelial progenitor cell type from the Atlas of Lung Development  \cite{negretti_single-cell_2021}.

The resulting AT1 and AT2 scTOP scores were plotted to visualize maturation into these two lineages (Figure \ref{LungMAP}(a)). As expected, early embryonic progenitor cells (E12.5, E15.5), which express few of the mature lineage markers, align poorly with both lineages. In contrast, the E17.5 alveolar epithelial cells fall along a wide spectrum with cells aligning to AT1 cells, AT2 cells, or both lineages. In the postnatal time points, this continuous spectrum gives way to more distinct clusters, eventually resolving into AT2 and AT1 cell clusters at P42. Similar to our analysis of adult mouse lungs (Figure \ref{herriges}), the scTOP scores seen over development correlate well with our lineage marker gene sets (Figure \ref{LungMAP} (b)-(c)).

In addition to the expected postnatal AT2 and AT1 cell clusters, this scTOP analysis illustrated a transient AT2/AT1 hybrid population that persisted through P7. The existence of these hybrid AT2/AT1 cells was already noticed in Zepp et al. using a combination of graph-based clustering and gene enrichment profiling (see Figure 1C, I in \cite{zepp2021genomic}). The authors conjectured that these AT2/AT1 cells were a transitional state of AT2 similar to Spock2+/Axin2+ AT2 cells previously identified in adult lungs \cite{frank2016emergence}. To assess whether this was the case, we displayed Axin2 expression on our scTOP plots (Figure \ref{LungMAP genes}(a)). These cells do not have significant upregulation of Axin2 relative to perinatal or adult AT2 cells, suggesting that the postnatal hybrid AT2/AT1 cells are distinct from adult Spock2+/Axin2+ AT2 cells \cite{frank2016emergence}.

We then performed a similar analysis of markers for other published postnatal hybrid AT1/AT2 or transitional alveolar cell states, including alveolar differentiation intermediate (ADI, {\it Krt8} and {\it Trp53}), pre-alveolar type-1 transitional cell state (PATS, {\it Cldn4} and {\it Krt19}), and damage-associated transient progenitors (DATPs, {\it Ilr1} and {\it Hif1a}) \cite{verheyden2020transitional, strunz2020alveolar, kobayashi2020persistence, choi2020inflammatory}. In each case, markers were not uniquely upregulated in the perinatal hybrid AT1/AT2 cells relative to either the perinatal or adult committed lineages (Figure \ref{LungMAP genes} (b) - (d)). Furthermore, differential gene expression analysis failed to identify any genes that were uniquely differentially expressed in the hybrid AT1/AT2 cells relative to either timepoint-matched AT1 or AT2 cells. These results suggest that this population is most likely transcriptionally distinct from previously identified adult AT2-transitional cell types.

\begin{figure}
	\centering
		\includegraphics[scale=0.52]{figs/LungMAP.png}
	\caption{Visualizing the developing mouse lung with scTOP. scTOP illustrates the specification of alveolar type I and type II cells in murine embryonic development. Alveolar cells from Zepp et al. are compared to a reference basis constructed from adult lung cells from Herriges et al. and early epithelial cells from the Atlas of Lung Development. (a) Each subplot corresponds to the age of the mouse from which the sample was extracted, from post-conception day 12,5 (E12.5) to 42 days post-birth (P42). (b), (c) Scatter plots of the same data as in (a), now with each cell colored by the average over that cell's expression of (b) AT1 marker genes and (c) AT2 marker genes. The scTOP scores of AT1 (AT2) cells are high when the average AT1 (AT2) marker gene expression is high. (d) The same data as above is displayed as a 2D histogram, with each bin colored corresponding to the average day of the data point(s) falling within that bin. The color trend shows that early cells generally have very low AT1 and AT2 scores, then the AT1 and AT2 scores increase over time.}
	\label{LungMAP}
\end{figure}

\begin{figure}
	\centering
		\includegraphics[scale=0.8]{figs/LungMAP genes.png}
	\caption{The AT1/AT2 combination state (most apparent in days P3 and P7) do not uniquely express genes of AT1/AT2 transitional states described by previous papers. Alveolar cells from Zepp et al. are plotted on AT1 and AT2 axes, separated by day. Each cell is colored by the expression z-score of a particular gene, and the particular gene is different in each row. The gene used to color the cells is indicated by the color bar on the far right of each row. (a) Alveolar cells colored by Axin2, showing that the AT1/AT2 combination cells do not appear to belong to the Axin2+ sublineage described by Frank et al. \cite{frank2016emergence}. (b) Alveolar cells colored by Krt8 and Trp53 expression, which are genes corresponding to alveolar differentiation intermediate cells (ADI) described by Strunz et al. \cite{strunz2020alveolar}. AT1, AT2, and the combination AT1/AT2 state all express Krt8 and Trp53 at similar levels. (c) Alveolar cells colored by expression of pre-alveolar type-1 transitional cell state (PATS) genes Kobayashi et al. \cite{kobayashi2020persistence}. AT1, AT2, and the combination AT1/AT2 state all express Cldn4 at similar levels. AT1 and AT1/AT2 combination cells both express Krt19 at similar levels, although AT1 cells express the gene at higher levels. (d) Alveolar cells colored by expression of damage-associated transient progenitor (DATP) genes described by Choi et al. \cite{choi2020inflammatory}. AT1, AT2, and the combination AT1/AT2 state all express Il1r1 and HIF1a at similar levels. }
	\label{LungMAP genes}
\end{figure}

\subsubsection{scTOP for hematopoiesis using lineage tracing data}

In the previous section, we used scTOP to visualize differentiation into two closely related lung epithelial cell fates. In this section, we show how scTOP can also be used to gain insights into more complicated developmental processes such as hematopoiesis, in which hematopoietic stem cells (HSCs) give rise to multiple differentiated blood cell types. To do so, we make use of the unique dataset by Weinreb et al. \cite{weinreb_lineage_2020} on hematopoiesis in mice using a new technique, lineage and RNA recovery (LARRY), that combines scRNA-seq and lineage tracing data. LARRY uses heritable barcodes that can be detected using scRNA sequencing, allowing for the tracking of clone families during differentiation. Here, we focus on visualizing the ex-vivo experiments in \cite{weinreb_lineage_2020}, where cells were sequenced at a sufficient depth to infer lineage relationships. In these experiments, HSCs were extracted from mice, barcoded, and allowed to expand in primary culture. scRNA-seq measurements were performed on days 2, 4, and 6 post-barcoding. The cells were then annotated using marker genes into 9 differentiated blood cell fates.

We restrict our analysis to a subset of these annotated cell types with at least 150 cells. We also excluded cells labeled as erythroids from our analysis since we found that this population was highly heterogeneous and gave rise to poor projections (likely due to the fact it was annotated using a single marker gene, \textit{Hbb-bs} \cite{weinreb_lineage_2020}, which did not sufficiently define a distinct population). Furthermore, we treated all day 2 undifferentiated cells as a single population we called progenitors. This resulted in a reference basis consisting of seven cell types: undifferentiated progenitors, neutrophils, monocytes, megakaryocytes, mast cells, eosinophils, and basophils. To construct the basis, we averaged the RNA counts for 150 random cells of each of these types, only including cells if they had no sister clones. Then we preprocessed the averaged gene expression profiles and inserted them into the reference basis.

Figure \ref{hem clones} shows the differentiation dynamics of three different clonal families using scTOP. The histograms show the distribution of scTOP projection scores on each cell type for days 2, 4, and 6 post-barcoding, while the scatter plots show two-dimensional cross-sections of scTOP projection scores as a function of time for the indicated cell types. In all the plots, the scTOP projections for the undifferentiated hematopoietic progenitor stem cell decreased over time, with a high projection on day 2 and a low projection on day 6, while the scores for the mature types increased in time.

Interestingly, the three clones show very different behaviors as they mature. Figure \ref{hem clones} (a) shows a clone where a single barcode detected at day 2 gives rise to multiple cells in the neutrophil lineage. This can be seen in the plots by noting that the day 6 cells have near zero scores on both progenitor cells and other differentiated lineages, such as monocytes, but project significantly on the neutrophils. Figure \ref{hem clones} (b) shows a barcode that was detected in three cells on day 2 that gives rise to a mix of neutrophils and monocytes on day 6. Interestingly, the day 4 cells move in the direction of both of these lineages before bifurcating at day 6. This data is consistent with Weinreb et al. \cite{weinreb_lineage_2020}, who previously highlighted the existence of such a bipotent neutrophil-monocyte progenitor population. We also find evidence that progenitor cells can even give rise to three distinct cellular populations. Figure \ref{hem clones} (c) shows a single barcode found on day 2 that results in day 6 cells with three distinct populations with non-zero scores on basophils, eosinophils, and neutrophils, suggesting a single clone can bifurcate into three different types of differentiated cell types.

Figure \ref{hem clones} focuses on scTOP scores for individual clonal families, while Figure \ref{hem progenitor} shows the scTOP projection scores across all 1,816 multi-generational clonal families in the dataset. In each subplot, the horizontal axes show progenitor scores, and the vertical axes correspond to the six differentiated cell types included in the reference basis. Each bin of the 2-dimensional histogram is colored according to the {\it average} day of all the cells that fall within that bin. For this reason, although the cells were only sampled on days 2, 4, and 6, the color bar is continuous, with a range between 2 and 6 days. The histograms show that, as expected, the day 2 cells generally have high projections on progenitor cells, but as time passes, progenitor scores decrease while the scores on differentiated cell types either increase (indicating cells that differentiate into the cell type on the y-axis) or remain close to zero (indicating cells that differentiate into a different cell type than that on the y-axis). Somewhat surprisingly, the plots for different cell types look qualitatively different from each other (for example, compare the plots for megakaryocytes and monocytes).

To try to better understand this, we made additional 2-dimensional histograms comparing scTOP scores between all differentiated cell types in our basis (see Figure \ref{hem mature}). What is striking about these plots is that they show two qualitatively different behaviors depending on which pairs of cells are being compared: the differentiation pathways for some pairs are mutually exclusive, while other pairs show a more graded behavior. For example, the histogram for megakaryocytes and monocytes exhibits a distinct L-shape, indicating that these developmental programs are mutually exclusive since cells never have significant projection scores of both cell types simultaneously. This is in stark contrast with the histogram for monocytes and neutrophils, where there exists a continuous spectrum of cells that are monocyte-like, neutrophil-like, and every step in between.

Our scTOP-based visualizations of developmental decisions during hematopoiesis complement the analysis in \cite{weinreb_lineage_2020} using pseudo-time. Like in the original paper, we find strong evidence for the existence of progenitor cells that can give rise to multiple lineages (Figure \ref{hem clones}), including a neutrophil/monocyte bivalent progenitor and possibly a basophil/eosinophil/neutrophils trivalent progenitor.  The existence of these multi-potent progenitors seems to give rise to graded developmental dynamics where intermediate cells have significant projections on multiple cell types. This is in stark contrast with developmental decisions between other cell lineages (megakaryocytes vs. monocytes, basophils vs. megakaryocytes, monocytes vs. basophils) that seem to be mutually exclusive. One compelling feature of these visualizations using scTOP is that they require no statistical fitting, dimensional reduction, or ordering of cells. Once we have chosen the relevant reference basis (in this case, progenitors and six differentiated lineages), the resulting plots give an interpretable way of assessing complex developmental dynamics.

\begin{figure}
	\centering
		\includegraphics[scale=0.75]{figs/hem. clone families.png}
	\caption{Tracing hematopoietic cell differentiation with scTOP. (a), (b), (c) display scatter plots and distributions of scTOP scores for individual in vitro clone families from  \cite{weinreb_lineage_2020}. The scatter plots show the x-axis scTOP score compared to the y-axis scTOP score for individual cells, while the distributions show the scTOP scores of the type indicated on the x-axis of the corresponding column. The color and shape of each marker (and the color of each distribution) indicate the days in culture on which the cells were extracted. (a) A clone family where all of the cells end up becoming neutrophils. (b) A clone family where some clones end up becoming neutrophils and some become monocytes. At day 2, the 3 sister cells have low neutrophil and monocyte scores and high progenitor scores. At days 4 and 6, the progenitor scores decrease, and a portion of the cells increase in monocyte scores while others increase in neutrophil scores. (c) A clone family where the descendant cells become neutrophils, eosinophils, and basophils.}
	\label{hem clones}
\end{figure}

\begin{figure}
	\centering
		\includegraphics[scale=0.75]{figs/hem progenitor hists.png}
	\caption{Tracing differentiation of hematopoietic cells from progenitor to mature type. The plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al., with the x-axis of each one indicating the scTOP score for hematopoietic progenitor type and the y-axis showing the score for a mature type. Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin. Early cells have high progenitor scores and low mature-type scores. As time passes, cells end up with high scores in mature types.}
	\label{hem progenitor}
\end{figure}

\begin{figure}
	\centering
		\includegraphics[scale=0.59]{figs/hem mature hists.png}
	\caption{Comparing differentiation of hematopoietic cells between mature types. As in figure \ref{hem progenitor}, the plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al. The x-axis and y-axis of each histogram indicate the scTOP score of the labeled mature type. Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin.  Some pairs of types have cells with high scores of both types. This is apparent in the neutrophil-monocyte histogram, which presents a continuous spectrum of scores from the horizontal axis to the vertical. Other pairs of types only have cells expressing one type or the other, as in the L-shaped neutrophil-megakaryocyte histogram.}
	\label{hem mature}
\end{figure}

\subsection{Using scTOP to compare endogenous and transplanted cells}

\begin{figure}[h!]
	\centering
		\includegraphics[scale=0.83]{figs/engineered cells comparison.png}[h!]
	\caption{Donor-derived and endogenous alveolar cells from Herriges et al. and Louie et al. The cell type annotations come from a separate Louvain clustering and are the same for all plots in this figure. (a) UMAP visualizations of the data, colored by cell type annotations. The three have different values for the UMAP parameter n\_neighbors, which changes whether the embedding preserves more local or global structure. (b) scTOP plot showing the AT1 score on the x-axis and the AT2 score on the y-axis. (c) Box-and-whisker plot showing the distributions of scTOP AT2 scores for the various cell populations. (d) Bar plot showing the top ten aggregate scores for the AT2 populations from different sources, sorted by the scores for the SCA1+ population.}
	\label{engineered}
\end{figure}
Recent advancements in cell culture and cell transplantation techniques have allowed researchers to generate cells that were maintained or differentiated in vitro to approximate and eventually replace endogenous cells generated during in vivo development. However, even with scRNA-seq, it is difficult to consistently quantify the transcriptomic similarity between donor-derived and endogenous lineages from transplant recipients. Since it provides a quantitative measure of cell type similarity, scTOP can be useful for comparing cell populations. A relevant example can be found in pulmonary cell replacement therapy, where one goal is to differentiate or maintain alveolar cells in vitro and then transplant them into injured lungs.

To compare two distinct pulmonary cell transplant protocols, we used scTOP to analyze the results of scRNA-seq from two recent publications featuring murine pulmonary cell transplantation \cite{herriges_durable_2022, louie_progenitor_2022}. In Louie et al., adult mouse AT2 cells were maintained in culture with support cells for 3 weeks to generate a donor cell population. In contrast, in Herriges et al. the authors developed a protocol for directed differentiation of mouse PSCs into lung distal tip-like cells, which mimic an embryonic progenitor of AT2 cells. In both papers, the donor cells were transplanted into injured mouse lungs, where they survived and gave rise to multiple cell lineages, including AT2-like cells.

To highlight the usefulness of scTOP, we first consider another dimensional reduction method. Dimensional reduction methods like UMAP and t-SNE are often used to compare similar cells from different sources, such as transplanted and endogenous cells. Figure \ref{engineered} shows the AT1 and AT2 cell populations from Herriges et al. and Louie et al., both donor-derived and endogenous. In figure \ref{engineered} (a), three possible UMAP plots are shown with varying algorithmic parameters. The qualitative relations between the clusters are generally consistent, with AT1 cells clustering together apart from the AT2 cells. However, the distances between cells and the relative positions of clusters are not consistent when the UMAP parameter is varied. This illustrates how UMAP is useful for visualization and gaining qualitative intuition for a particular dataset but cannot be used to make a rigorous quantitative comparison between cells.

On the other hand, figure \ref{engineered} (b) shows the scTOP scores for AT1 and AT2 for the same populations. The distances between cells will only change if the reference basis is changed, and even then will not change significantly (see \ref{robustness hists} for discussion). The AT1 and AT2 populations are separated along the relevant axes, and the highly-similar cell populations are largely overlapping. Because scTOP provides many relevant dimensions to choose from instead of trying to capture high-dimensional distances in two dimensions, it is possible to directly compare different populations to the same cell type. Figure \ref{engineered} (c) shows the distributions of AT2 scores for each of the populations, providing a clear quantitative way to compare the various engineered cells. As expected, the endogenous populations from both data sources are similar in distribution. The means and overall distributions of the AT2-derived (Louie et al.) cells are higher than those for the PSC-derived (Herriges et al.) cells, suggesting that the primary cell transplantation better emulates the endogenous AT2 cells.

Figure \ref{engineered} (d) provides a comparison between cell populations using aggregate scTOP scores. Similar to figure \ref{engineered} (c), (d) provides a quantitative comparison between the populations. This figure confirms that while both AT2-derived and PSC-derived transplants generate cells that have high AT2 scores, the AT2-derived cells are quantitatively more similar to the reference AT2 profile. Altogether this analysis demonstrates how scTOP can be used to compare two distinct cell populations against a primary control. scTOP satisfies the need for quantitative assessment of transplantation protocols and can be used in combination with functional tests to identify protocols that most effectively replace endogenous cell lineages.

\section{Discussion}

In this paper, we have introduced a new physics-inspired method, scTOP, for analyzing scRNA-seq data. scTOP projection scores provide clear and meaningful axes to visualize differentiation and classify cells. Since scTOP requires no statistical fitting, clustering, or dimensional reduction techniques, plots can be made quickly and easily, even for an individual cell. The input to scTOP is a reference basis of cell fates of interest, allowing us to take advantage of the wealth of new scRNA-seq expression atlases now being generated. Importantly, scTOP scores are robust to the choice of basis and reproducible across experiments, labs, and datasets, allowing integration of disparate data sources.

scTOP is especially suited for tasks where the cell types of interests are well characterized. For example, we have found in our own work that scTOP is an extremely sensitive measure for assessing the fidelity of engineered cells from directed differentiation protocols and reprogramming \cite{herriges_durable_2022}. Since scTOP does not require data harmonization, joint embeddings, or make use of cellular neighborhood information,  scTOP is able to distinguish between technical noise and biologically meaningful differences. scTOP also can be used to classify cells with minimal computational costs and thus represents a potentially simple, scalable, and reproducible way of dealing with the proliferation of scRNA-seq datasets. Since scTOP yields an alternative coordinate system for representing developmental dynamics in cell-fate space, scTOP-based visualizations are a powerful way of thinking about complex developmental processes.

As an illustration, we analyzed the development of the lung epithelial AT1 and AT2 cell fates \cite{zepp2021genomic} and developmental dynamics during hematopoiesis  \cite{weinreb_lineage_2020}. Using scTOP, we showed that it was easy to identify a transient hybrid AT2/AT1 population 3 to 5 days post-birth whose gene expression profile is distinct from previously investigated transient AT2-to-AT1 states. By plotting scTOP scores as a function of time for AT1 and AT2, hybrid AT2/AT1 cells were easily identified since they formed a distinct cluster that was well separated from mature AT1 and AT2 cells. This example illustrates how given an accurate reference basis, scTOP is able to pick up on small but biologically meaningful differences in global gene expression profiles.

In the context of hematopoietic development, where HSCs give rise to multiple downstream cell fates, we took advantage of the fact that scTOP maps each cell to a seven-dimensional coordinate (each direction in the coordinate systems corresponds to one of the seven cell types in the reference basis: undifferentiated progenitors, neutrophils, monocytes, megakaryocytes, mast cells, eosinophils, and basophils) to generate multiple two-dimensional visualizations using scTOP. We found that a single clonal family can often give rise to up to three different mature cell fates. Our visualizations also suggest that developmental decisions between different cell fates can be classified into two broad categories: mutually exclusive and graded. For cell fates such as megakaryocytes and monocytes, cells never have significant projection scores on both lineages simultaneously, suggesting that the underlying developmental decisions are mutually exclusive. In contrast, for other pairs of cell fates, such as monocytes and neutrophils, cells often exhibit significant projections on both members of the pair suggesting these developmental switches function in a more graded manner. It will be interesting to see if this general distinction is also present in other developmental systems or is specific to hematopoiesis.

In the analysis here, we have limited ourselves to considering scRNA-seq data. However, in principle, our approach can be easily extended to include other data modalities, such as chromatic accessibility and histone modifications. Recently, several interesting works such as Dynamo have used RNA-velocity to directly learn cellular dynamics \cite{xing2022reconstructing, qiu2022mapping}. It will be interesting to see if these methods can be combined with scTOP to learn vector flows directly in cell-type space. Similarly, methods based on optimal transport, such as Waddington-OT \cite{schiebinger2019optimal}, or developmental directory reconstruction based on graph embedding, such as Monocle \cite{qiu2017reversed}, can also be performed in the cellular identity space instead of gene expression space. These represent promising directions for future research since the scTOP cell identity coordinates are often better suited for analyzing developmental dynamics than gene expression coordinates.

We have found that scTOP can be used to quantitatively assess the fidelity of cultured cells, as in the case of donor-derived lung alveolar cells. However, we currently have no systematic manner within the scTOP framework for providing guidance on adjusting gene expression or designing differentiation protocols to steer cells toward desired fates. Developing such methods would provide powerful new computational tools for engineering cells via directed differentiation.

We are also interested in extending scTOP to better understand the signaling pathways and genetic signatures underlying developmental dynamics. While scTOP provides a way to translate between gene expression space and cell type space, there is currently no direct interface with the biologically-essential signaling space. Further work in uniting the three relevant spaces, combined with studying the possible bifurcations in these spaces, may provide insight into the complex process of differentiation.

\section*{Acknowledgments}
We acknowledge useful discussions with Jason Rocks, Robert Marsland, and Alex Lang. We would like to thank the Mehta group and Kotton lab for their comments on the manuscript. The work was funded by a grant to the Boston University Kilachand Multicellular Design Program (to DK and PM) and NIH NIGMS 1R35GM119461 to PM. We also acknowledge programming support from Alena Yampolskaya and Sumner Hearth.

\section*{Appendix}
\beginsupplement
\subsection{Preprocessing} \label{preprocessing}
The raw scRNA-seq count data for each dataset was downloaded as a matrix of mRNA counts corresponding to genes and cells. An example of such a matrix is shown in step 0 of figure \ref{FIG:supp1}, where each row is a different gene, and each column is a different cell. Throughout this figure, the histogram to the right of each step shows the distribution of entry values for the first column (cell) in the matrix, and the histogram below the step shows the distribution of entry values for the second row (gene Rpl7) of the matrix. The histograms are presented as examples of how the distributions (for each cell and gene) of gene expressions change as the data is preprocessed.

In the first step of preprocessing, each cell is normalized independently. This step is taken because scRNA-seq generally measures relative rather than absolute expression levels within a cell. In order to facilitate the comparison of cells across sequencing conditions, we transform the normalized RNA counts into quantities that are centered around the mean expression level. Z-scores are exactly such a quantity; they measure the number of standard deviations away from the mean. To calculate the z-scores after step 1, we convert the normalized gene expressions into percentiles by ranking each gene in a given cell from lowest to highest, then dividing by the number of genes (plus one so that the highest rank is not assigned the 100th percentile). For example, if a cell has 1000 genes sampled and Rpl7 is the highest-expressing gene, it will be given rank 1000 and labeled as the 99th percentile. If two or more genes are tied for rank, as is the case for zero-values, they are assigned the average of the ranks they would have been assigned if they were not tied (see rankdata function from the Python package Scipy). To convert from percentiles to z-scores, we assumed a normal distribution with mean 0 and standard deviation 1 and applied the corresponding quantile function.

\begin{figure}
	\centering
		\includegraphics[scale=0.6]{figs/fig1a supplement.pdf}
	\caption{Data distributions at each preprocessing step. Step 0 shows the original RNA count matrix before preprocessing, where each entry corresponds to the number of RNA detected for a particular gene in a particular cell. The rows of each matrix correspond to genes, and the columns correspond to cells. The histograms to the right of each matrix show the distribution of an example cell, while the histograms below each matrix show the distribution of an example gene. Step 1 of preprocessing is to normalize each cell. Steps 2 and 3 convert the normalized gene expressions to z-scores. Step 2 ranks the genes in each cell according to magnitude and converts the rank into a percentile. Step 3 uses a quantile function, assuming a normal distribution with mean zero and standard deviation one, to convert from percentiles to z-scores.}
	\label{FIG:supp1}
\end{figure}

\subsection{Reference basis construction}\label{basis construction}
Creating an appropriate reference basis is vital to the accuracy of scTOP. There are many existing scRNA-seq atlases, such as the Mouse Cell Atlas; using them with scTOP requires some amount of curation. For each reference dataset, we took an average over each population corresponding to a given cell type. The average gene expression profiles were preprocessed and then used as the reference profiles for each cell type. Often, it was necessary to perform additional curation of the reference basis. Cell types were dropped from an atlas if they did not include enough cells (i.e., if they contained $<100$ cells) or were found to be poorly defined (i.e., not enough marker genes to distinguish that cell type from others). Common instances of the latter were cell types that differed only in the expression of a single gene.

Ensuring that each reference gene expression profile sampled enough cells was essential. This is because the number of cells determines how well-sampled the reference cell type is. Although cell types have distinct transcriptomic patterns, they are not all transcriptomically identical. There are variations in gene expression levels between cells of the same type due to differences in cell-level processes such as cell division or environmental stimuli. Averaging over a population of cells which are some cell type X creates an approximate gene expression profile of the archetypal X cell.

\subsubsection{Mouse Cell Atlas}
The Mouse Cell Atlas (MCA) contains hundreds of thousands of cells from tissues across the mouse body at various stages of development. For the reference basis created from the Mouse Cell Atlas (MCA), we dropped cell types that had fewer than 100 cells. For cell types that had greater than or equal to 100 cells, we averaged across the entire population for each one, then preprocessed. For cell types that were divided according to high expressions of various genes, we combined them into one cell type. For example, the original MCA contained dozens of clusters of mammary gland secretory alveoli cells such as ``secretory alveoli cell, Hes1 high" or ``secretory alveoli cell, Gpx3 high." These were all combined into one cell type, ``mammary gland secretory alveoli." This is because we wanted each reference basis type to act as an archetype rather than a perfect representation of every possible iteration of a cell type; each reference type represents a basin of attraction, and the basins were defined by the overall cell type, not whether individual genes were particularly high within that cell type. We cleaned each of the MCA type labels and ended up with cell types defined by the organ of origin, specific cell type, and time point of collection (e.g., ``lung ciliated cell week 6-10"). The time points of cell types in the MCA were either E14.5 or somewhere between week 6 and week 10 post-birth.   The resulting reference basis contained 221 cell types.

\subsection{Accuracy measures}
For table \ref{table:1}, the top1 accuracy was determined by applying scTOP to a cell, finding the type with the highest score, and checking whether that type matched the cell's true type. Top3 was similarly determined, except instead of just checking the highest-scoring type, we checked whether the true type was included in the top 3 highest-scoring types.

By looking at the scTOP scores for cells whose true type did not match the reference type, we set a score of 0.1 as the lower bound for identifying a type. As shown in figure \ref{robustness hists} (c) and (f), when the reference type does not match the sample type, the scores for individual cells fall well within [-0.1, 0.1], even when most of the reference basis is missing.

In table \ref{table:1}, we labeled cells as unspecified if the highest scTOP score was less than 0.1. This provides a useful metric for determining whether a query cell belongs to one of the cell types in the reference basis. If the highest scTOP score is lower than 0.1 for a query cell, this indicates that the reference basis lacks the relevant types or was poorly constructed (see appendix \ref{failure_cases} for a discussion of ill-defined bases).

scTOP can only identify cells that exist within the reference basis. As such, the accuracy scores were calculated only for the cells whose true type (as determined by the annotations from the authors) was included in the reference basis for the corresponding analysis. For cells whose true types were not represented in the basis, the unspecified rate was very high. This indicates that the rate of false positives is very low.

\subsection{Robustness of scTOP}
scTOP is consistent between iterations of the algorithm, and it is robust to changes in the reference basis. As shown in figure \ref{robustness hists}, many scTOP scores do not change significantly even when most of the reference basis is removed. Generally, scores for cells that match the reference type do not change much even when only 25\% of the original basis is retained (figure \ref{robustness hists} (d), (e)). The largest difference occurs for cells that are similar to the reference type, such as the AT1 score for AT2 cells (figure \ref{robustness hists} (g)). These scores tend to increase because removing other types from the basis reduces the ability to de-correlate effectively. In other words, scTOP is more likely to confuse similar types when it has fewer reference types to compare. This effect is most apparent when relevant types are removed from the basis. Figures \ref{robustness hists} (j) and (k) show that all lung types score higher for AT1 and AT2 when only lung alveolar types are included in the basis. Figures \ref{robustness hists} (c) and (f) show that irrelevant scores, such as bladder and thymus cells, do not change much even when large portions of the basis are removed. scTOP does not confuse sample cells for types that are dissimilar even with a small reference basis because there is no need to de-correlate between types that are naturally not correlated.

\subsubsection{Cases where it does not work}\label{failure_cases}
The performance of scTOP relies on the quality of the reference basis. scTOP is expected to give inaccurate results in cases where the reference basis is ill-defined. The underlying assumption of the algorithm is that the reference basis consists of accurate gene expression profiles of cell types that are attractor basins. In other words, the reference types are assumed to represent stable cell types rather than cell states that are on the path to the true cell type basins. For example, scTOP performs poorly when using a reference basis consisting of early embryonic identities because the cell types are not sufficiently differentiated from one another. Adapting scTOP to such early lineage data is an important and interesting problem.

The other part of the assumption is that the reference types should be accurate profiles of the cell type of interest. As discussed in appendix \ref{basis construction}, reference types are created from averaging over cells, and the more cells included in this average, the better-sampled the type is. As expected, scTOP gives unreliable results when the reference gene expression levels do not accurately reflect the true cell types' gene expression levels. An important factor in the accuracy of scRNA-seq is the amplification step: after RNA is extracted, it is amplified to improve the sensitivity of sequencing, but not all RNA is amplified at the same rate. This results in inflated counts for some genes but not others. Unique molecular identifiers (UMIs) are vital for reducing the inflating effect of amplification \cite{islam2014quantitative}. UMIs tag the RNA before amplification and are present post-amplification; this makes it possible to see which amplified RNA came from the same original molecule and prevents uneven double-counting of RNA. We have found that scTOP gives more accurate results when used with reference and query datasets that use UMIs, compared to scTOP results using data that did not use UMIs.

\begin{figure}
	\centering
		\includegraphics[scale=0.72]{figs/hem genes.png}
	\caption{Scatter plots of in vitro hematopoietic clone families from Weinreb et al., with scTOP scores for progenitor, neutrophil, and monocyte types, and colored by normalized gene expression. In Weinreb et al. figure 4 (a), cells visualized using pseudo-time inference were separated into branches. The monocyte branch appeared to have half of its cells expressing dendritic cell genes and the other half expressing neutrophil genes. This figure plots those same cells on progenitor, neutrophil, and monocyte axes. The colors correspond to neutrophil, dendritic cell, and monocyte marker genes. (a) Hematopoietic cells colored according to the expression of neutrophil marker genes S100a8 and Elane. (b) Hematopoietic cells colored according to the expression of dendritic cell marker genes Cd74 and H2-Aa. (c) Hematopoietic cells colored according to the expression of monocyte marker genes Csf1r and Klf4. }
	\label{hem genes}
\end{figure}

\begin{figure}
	\centering
		\includegraphics[scale=0.82]{figs/robustness hists.png}
	\caption{Distributions of scTOP scores for lung cells from Herriges et al., using varying portions of the Mouse Cell Atlas as the basis. The label above each plot indicates the Herriges et al. sample subset, while the x-axis label indicates the scTOP reference type. "All lung types" means that all of the lung types from the data set were included. (a) - (i) show how scTOP scores change when 25-100\% of the basis is used. (j) - (k) show how scTOP scores change when the basis is changed from the full basis (which includes many tissues and organs) to just the lung types and to just the two alveolar types. (c) and (f) show scTOP scores for random irrelevant types, Bladder Epithelium and Thymus B Cell, to show that the scTOP scores remain low for types that do not match the sample.}
	\label{robustness hists}
\end{figure}

\subsection{AT1/AT2 marker genes}\label{alveolar markers}
In figure \ref{LungMAP}, plots in (b) and (c) were colored according to AT1 and AT2 marker gene expression. For each cell, the average was taken over the expressions of a set of marker genes. The AT1 marker gene set was:

\begin{itemize}
    \item Ager
    \item Akap5
    \item Aqp5
    \item Cldn18
    \item Clic3
    \item Clic5
    \item Col4a3
    \item Col4a4
    \item Cryab
    \item Cyp2b10
    \item Emp2
    \item Fam189a2
    \item Gprc5a
    \item Hopx
    \item Hs2st1
    \item Igfbp2
    \item Krt7
    \item Lmo7
    \item Mal2
    \item Pdpn
    \item Prdx6
    \item Pxdc1
    \item Rtkn2
    \item Scnn1g
    \item Spock2
    \item Tspan8
    \item Vegfa
\end{itemize}

The AT2 marker gene set was:

\begin{itemize}
    \item Abca3
    \item Acot7
    \item Acoxl
    \item Acsl4
    \item Ank3
    \item Atp8a1
    \item Cd74
    \item Cebpa
    \item Cpm
    \item Ctsh
    \item Cxcl15
    \item Dram1
    \item Egfl6
    \item Elovl1
    \item Etv5
    \item Fabp5
    \item Fasn
    \item Hc
    \item Lamp3
    \item Lgi3
    \item Lpcat1
    \item Lyz2
    \item Mlc1
    \item Muc1
    \item Napsa
    \item Npc2
    \item Ppp1r14c
    \item S100g
    \item Scd1
    \item Sfta2
    \item Sftpa1
    \item Sftpb
    \item Sftpc
    \item Sftpd
    \item Slc34a2
    \item Zdhhc3
\end{itemize}

\bibliography{export-data.bib}

\end{document}

### 2023-01-25 scripts/brain_atlas_accuracies.py

@@ -1,0 +1,149 @@
+import numpy as np
+import pandas as pd
+
+import sys
+# Add path of module (temporary measure until pip install works)
+sys.path.append('/project/biophys/mouse scRNA-seq/scTOP explorations/scTOP/sctop')
+import sctop as top
+
+import h5py
+
+fname = '/projectnb/biophys/mariay/scTOP_manuscript_data/Brain Atlas/Analysis_Zeng_Hippocampus_10X/data/10x_v2/mouse/processed/YaoHippo2020/CTX_Hip_counts_10x.h5'
+h5f = h5py.File(fname,'r')
+genes = np.array(h5f['data/gene'].asstr())
+annotations = pd.read_csv('/projectnb/biophys/mariay/scTOP_manuscript_data/Brain Atlas/Analysis_Zeng_Hippocampus_10X/data/10x_v2/mouse/processed/YaoHippo2020/CTX_Hip_anno_10x/CTX_Hip_anno_10x.csv')
+annotations = annotations.set_index('sample_name')
+
+h5_names = np.array(h5f['data/samples']).astype(str)
+
+specifier = 'subclass_label'
+class_basis = pd.read_csv('/project/biophys/mouse scRNA-seq/scTOP explorations/scTOP/manuscript/manuscript data/brain_atlas.csv')
+training_indices = pd.read_csv('/project/biophys/mouse scRNA-seq/scTOP explorations/scTOP/manuscript/manuscript data/brain_atlas_training_indices.csv')
+
+class_basis = class_basis.set_index('Unnamed: 0')
+training_indices = training_indices.loc[:,'0'].values
+
+training_indices_converted = []
+for index in training_indices:
+    training_indices_converted += [np.where(h5_names == annotations.index[int(index)])[0][0]]
+    
+training_sample_names = h5_names[training_indices_converted]
+
+# Find accuracy scores one batch at a time, so we don't have to load 1.2 million samples at once
+batch_size = 50000
+sample_index = 0
+
+excluded_types = np.setdiff1d(np.unique(annotations.loc[:, specifier]), class_basis.columns.values)
+number_excluded_cells = np.sum([len(np.where(annotations.loc[:, specifier] == cell_type)[0]) for cell_type in excluded_types])
+total_cells = len(h5_names) - len(training_sample_names) # total cells in test data (total - training)
+
+included_types_accuracies = {'top1': 0,
+                             'top3': 0,
+                             'unspecified': 0,
[…]

### 2023-01-25 scripts/brain_atlas_bubbleplot.py

@@ -1,0 +1,52 @@
+import numpy as np
+import pandas as pd
+
+import sys
+# Add path of module (temporary measure until pip install works)
+sys.path.append('/project/biophys/mouse scRNA-seq/scTOP explorations/scTOP/sctop')
+import sctop as top
+
+import h5py
+
+fname = '/projectnb/biophys/mariay/scTOP_manuscript_data/Brain Atlas/Analysis_Zeng_Hippocampus_10X/data/10x_v2/mouse/processed/YaoHippo2020/CTX_Hip_counts_10x.h5'
+h5f = h5py.File(fname,'r')
+genes = np.array(h5f['data/gene'].asstr())
+annotations = pd.read_csv('/projectnb/biophys/mariay/scTOP_manuscript_data/Brain Atlas/Analysis_Zeng_Hippocampus_10X/data/10x_v2/mouse/processed/YaoHippo2020/CTX_Hip_anno_10x/CTX_Hip_anno_10x.csv')
+annotations = annotations.set_index('sample_name')
+
+h5_names = np.array(h5f['data/samples']).astype(str)
+
+specifier = 'subclass_label'
+class_basis = pd.read_csv('/project/biophys/mouse scRNA-seq/scTOP explorations/scTOP/manuscript/manuscript data/brain_atlas.csv')
+training_indices = pd.read_csv('/project/biophys/mouse scRNA-seq/scTOP explorations/scTOP/manuscript/manuscript data/brain_atlas_training_indices.csv')
+
+class_basis = class_basis.set_index('Unnamed: 0')
+training_indices = training_indices.loc[:,'0'].values
+
+training_indices_converted = []
+for index in training_indices:
+    training_indices_converted += [np.where(h5_names == annotations.index[int(index)])[0][0]]
+    
+# Choose ten populations to make heatmaps
+populations = ['Astro', 'Oligo', 'CT SUB', 'CA3', 'Sst', 'L2/3 IT CTX', 'L5 PT CTX', 'Pvalb', 'Vip', 'Lamp5']
+population_scores = pd.DataFrame(index=populations, columns=populations)
+
+for population in populations:
+    # Random sample of indices for population type, excluding indices used for training data
+    current_condition = annotations.loc[:, specifier] == population
+    index_options = np.setdiff1d(np.where(current_condition)[0], training_indices)
+    random_indices = np.random.choice(index_options, size=min(50000, len(index_options)), replace=False)
+
+    converted_indices = []
+    for index in random_indices:
[…]

### 2023-01-25 scripts/brain_atlas_bubbleplot.sh

@@ -1,0 +1,16 @@
+#!/bin/bash -l
+module load python3/3.8.10
+
+# Set SCC project
+#$ -P biophys
+
+# Request at least 128 GB of ram
+#$ -pe omp 8
+#$ -l mem_per_core=16G
+
+# Load packages
+source /projectnb/biophys/mariay/venvs/myenv/bin/activate
+export PYTHONPATH=/projectnb/biophys/mariay/venvs/myenv/lib/python3.8/site-packages/:$PYTHONPATH
+export PATH=/projectnb/biophys/mariay/venvs/myenv/bin:$PATH
+
+python /project/biophys/mouse\ scRNA-seq/scTOP\ explorations/scTOP/manuscript/brain_atlas_bubbleplot.py

### 2023-01-25 scripts/brain_atlas_script.sh

@@ -1,0 +1,16 @@
+#!/bin/bash -l
+module load python3/3.8.10
+
+# Set SCC project
+#$ -P biophys
+
+# Request at least 256 GB of ram
+#$ -pe omp 16
+#$ -l mem_per_core=16G
+
+# Load packages
+source /projectnb/biophys/mariay/venvs/myenv/bin/activate
+export PYTHONPATH=/projectnb/biophys/mariay/venvs/myenv/lib/python3.8/site-packages/:$PYTHONPATH
+export PATH=/projectnb/biophys/mariay/venvs/myenv/bin:$PATH
+
+python /project/biophys/mouse\ scRNA-seq/scTOP\ explorations/scTOP/manuscript/brain_atlas_accuracies.py

### 2023-01-25 scripts/tabula_sapiens_accuracies.py

@@ -1,0 +1,61 @@
+import numpy as np
+import pandas as pd
+from scanpy import read_h5ad
+
+import sys
+# Add path of module (temporary measure until pip install works)
+sys.path.append('/project/biophys/mouse scRNA-seq/scTOP explorations/scTOP/sctop')
+import sctop as top
+
+tabula_sapiens_file = read_h5ad('/projectnb/biophys/mariay/scTOP_manuscript_data/TS/TabulaSapiens.h5ad')
+
+# Ignore immune types and other types that aren't specific to organs
+types_to_keep = pd.read_csv('/project/biophys/mouse scRNA-seq/scTOP explorations/scTOP/tutorial/manuscript data/ts_types.csv', header=None)
+types_to_keep = types_to_keep.loc[:,0].str.replace(' \'|\'|,', '', regex=True)
+
+cell_id_types = tabula_sapiens_file.obs.loc[:,'cell_ontology_class'].astype('string') + " (" + tabula_sapiens_file.obs.loc[:,'organ_tissue'].astype('string') + ")"
+keeping_condition = tabula_sapiens_file.obs.loc[:,'cell_ontology_class'].str.contains("|".join(types_to_keep))
+kept_obs = tabula_sapiens_file.obs[keeping_condition]
+all_types, all_counts = np.unique(kept_obs.loc[:,'cell_ontology_class'].astype('string') + " (" + kept_obs.loc[:,'organ_tissue'].astype('string') + ")", return_counts=True)
+
+# Drop types with counts below 200 cells
+ts_types = all_types[all_counts>200]
+
+type_expressions_list = []
+training_indices = []
+
+for ts_type in ts_types:
+    current_type_indices = np.where(cell_id_types==ts_type)[0]
+    
+    # Choose n_training cells to use for the basis
+    n_training = 200
+    random_indices = np.random.choice(current_type_indices, n_training, replace=False)
+    training_indices.append(random_indices)
+    
+    current_raw = pd.DataFrame(tabula_sapiens_file.layers['raw_counts'][random_indices].T.toarray(),
+             index = tabula_sapiens_file.var['gene_symbol']
+            )
+    current_processed = top.process(current_raw, average=True)
+    type_expressions_list.append(current_processed)
+
+training_indices = np.concatenate(training_indices)
+ts_basis = pd.concat(type_expressions_list, axis=1)
+ts_basis.columns = ts_types
+
+all_organs = set(kept_obs.loc[:,'organ_tissue'].astype('string'))
+
[…]

### 2023-01-25 scripts/tabula_sapiens_script.sh

@@ -1,0 +1,16 @@
+#!/bin/bash -l
+module load python3/3.8.10
+
+# Set SCC project
+#$ -P biophys
+
+# Request at least 256 GB of ram
+#$ -pe omp 16
+#$ -l mem_per_core=16G
+
+# Load packages
+source /projectnb/biophys/mariay/venvs/myenv/bin/activate
+export PYTHONPATH=/projectnb/biophys/mariay/venvs/myenv/lib/python3.8/site-packages/:$PYTHONPATH
+export PATH=/projectnb/biophys/mariay/venvs/myenv/bin:$PATH
+
+python /project/biophys/mouse\ scRNA-seq/scTOP\ explorations/scTOP/tutorial/tabula_sapiens_accuracies.py

### 2023-07-21 manuscript/scTOP response to reviewers/review_response.tex
added revised manuscript

\documentclass[12pt,american]{scrartcl}
\usepackage{babel}
\usepackage[babel]{microtype}
\usepackage[babel]{csquotes}
\usepackage[margin=0.75in]{geometry}

\usepackage[journal={Development},
			manuscript={DEVELOP/2023/201873},
			editor={Paul Francois}]{reviewresponse}

\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{newcent}
%\usepackage[scaled]{beramono}

\usepackage[backend=biber,style=ieee,dashed=false,url=false,isbn=false,defernumbers=true,refsection=section]{biblatex}
\bibliography{literature.bib}

\usepackage{hyperref}

\title{scTOP: physics-inspired order parameters for cellular identification and visualization}
\author{Maria Yampolskaya, Michael J Herriges, Laertis Ikonomou, Darrell N Kotton, and Pankaj Mehta}

\begin{document}
% \maketitle

% % Cover Letter
% \input{cover_letter.tex}

% Reviewer 1
\reviewer

\begin{revcomment}
	the important fact that the method is linear is mentioned but its implication is not really discussed. Many people will think that linear method = PCA (for instance), and it should be mentioned that the linear computation used is different from unsupervised linear methods such as PCA and SVD ( because of the reference set). Also the method could prove useful way beyond single cell data because of this (for instance many people use PCA for morphometrics analysis, and I could see how having a reference dataset to perform computation would help there).
\end{revcomment}
\begin{revresponse}
	We agree that this method is general enough to be applied to other contexts. We have added the following section to the Supplementary Information to explain how scTOP relates to other linear methods:
	\begin{changes}
		\subsection*{Comparison to other linear methods}
          scTOP is mathematically and conceptually distinct from other linear methods such as Principal Component Analysis (PCA). PCA takes input data and finds the vectors in feature space along which the most sample-to-sample variation occurs. PCA is unsupervised and assumes nothing is known about the input data. The principal components need to be examined to determine their biological relevance. The major difference with scTOP is that scTOP takes input data and a reference basis of known cell types. The input data is decomposed onto the vectors defined by the reference basis instead of vectors corresponding to the most sample-to-sample variation. With scTOP, the resulting decomposition has a direct interpretation: alignment with each of the known cell types within the reference basis. It’s not possible to find the decomposition onto known cell types with PCA, because PCA doesn’t have the extra information about pre-defined cell type expression profiles.

          To summarize, scTOP and PCA both transform the sample data to a different coordinate system, but there are major differences between the spaces that these methods project onto. The following table breaks down the differences between these spaces.

          \begin{tabular}{ |p{2.5in}|p{2.5in}|} 
             \hline
             scTOP space & PCA space\\ 
             \hline
             \hline
             Defined by an input reference basis of known cell types & Calculated by PCA by finding the orthogonal vectors (principal components) that best explain the variation in data  \\ 
             \hline
             Can be non-orthogonal (e.g. can have highly-correlated cell types) & Orthogonal by definition  \\ 
             \hline
             Dimensions have clear biological interpretation: alignment with cell types & Dimensions need to be closely analyzed to be interpreted \\
             \hline
         \end{tabular}
	\end{changes}
\end{revresponse} \label{PCA}

\begin{revcomment}
	I see what the authors mean when they say it is important that the algorithms should give deterministic results, but in some cases, well controlled stochastic results could be necessary/important. I was thinking of MCMC based models, or similarly, instead of computing a deterministic order parameter, one could imagine algorithms giving meaningful probability scores while not being fully deterministic. In the end, the choice of the reference base certainly adds a quenched disorder component. Maybe the authors can elaborate a bit on this deterministic/stochastic issue because I suspect some people might not necessarily see the deterministic aspect as a strength 
\end{revcomment}
\begin{revresponse}
    We agree that stochastic results can be valuable. Rather than focus on the deterministic aspect of scTOP, we have edited the text to emphasize the importance of reproducibility rather than determinism. As you say, stochastic likelihood scores can also be useful; the important thing is the ability to make comparisons across analyses and data sets. We have changed the manuscript's mentions of determinism and stochasticity. For example, instead of ``Second, the algorithm should be deterministic, further enhancing reproducibility,'' the text now states:

    \begin{changes}
        Second, the results of the algorithm should be reproducible, so that direct comparison between data sets is possible.
    \end{changes}
\end{revresponse}

\begin{revcomment}
    the ‘mathematical details’ part is very condensed. It is certainly 
very readable for a physicist, but I do not think developmental 
biologists (even the ones interested in ‘modelling’) will be able to 
understand statements like ‘decorrelated versions of the conventional 
spin-glass magnetization’. I would recommend a short crash course for 
biologists on what an order parameter is, since it is the crucial 
concept to define scores. I like equations but I am not even sure they 
should be in the main text, I could for instance imagine a version of 
the paper where those concepts are explained in an intuitive/graphical 
way. At the very least more ‘intuitive’ details should be given to 
biologists.
\end{revcomment}
\begin{revresponse}
    As per the formatting guidelines for Development, we have moved the mathematical details to the ``Materials and Methods'' section after ``Discussion.'' We have also expanded the introduction section to give a more detailed explanation of what an order parameter is (the additions are in \textcolor{blue}{blue}).

    \begin{changes}
        Here, we present a new physics-inspired algorithm, single-cell Type Order Parameters (scTOP), that has all of these desired properties. The algorithm is based on the statistical physics concept of an order parameter -- a coarse-grained quantity that can distinguish between phases of matter. \textcolor{blue}{An order parameter takes on different values depending on the phase of the system. Typically, in a disordered phase, the corresponding order parameter is zero, and in an ordered phase it is nonzero.} Arguably the most famous example of an order parameter is magnetization in a ferromagnetic system, which is a collection of spins \textcolor{blue}{(in the simplest case, spins can be ``up'' or ``down'')} which align with one another. In a ferromagnet, the magnetization measures the alignment of spins and can be used to distinguish magnetic and paramagnetic phases. \textcolor{blue}{Below a particular temperature (known as the ``critical temperature''), the spins in the ferromagnet align with one another. In a simple model, the order parameter (magnetization, which measures the average spin) is approximately one if all the spins are up and approximately negative one if all the spins are down. When the temperature exceeds the critical temperature, the system changes from the magnetic phase to the paramagnetic phase, and the spins do not align; instead, they are randomly up or down, and the order parameter is approximately zero.}

        It is straightforward to generalize the idea of magnetization to more complex settings such as spin glasses and attractor neural networks. In the context of attractor-network epigenetic landscape models of cell identity, this takes the form of generalized magnetizations that measure how aligned a gene expression state is with each of the attractor basins for different cell fates. \textcolor{blue}{Whereas magnetization for the ferromagnetic system measured whether the system was in a magnetic or paramagnetic phase, these generalized magnetizations measure which phase a gene regulatory network is in (where the different phases correspond to different cell types).} These generalized magnetizations serve as natural order parameters for cellular identity and can be calculated directly from data. The scTOP algorithm builds on previous work by developing methods for integrating scRNA-seq data, greatly expanding the power and utility of this approach. 
    \end{changes}
\end{revresponse}

\begin{revcomment}
    in physics magnetization and spins could range from -1 to +1, here the
 order parameter is always positive (and lower than 1) as far as I can 
see. why ? In particular it is not entirely clear to me how things were 
put back between 0 and 1. Also, the maximum value of the order parameter
 computed from data seems to be around 0.5, can the authors comment on 
the magnitudes computed by their algorithm and how to interpet them ?
\end{revcomment}
\begin{revresponse}
    scRNA-seq data is very sparse, so it’s not possible to reach perfect alignment (score of +1). The reference basis contains aggregate profiles, where expressions have been averaged across cells of the same type. This removes many of the zeroes, but the single-cell samples still have many zeroes, resulting in misalignment. When samples are averaged in order to find aggregate scores, it’s possible to reach higher scores thanks to the reduction in zeroes. The figure with the aggregate samples from the Mouse Cell Atlas compared to organs from Tabula Muris illustrates these higher scores. It also shows a few scores that are negative.

    We have also added the following section to the Supplementary Information:
    \begin{changes}
        \subsection*{Interpreting scTOP scores}
        In theory, the attractor-network order parameters we have adapted for gene expression data range between negative one and positive one. An attractor state order parameter with a value of positive one means the system is perfectly aligned with that particular attractor state, and negative one means it is the exact opposite of that attractor state. Zero means that it is neither correlated nor anti-correlated with that attractor state.

        In practice, scTOP scores generally do not reach positive one. scRNA-seq data contains significant numbers of dropouts, and these zeroes decrease the magnitude of the scTOP score. When an aggregate score is taken, the effect of dropouts is decreased and the scTOP score increases. This is important to keep in mind when interpreting scTOP scores; any score above 0.1 is notable, and since the data is so noisy even the best-case score may be well below 1. It's helpful to have a control population to compare with, to have a measure of the range of best-case scTOP scores. This is also why it's best to compare distributions of scores rather than scores for individual cells, since noise plays a non-negligible role.

        scTOP scores are rarely negative. Negative values for scTOP scores correspond to cases where the sample gene expression profile is the opposite of the reference profile. For example, a score of -1 would indicate that all the genes which are highly-expressed in the reference are lowly-expressed in the sample, and vice versa. There are cases where the scores are low, negative values, indicating that the sample does not align well with the reference and that some genes which are highly- (lowly-) expressed in the reference are lowly- (highly-) expressed in the sample. From the data, it appears that it’s unlikely for gene networks to contain cell types which are opposites of one another. One possible reason is that some genes, like housekeeping genes, are similarly expressed across all cell types. Also, it’s possible that cells which expressed the exact opposite genes of any given cell type would not be able to survive.
    \end{changes}
\end{revresponse}

\begin{revcomment}
    when commenting on $S^\perp$, it is qualified as ‘biological noise’, but
 then reference to cell cycle genes, etc… is made. I can imagine genes 
related to, say, oscillating genes in the segmentation clocks, which 
would definitely be thrown away by the method (because they do not 
correspond to actual cell fates) but are definitely not developmental 
noise either. Maybe some comment should be added on situations like 
this, where gene dynamics not clearly associated with ‘fates’ (as 
defined by humans in their reference dataset) can not really be 
detected/visualized with this method. Along the same lines, it could be 
useful to show what happens with ‘virtual’ experiments where some 
reference fates are removed to contrast them with the situation with all fates
\end{revcomment}
\begin{revresponse}
    We agree that one of the limitations of scTOP is that it gives only information on cell differentiation. It cannot capture the dynamics of gene expression beyond the changing of alignment with cell identities defined by the reference basis. To clarify this point, we have edited the $S^{\perp}$ paragraph:

    \begin{changes}
        Biologically, $S^{\perp}$ contains information about processes that affect gene expression but are not associated with cell identity, such as the cell cycle. We refer to the $a^{\mu}$ as the scTOP scores. Each component of this vector measures the projection of the cell with gene expression profile $S_i$ onto the $\mu$ - the cell type in the reference basis. These scores can be used to accurately classify cell identity and provide a natural visualization of gene expression in the space of possible cell fates. Gene regulatory dynamics that are unrelated to transitions between cell types in the reference basis would not be captured by these scores; instead, these would contribute to $S^{\perp}$. As a result, these fate-unrelated dynamics cannot be visualized using scTOP.
    \end{changes}
\end{revresponse}

\begin{revcomment}
The authors only show ‘2D’ representation, for two order parameters. 
Is this because they only see binary choices in the data? Could we see 
in 3D a sequence of two binary choices for instance ? Is there anything 
biological to tell there?
\end{revcomment}
\begin{revresponse}
    It can be difficult to distinguish the positions of points in a 3D scatter plot. We have included figure \ref{3D plots} here to illustrate this. In this figure, we plot a clonal family from the hematopoietic data set in the main text. This is a case where multiple cell identities are of interest, but it is difficult to discern their relationships from a 3D plot. In cases where multiple fates are of interest, in binary choices or otherwise, we use pair-plots to show multiple 2D projections simultaneously.

    \begin{figure}
	\centering
    \includegraphics[scale=0.3]{figs/hem 3d neutrophil-monocyte.png}
	\caption{An example of a three-dimensional representation of scTOP scores, in the case of hematopoietic differentiation. Even with the inclusion of "shadows" (the corresponding 2D projections on the boundaries) it is difficult to determine the relationships between cells, which is why we use pair-plots in the main text instead. }
	\label{3D plots}
    \end{figure}
\end{revresponse}

\begin{revcomment}
    Table I : it is a bit unclear to me how the method was validated when 
Query data are used on the same reference set. I was expecting something
 classic like an initial separation between training/validating 
ensembles, etc… But it rather seems from the Supplement that the 
training set essentially was based on ‘well-chosen’ cells or something 
like this. It could be useful to give more details in the main text.
\end{revcomment}
\begin{revresponse}
    We do have the initial separation between training and test data. For clarity, we have changed the table column names to ``Test data source'' and ``Reference data source,'' and we have added the following line to the main text:
    \begin{changes}
        For cases where the training (or ``reference'') data source is the same as the test data source, we set aside a randomly-selected subset of cells for the training set and do not use them in the test data.
    \end{changes}

    The discussion of curating data sets in the SI refers to the cell types, rather than the cells, that were chosen for the reference basis. In other words, we discard cell types that have fewer than 100 corresponding cells. We do this because scRNA-seq data has a lot of dropouts, so we need to average across many cells to get a reasonable reference basis. We want to make sure the reference gene expression profiles are reflective of the true gene expression profiles of the cell types; making sure there are enough training data points per cell type is what we mean by having "well-sampled" reference profiles.

    To demonstrate the effect of not having enough cells per cell type, we have included figure \ref{sampling}. In this figure, we vary the number of cells sampled for the reference cell type profile, and find the corresponding scTOP score for the query data. With very few cells, the score is low since the reference type is poorly sampled. As the number of cells increase, the score plateaus. Ideally, we would have thousands of cells for every cell type, but unfortunately many single-cell data sets do not have so many cells per cell type.

    \begin{figure}
	\centering
    \includegraphics[scale=0.4]{figs/sampling.png}
	\caption{A plot of adult mouse lung stromal cells from the Mouse Cell Atlas projected onto a reference profile calculated from mesenchymal cells from the Atlas of Lung Development \cite{negretti2021single}. The x-axis indicates the number of cells sampled for the reference cell type, and the y-axis indicates the scTOP score. }
	\label{sampling}
    \end{figure}
\end{revresponse}

\begin{revcomment}
along the same line, all results depend on the quality of the 
reference set. Is there a way to somehow benchmark it ? For instance, I 
could imagine that a bad reference set could give a lot of false 
negatives or false positives, how would we see this ? (maybe by 
comparing with some unsupervised method ?)
\end{revcomment}
\begin{revresponse}
    In order to benchmark the quality of the reference bases, we measured the accuracy scores of control samples scored with those bases. In the manuscript, we provide examples of comparing the accuracy metrics with unsupervised methods from Abdelaal et al. In particular we compare the F1 score for scTOP using the CellBench and Allen Mouse Brain atlas with the F1 scores achieved for the same data sets using a variety of automatic cell identification algorithms, including unsupervised methods. The F1 score is a metric that is calculated from the true positives, false positives, and false negatives.

    Similar algorithms are often black boxes that are difficult to troubleshoot, but since scTOP simply uses linear algebra, it is possible to determine which genes are contributing to the scTOP scores. If a reference basis is giving a dubious result, it's possible to check if the top-contributing genes correspond to relevant biological processes. If not, this indicates that the reference basis may be poorly constructed, whether due to not having enough cells per cell type or not having distinct enough cell types.
\end{revresponse}

\begin{revcomment}
    The paper reads a bit long to me (12 figures, please check ). Maybe 
some figures with very similar messages/methods could be combined (e.g. 
4,5,6) and the writing could be a bit tightened there as well. There is a
 lot of ‘empty space’ in some figures (eg 9, 12) which could be 
tightened
\end{revcomment}
\begin{revresponse}
    In addition to ensuring that all figures match the formatting guidelines, we have made the following changes:
    \begin{enumerate}
        \item We have adjusted the layout and removed whitespace from figures 1, 5, and 6
        \item We have removed redundant panels (or moved them to the Supplementary Information), as in figures 5 and 9
        \item We have moved figure 8 to the Supplementary Information
        \item We have combined three figures into the panels of figure 4, and tightened the writing accordingly
    \end{enumerate}
\end{revresponse}

\begin{revcomment}
    1st sentence : « humans and other mammals… » I guess other animals too
\end{revcomment}
\begin{revresponse}
    We have fixed this line to say "humans and other animals."
\end{revresponse}

% Reviewer 2
\reviewer

\begin{revcomment}
	What is the relationship of scTOP to PCA? Presumably one could do PCA with C principal components for C distinct cell types and construct a new coordinate system as the average vector to each celltype. How should one think about this wrt to scTOP’s basis vectors? Can you perhaps provide a comparison of how the two methods fair when thinking about cell types or intermediates and even lineages
\end{revcomment}
\begin{revresponse}
 Although scTOP and PCA are both linear methods, there are significant differences between them. The principal components computed by PCA are the orthogonal dimensions that describe the most variance in the data. For example, the first principal component is the dimension along which the data varies the most. Taking C principal components would not provide information about alignment to the C cell types; instead, this coordinate system would give information about the top C dimensions along which there is the most variance. For example, if the input data corresponded to cells that were all of cell type A, the principal components would give information about how the gene expression varies between these cells, regardless of cell type. On the other hand, scTOP would identify all these cells as cell type A.

 We have added the following section to the Supplementary Information to explain how scTOP relates to other linear methods:
	\begin{changes}
		\subsection*{Comparison to other linear methods}
          scTOP is mathematically and conceptually distinct from other linear methods such as Principal Component Analysis (PCA). PCA takes input data and finds the vectors in feature space along which the most sample-to-sample variation occurs. PCA is unsupervised and assumes nothing is known about the input data. The principal components need to be examined to determine their biological relevance. The major difference with scTOP is that scTOP takes input data and a reference basis of known cell types. The input data is decomposed onto the vectors defined by the reference basis instead of vectors corresponding to the most sample-to-sample variation. With scTOP, the resulting decomposition has a direct interpretation: alignment with each of the known cell types within the reference basis. It’s not possible to find the decomposition onto known cell types with PCA, because PCA doesn’t have the extra information about pre-defined cell type expression profiles.

          To summarize, scTOP and PCA both transform the sample data to a different coordinate system, but there are major differences between the spaces that these methods project onto. The following table breaks down the differences between these spaces.

          \begin{tabular}{ |p{2.5in}|p{2.5in}|} 
             \hline
             scTOP space & PCA space\\ 
             \hline
             \hline
             Defined by an input reference basis of known cell types & Calculated by PCA by finding the orthogonal vectors (principal components) that best explain the variation in data  \\ 
             \hline
             Can be non-orthogonal (e.g. can have highly-correlated cell types) & Orthogonal by definition  \\ 
             \hline
             Dimensions have clear biological interpretation: alignment with cell types & Dimensions need to be closely analyzed to be interpreted \\
             \hline
         \end{tabular}
	\end{changes}
\end{revresponse}

\begin{revcomment}
    The authors say scTOP does not require data harmonization and 
is able to distinguish between technical and biological noise, but 
couldn’t batch effects between the reference dataset and the sample 
dataset effect the results? Can you use some dataset to illustrate this,
 there are few discussed in the several batch correction papers?
\end{revcomment}
\begin{revresponse}
    scTOP is able to identify cell types and give similar score distributions even across data sets. To demonstrate this, we analyzed dataset 4 from Tran et al. \cite{tran2020benchmark}. This data set contains pancreatic data from 5 different sources. We have added the following section to the Supplementary Information:

    \begin{changes}      
        \subsection*{Batch effects}
        scRNA-seq data produced using different sequencing platforms from different experiments may introduce variation in the form of batch effects. Because the pre-processing step in scTOP prioritizes the relative rather than absolute expression of genes, batch effects are mitigated.

        Tran et al. \cite{tran2020benchmark} consolidated datasets from different experiments for the purposes of benchmarking various batch-effect correction methods. One of these is a dataset containing scRNA-seq human pancreatic data from five different sources \cite{baron2016single} \cite{muraro2016single} \cite{segerstolpe2016single} \cite{wang2016single} \cite{xin2016rna}, using various sequencing technologies. We used scTOP to analyze these datasets. Figure \ref{batch effects} illustrates that scTOP is not significantly affected by differences between batches. Because the Baron et al. dataset had the most cells per cell type, we used it to create a reference basis consisting of alpha, beta, ductal, acinar, delta, gamma, stellate, and endothelial gene expression profiles. We projected cells from the other four experiments onto the Baron reference basis. The top row shows kernel density estimate (KDE) plots for cells that match the reference cell type (e.g. alpha scores for alpha cells) with solid lines, and cells that do not match the reference type but have higher scores than the other non-matching types (e.g. alpha scores for beta cells) with dashed lines. The bottom row shows box-and-whisker plots for the cells that match the reference type.

        Despite coming from different sources, the distributions of scores are similar. There is also a clear bimodal distribution for cells that match the reference type and cells that don't. This bimodality is a clear indication that batch effects don't cause misidentification.

        According to Ziegenhain et al. \cite{ziegenhain2017comparative}, Smart-seq2 detects the most genes per cell when compared to other sequencing methods of that time. This decrease in dropouts may explain why the Segerstolpe data (which is the only experiment that used Smart-seq2) has slightly higher projections than the other datasets.
    \end{changes}

    \begin{figure}
	\centering
    \includegraphics[scale=0.95]{figs/batch effects.pdf}
	\caption{Pancreatic data from different experiments using different scRNA-seq technologies. The reference basis uses data from Baron et al. \cite{baron2016single}, and the cells that are projected onto this basis are from Muraro et al. \cite{muraro2016single}, Segerstolpe et al. \cite{segerstolpe2016single}, Wang et al. \cite{wang2016single}, and Xin et al. \cite{xin2016rna} (A) Kernel density estimate (KDE) plot of alpha scTOP scores for alpha cells (solid) and beta cells (dashed), and a box-and-whisker plot of alpha scTOP scores for alpha cells. (B) KDE plot of beta scTOP scores for beta cells (solid) and delta cells (dashed), and a box-and-whisker plot of beta scTOP scores for beta cells. }
	\label{batch effects}
    \end{figure}
\end{revresponse}

\begin{revcomment}
    Authors mention that pre-processing to generate z-scores is essential.
 It is not clear why that should be, at least authors don’t explain it. 
For example, does a simpler, more intuitive normalization per cell of 
converting counts to fractions, work? I think there are many schemes of 
normalizations that have been proposed and having a clear analysis and 
discussion around that is essential for “users” who may be interested in
 using scTOP.
\end{revcomment}
\begin{revresponse}
    To address this comment, we have added the following subsection under the ``Preprocessing'' section of the Supplementary Information.

    \begin{changes}
    \subsection*{Preprocessing rationale}

    In order to mitigate batch effects, the preprocessing step in scTOP takes each cell and ranks each gene according to expression. These ranks are converted into percentiles, which are then transformed into z-scores assuming a normal distribution.

    By ranking the genes before converting to z-scores, we are ignoring the absolute values of RNA counts and instead focusing on the relative values. Some sequencing platforms may measure many more counts than others; although we can't recover the unmeasured counts that are lost via dropouts, we can remove the effect of differences in counts by ranking. We assume that each cell contains similar numbers of mRNA and that the actual number of RNA doesn't matter as much as the relative expression of different genes. For example, the highest-expressed gene in one cell should be comparable to the highest-expressed gene in another cell, regardless of RNA count. Ignoring the absolute values of RNA counts reduces the differences caused by different scRNA-seq technologies and experiments, as shown by the ``Batch effects'' section.

    The second preprocessing step is to convert the ranks to z-scores, assuming a normal distribution. This step results in a mean of zero and standard deviation of one. In data analysis, it is beneficial to standardize the data before comparing across datasets because this prevents a disproportionate bias towards features with large values. In the case of scTOP, having a mean of zero and a standard deviation of one is especially important because the underlying math relies on the dot product of two uncorrelated vectors to be zero and the magnitude of a vector to be one.

    By ranking and z-scoring, we remove undue weighting of genes according to magnitude of expression, instead prioritizing the relative expression.
    \end{changes}

\end{revresponse}

\begin{revcomment}
    About lineages, this is perhaps the most important application of 
scTOP as it brings out something beyond cell types that was the input to
 the system. The authors have some discussion of how having fewer cell 
types make it harder to map out the space and hence the lineages. But 
many applications where one is interested in cell fate dynamics of only a
 handful of cell types, sometimes one cell type going to two, is central
 to understand the underlying genetic logic of that drives the cascade 
of “bifurcations”. This feels like a big limitation, and I am curious if
 authors to say a little more on this, perhaps again by taking a 
concrete example, either with synthetic or real data.
\end{revcomment}
\begin{revresponse}
    One limitation of scTOP is that a certain number of cells are needed per cell type in order to create an accurate expression profile for that cell type. While having more cell types can indeed help map out the cell type subspace, most cell types can be removed from the reference basis and scTOP scores are still able to distinguish the remaining types (see Supplementary Information section ''Robustness of scTOP``, which shows the result of having few cell types in the reference basis).

    scTOP is able to visualize the cell fate dynamics of one cell type going to two. A concrete example can be found in the hematopoiesis lineage tracing data. In this dataset, clonal families with three time points reveal the dynamics of hematopoietic stem cells differentiating into mature blood cells. Figure 6 shows the scTOP scores of individual clonal families (i.e. sister cells that originated from the same stem cell) over time. One of these plots shows the case of one cell type going to two: progenitor cells going to neutrophil and monocytes. The plot with basophils, neutrophils, and eosinophils shows one cell type going to three other cell types. Another example of one cell type going to two is shown by the AT1/AT2 plots, where a lung alveolar progenitor cell differentiates into either AT1 or AT2.

    The biggest limitation of scTOP is the case where the cell type of a sample of interest is not included in the reference basis. In this case, all the corresponding scTOP scores will be low, unless the cell type of the sample is highly correlated with a cell type in the basis. For this reason, the scTOP scores may be sensitive to change when highly-correlated cell types are added or removed.
\end{revresponse}

\begin{revcomment}
    Also, on lineages, could some analysis be provided to see if expected gene expression changes happen along the directions of cell fate changes? This would provide some confidence on the predicted cell fate 
dynamics.
\end{revcomment}
\begin{revresponse}
    Figure 5 of the main text shows the development of mouse alveolar cells. The scTOP scores for AT1 and AT2 are shown for each time point of this time series data, and each cell is colored by the expression of AT1 or AT2 marker genes. These marker genes are the genes that are expected to change and become highly expressed when AT1 or AT2 fates are adopted. As shown by the plots, expression of these marker genes occurs along the directions of cell fate indicated by scTOP. The expected gene expression changes do occur with the dynamics visualized by scTOP.
\end{revresponse}

\begin{revcomment}
    You should explicitly mention “xi” is the basis in methods 
section II A. I don’t think its explicitly mentioned anywhere and would 
be useful for readers not familiar with these types of methods.
\end{revcomment}
\begin{revresponse}
    To address this comment, we have added the following line:

    \begin{changes}
         We refer to $\xi$ as the reference basis.
    \end{changes}
\end{revresponse}

\begin{revcomment}
    Figure 2 a,b,c are referred to in the text but the figure does not contain these labels.
\end{revcomment}
\begin{revresponse}
    To address this comment, we have edited the figure and added labels.
\end{revresponse}

\begin{revcomment}
    What is the interpretation of a negative scTOP score? Some of the plots contain small negative values. For example, in Figure 2 when plotting Ciliated vs AT2, cells high along the Ciliated axis become increasingly negative along the AT2 axis. Are these cells more or less like AT2 cells than the zero point? A similar result is seen in Figure 7 where high AT1 axis cells become increasing negative along the AT2 axis. Some other moderately high negative values are also seen in Figure 4 and 6.
\end{revcomment}
\begin{revresponse}
    To address the interpretation of negative scores, we have added the following section to the Supplementary Information:
    \begin{changes}
        \subsection*{Interpreting scTOP scores}
        In theory, the attractor-network order parameters we have adapted for gene expression data range between negative one and positive one. An attractor state order parameter with a value of positive one means the system is perfectly aligned with that particular attractor state, and negative one means it is the exact opposite of that attractor state. Zero means that it is neither correlated nor anti-correlated with that attractor state.

        In practice, scTOP scores generally do not reach positive one. scRNA-seq data contains significant numbers of dropouts, and these zeroes decrease the magnitude of the scTOP score. When an aggregate score is taken, the effect of dropouts is decreased and the scTOP score increases. This is important to keep in mind when interpreting scTOP scores; any score above 0.1 is notable, and since the data is so noisy even the best-case score may be well below 1. It's helpful to have a control population to compare with, to have a measure of the range of best-case scTOP scores. This is also why it's best to compare distributions of scores rather than scores for individual cells, since noise plays a non-negligible role.

        scTOP scores are rarely negative. Negative values for scTOP scores correspond to cases where the sample gene expression profile is the opposite of the reference profile. For example, a score of -1 would indicate that all the genes which are highly-expressed in the reference are lowly-expressed in the sample, and vice versa. There are cases where the scores are low, negative values, indicating that the sample does not align well with the reference and that some genes which are highly- (lowly-) expressed in the reference are lowly- (highly-) expressed in the sample. From the data, it appears that it’s unlikely for gene networks to contain cell types which are opposites of one another. One possible reason is that some genes, like housekeeping genes, are similarly expressed across all cell types. Also, it’s possible that cells which expressed the exact opposite genes of any given cell type would not be able to survive.
    \end{changes}
\end{revresponse}

\begin{revcomment}
    Are there any dataset-specific preprocessing steps done? For example filtering cells or genes based on quality control metrics. If so, they 
should be reported so that the results are reproducible.
\end{revcomment}
\begin{revresponse}
    We took the datasets as they were deposited and did not filter any genes. As described in the Supplementary Information section ``Reference basis construction,'' we discarded cells if there were fewer than 100 cells of the corresponding cell type. This is because the reference basis needs at least 100 cells per cell type in order to represent accurate expression profiles for the given cell types.
\end{revresponse}

\begin{revcomment}
    The code in the linked Github repository contains a tutorial, but the
 code used to run scTOP on the other datasets should also be provided.
\end{revcomment}
\begin{revresponse}
    We have made the following changes to address this comment:

    \begin{changes}
        We have implemented scTOP in an easy-to-use Python package, available on the \href{https://pypi.org/project/scTOP/}{Python Package Index} and \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP}{Github}. The code for the analyses in this paper is available in a separate \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP-manuscript}{Github repository}.
    \end{changes}
\end{revresponse}

\printbibliography

\end{document}

### 2023-07-21 manuscript/scTOP (Company of Biologists LaTeX template)/SI.tex
added revised manuscript

--- 2023-01-25 manuscript/scTOPv5.tex
@@
 \usepackage{xcolor}
-\usepackage{hyperref}
 \setlength {\marginparwidth }{2cm}
@@
 
-\title{ scTOP: physics-inspired order parameters for cellular identification and visualization}
+\title{scTOP: physics-inspired order parameters for cellular identification and visualization}
 
@@
 \affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
-\author{Michael Herriges}
+\author{Michael J Herriges}
 \affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
@@
 \affiliation{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
-\author{Darrell Kotton}
+\author{Darrell N Kotton}
 \affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
@@
 
-\begin{abstract}
-Advances in single-cell RNA-sequencing (scRNA-seq) provide an unprecedented window into cellular identity.
-The increasing abundance of data requires new theoretical and computational frameworks for understanding cell fate determination, accurately classifying cell fates from expression data, and integrating knowledge from cell atlases.
-Here, we present single-cell Type Order Parameters (scTOP): a statistical-physics-inspired approach for constructing ``order parameters'' for cell fate given a reference basis of cell types.
-scTOP can quickly and accurately classify cells at a single-cell resolution, generate interpretable visualizations of developmental trajectories, and assess the fidelity of engineered cells.
-Importantly, scTOP does this without using feature selection, statistical fitting, or dimensional reduction (e.g., UMAP, PCA, etc.).
-We illustrate the power of scTOP utilizing a wide variety of human and mouse datasets (both {\it in vivo} and {\it in vitro}).
-By reanalyzing mouse lung alveolar development data, we characterize a transient perinatal hybrid alveolar type 1/alveolar type 2 (AT1/AT2) cell population that disappears by 15 days post-birth and show that it is transcriptionally distinct from previously identified adult AT2-to-AT1 transitional cell types.
-Visualizations of lineage tracing data on hematopoiesis using scTOP confirm that a single clone can give rise to as many as three distinct differentiated cell types.
-We also show how scTOP can quantitatively assess the transcriptional similarity between endogenous and transplanted cells in the context of murine pulmonary cell transplantation.
-Finally, we provide an easy-to-use Python implementation of scTOP.
-Our results suggest that physics-inspired order parameters can be an important tool for understanding development and characterizing engineered cells.
-\end{abstract}
-
 \maketitle
 
-\section{Introduction}
-Humans and other mammals have many specialized cells that arise through a complex developmental process  \cite{zeng_what_2022}.
-At the blastocyst/epiblast stages of embroynic development, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers \cite{Wolpert_2019}.
-However, as embryos mature, cellular identity becomes less plastic and more specified through a complex spatiotemporal process involving signaling and other environmental cues.
-The questions of how and why cells end up in one fate or another are fundamental to developmental biology \cite{stephenson_intercellular_2012, zhu2020principles, rand2021geometry}.
-More practically, understanding the process of cell fate specification is crucial for developing directed differentiation protocols for engineering cells \emph{in vitro}, with important clinical applications for human health and disease \cite{rossant_impact_2011, rowe2019induced, basil2020cellular}.
+\section*{Supplementary Information}
+\beginsupplement
 
-One powerful set of techniques for characterizing cellular phenotype is transcriptome-wide measurement using single-cell RNA sequencing (scRNA-seq).
-Typical scRNA-seq experiments involve thousands of genes and dozens to millions of cells.
-For this reason, this data is intrinsically high-dimensional and requires specialized computational tools for analysis and interpretation \cite{zhu2020principles, setty2019characterization, schiebinger2019optimal}.
-This task is especially difficult because the technical challenges of sequencing RNA from individual cells result in data that is noisy and sparse \cite{hicks2018missing, lahnemann_eleven_2020, argelaguet2021computational}.
-Genes with relatively low expression are often recorded as not being expressed at all, resulting in zero counts called dropouts.
-Experiment-to-experiment differences in sequencing methods, platforms, and other difficult-to-standardize conditions result in technical variations in scRNA-seq measurements called batch effects.
-These challenges merit careful consideration of how to normalize and analyze the data in a way that permits the study of biological variation without conflation by technical variation.
+\subsection{Interpreting scTOP scores}\label{interpretation}
+In theory, the attractor-network order parameters we have adapted for gene expression data range between negative one and positive one.
+An attractor state order parameter with a value of positive one means the system is perfectly aligned with that particular attractor state, and negative one means it is the exact opposite of that attractor state.
+Zero means that it is neither correlated nor anti-correlated with that attractor state.
 
-scRNA-seq is often used for cellular identification.
-The most frequently used workflow involves applying dimensional reduction (such as Uniform Manifold Approximation and Projection (UMAP) \cite{mcinnes_umap_2018} or SPRING \cite{weinreb_spring_2018}) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells.
-These groups are then often manually annotated according to expression levels of known marker genes.
-Machine learning classifiers such as support vector machines are also commonly used to identify cells after training with labeled data \cite{abdelaal_comparison_2019}.
-Another common analysis task is trajectory inference, where the goal is to infer developmental trajectories from static time shots (reviewed in \cite{saelens2019comparison,wagner2020lineage}).
-Trajectory inference arranges cells on a low-dimensional manifold and assigns pseudo-times according to transcriptomic similarity.
-There are also methods such as partition-based graph abstraction \cite{wolf_paga_2019} that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions.
-RNA-velocity-based methods like Dynamo \cite{qiu2022mapping} use splicing information or time-resolved data to predict paths of differentiation from RNA data.
+In practice, scTOP scores generally do not reach positive one.
+scRNA-seq data contains significant numbers of dropouts, and these zeroes decrease the magnitude of the scTOP score.
+When an aggregate score is taken, the effect of dropouts is decreased and the scTOP score increases.
+This is important to keep in mind when interpreting scTOP scores; any score above 0.1 is notable, and since the data is so noisy even the best-case score may be well below 1.
+It's helpful to have a control population to compare with, to have a measure of the range of best-case scTOP scores.
+This is also why it's best to compare distributions of scores rather than scores for individual cells, since noise plays a non-negligible role.
 
-Despite the power and utility of these computational tools, most commonly employed methods for analyzing scRNA-seq data share certain limitations.
-Dimensional reduction is often one of the first steps in visualizing scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships \cite{chari_specious_2021}.
-Distorted embeddings inconsistently represent distances between individual data points and between cell types, making it difficult to interpret cell fate transitions.
-Similar issues arise in trajectory inference and RNA-velocity methods because local neighborhoods can also vary substantially with the choice of embedding.
-Another limitation of these methods is the lack of reproducibility inherent to stochastic machine learning methods like t-SNE and UMAP, which output different results on separate executions even when hyperparameters remain the same \cite{wattenberg_how_2016}.
-Many algorithms also perform feature selection (e.g., limiting their analysis to the most variable genes or highly-expressed genes) or require choosing parameters such as perplexity, making results even more sensitive to the inclusion or exclusion of different datasets in the analysis.
+scTOP scores are rarely negative.
+Negative values for scTOP scores correspond to cases where the sample gene expression profile is the opposite of the reference profile.
+For example, a score of -1 would indicate that all the genes which are highly-expressed in the reference are lowly-expressed in the sample, and vice versa.
+There are cases where the scores are low, negative values, indicating that the sample does not align well with the reference and that some genes which are highly- (lowly-) expressed in the reference are lowly- (highly-) expressed in the sample.
+From the data, it appears that it’s unlikely for gene networks to contain cell types which are opposites of one another.
+One possible reason is that some genes, like housekeeping genes, are similarly expressed across all cell types.
+Also, it’s possible that cells which expressed the exact opposite genes of any given cell type would not be able to survive.
 
-An ideal analysis algorithm for scRNA-seq data should possess certain desirable properties.
-First, the algorithm should incorporate explicit assumptions that can be kept consistent between analyses.
-Second, the algorithm should be deterministic, further enhancing reproducibility.
-Third, the axes in visualization should be interpretable, intuitive, and remain fixed across analysis and datasets.
-Fourth, the algorithm should assign a continuous value to cell identity instead of assigning discrete identity labels, thus enabling a more subtle and fine-grained analysis of cellular identity and developmental trajectories that reflect the continuous nature of biology.
-Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single-cell atlases now being generated \cite{quake2022decade}.
+\subsection{Comparison to other linear methods}
+scTOP is mathematically and conceptually distinct from other linear methods such as Principal Component Analysis (PCA).
+PCA takes input data and finds the vectors in feature space along which the most sample-to-sample variation occurs.
+PCA is unsupervised and assumes nothing is known about the input data.
+The principal components need to be examined to determine their biological relevance.
+The major difference with scTOP is that scTOP takes input data and a reference basis of known cell types.
+The input data is decomposed onto the vectors defined by the reference basis instead of vectors corresponding to the most sample-to-sample variation.
+With scTOP, the resulting decomposition has a direct interpretation: alignment with each of the known cell types within the reference basis.
+It’s not possible to find the decomposition onto known cell types with PCA, because PCA doesn’t have the extra information about pre-defined cell type expression profiles.
 
-Here, we present a new physics-inspired algorithm, single-cell Type Order Parameters (scTOP), that has all of these desired properties.
-The algorithm is based on the statistical physics concept of an order parameter -- a coarse-grained quantity that can distinguish between phases of matter.
-Arguably the most famous example of an order parameter is magnetization in a ferromagnetic system (i.e., a system that can spontaneously magnetize).
-In a ferromagnet, the magnetization measures the alignment of spins and can be used to distinguish magnetic and paramagnetic phases \cite{landau2013statistical,sethna2021statistical}.
-It is straightforward to generalize the idea of magnetization to more complex settings such as spin glasses and attractor neural networks \cite{amit1985spin}.
-In the context of epigenetic landscape models of cell identity, this takes the form of generalized magnetizations that measure how aligned a gene expression state is with each of the attractor basins for different cell fates \cite{lang_epigenetic_2014,pusuluri_cellular_2017}.
-These generalized magnetizations serve as natural order parameters for cellular identity and can be calculated directly from data \cite{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}.
-The scTOP algorithm builds on previous work by developing methods for integrating scRNA-seq data, greatly expanding the power and utility of this approach.
+To summarize, scTOP and PCA both transform the sample data to a different coordinate system, but there are major differences between the spaces that these methods project onto.
+The following table breaks down the differences between these spaces.
 
-The inputs to scTOP are the scRNA-seq measurements of query data as well as a reference basis containing the archetypal gene expression profiles for cell types of interest.
-The outputs of scTOP are the coordinates of the query data in cell type order parameter space (i.e., generalized magnetizations), with each output dimension serving as a measure of similarity between the query data and a distinct cell type in the reference basis.
-In this way, scTOP naturally projects gene expression deterministically onto a space with dimension equal to the number of cell types in the reference basis.
-These generalized coordinates can be used for both cell type identification and visualization.
-One appealing feature of scTOP is that the assumptions of the algorithm are explicit and are contained entirely in the choice of reference basis.
-This allows direct comparison across datasets and settings since the coordinates of and distances between data points do not change as long as one uses the same reference basis.
-One important limitation of scTOP worth noting is that it cannot be easily used to identify novel cell fates.
-The underlying reason for this is that scTOP assumes one already knows what cell types are of interest in the form of a reference basis and, for this reason, is less suited for exploratory work with undocumented cell types.
-
-In this paper, we show that scTOP can be used to reliably assess cell identity and accurately classify cell type with high accuracy at the resolution of individual cells.
-We demonstrate this using mouse and human samples from lungs, brains, and other organs, including transplanted cells that were generated through the directed differentiation of pluripotent stem cells (PSCs).
-By analyzing lineage tracing data and embryonic development data, we also establish scTOP's capacity to quantify transitions between cell types, including bifurcations between closely related cell fates.
-We also show that scTOP can be used to quantitatively assess transcriptional similarity of engineered and natural cells by analyzing murine pulmonary transplantation experiments.
-
-We have implemented scTOP in an easy-to-use Python package, available on the \href{https://pypi.org/project/scTOP/}{Python Package Index} and \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP}{Github}.
-The code written for the analyses in this paper is available on \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP-manuscript}{Github}.
-
-\section{Methods}
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.8]{figs/scTOP manuscript fig1 - no aggregate.pdf}
-	\caption{Steps involved in the scTOP algorithm.
-(a) Each cell in the query data is preprocessed independently of the other cells in the dataset.
-(b) scRNA-seq atlases, such as the Mouse Cell Atlas, are used to define the reference basis in the algorithm.
-(c) scTOP scores are the projections of query data onto the non-orthogonal subspace of cell types.
-Since there is no statistical fitting, no tuning parameters are involved.
-(d) The process of finding scTOP scores is shown in the grey section, and the blue section illustrates cell-type space.
-This space has as many dimensions as there are cell types in the reference basis.
-We can visualize this concept in 1, 2, or 3 dimensions, depending on which cell types are most relevant.
-In the 3-dimensional case, the shadows on the bounds of the scatter plot are included to better illustrate the position of points in 3-dimensional space.}
-	\label{FIG:1}
-\end{figure}
-
-\subsection{Mathematical Details}
-To calculate cell type similarity scores, scTOP uses a linear algebra projection method inspired by the concept of cell types as attractors in a dynamical system \cite{huang_cell_2005}.
-This projection method has been applied previously to bulk RNA-seq samples, and we will summarize the theoretical background here \cite{lang_epigenetic_2014}.
-A cell type is a state of the gene regulatory network.
-A natural way to represent attractors in this system is with an attractor neural network, where attractors correspond to different cell types.
-If we represent each gene as a node in a network with an associated value measuring gene expression (positive value representing high gene expression, negative value indicating low gene expression), then a cell type may be denoted by a vector of gene expression values $S_i$, where $i=1,\ldots, G$ spans the $G$ genes being measured.
-The gene expression profiles
-corresponding to the $C$ cell type attractors are also $G$ dimensional vectors $\xi_i^\mu$, where $\mu=1,\ldots, C$ spans the $C$ cell types being stored in the network.
-In what follows, we assume that $G > C$; in other words, the number of genes being measured is greater than the number of cell types.
-
-Cell types are often highly similar in their patterns of gene regulation.
-Kanter and Sompolinsky \cite{kanter_associative_1987} defined a pattern storage method for spin-glass-like neural networks where even correlated patterns robustly act as attractors.
-For the same system, they also defined order parameters: generalized magnetizations $a^{\mu}$, where $\mu$ iterates through each of the stored cell type attractors.
-The $a^\mu$ can be understood as de-correlated versions of the conventional spin glass magnetization order parameter that measures the similarity between a given network state and the network state $\xi^\mu$.
-Explicitly, the $C$ generalized magnetizations $a^\mu$ can be calculated for a gene expression state $S_i$ via the expression
-\be \label{op_eqn}
-a^\mu= \sum_{\nu=1}^C\sum_{j=1}^G [A_{\mu \nu}]^{-1}\xi_j^\nu S_j.
-\ee
-where
-\be
-A_{\mu \nu}= \sum_{j}^G  \xi_j^\mu \xi_j^\nu,
-\ee
-is the overlap matrix of gene expression profiles of different cell types.
-Previous works have shown the order parameters $a^{\mu}$ provided an excellent similarity score for analyzing bulk RNA-seq data \cite{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}.
-However, bulk RNA-seq measurements average gene expression over many cells, rendering it impossible to measure cell fate transitions beyond an average tissue state.
-
-scTOP uses the same order parameters, $a^{\mu}$, to track cell type at the resolution of individual cells, making it possible to directly observe cells in different stages of differentiation.
-In both attractor neural networks and previous applications to bulk RNA-seq data, the expression vectors $S_i$ were binary variables.
-However, we have found that for working with scRNA-seq data, it is helpful to treat $S_i$ as continuous.
-
-\subsection{Preprocessing}
-
-As shown in figure \ref{FIG:1} (a), the first stage of the algorithm is preprocessing and normalization of the input data.
-In our algorithm, measurements from each cell are normalized independently and do not depend on the expression profiles of any other cells in the dataset.
-We assume that the scRNA-seq data has been processed, resulting in a gene-wise count matrix: for each cell, each gene has an integer value that counts the number of assigned RNA reads.
-Genes are then rank-ordered.
-The rank order is subsequently converted to a z-score representing the percentile rank of the expression of a gene within the cell relative to all other genes being measured (assuming a normal distribution with mean zero and standard deviation one, for the sake of simplicity).
-For example, a gene that has higher expression than $97.8\%$ of genes being measured is assigned a score of $z=2$, while a gene that has higher expression than half of the genes is assigned a score of $z=0$.
-The details of the process and its effects on data distributions are described in Appendix section \ref{preprocessing}.
-The preprocessing step is a key component of the algorithm.
-By processing each cell independently, the output for one cell is independent of other cells included in the analysis.
-This is in contrast with other algorithms which normalize gene expression across cells by, for example, selecting the most variable genes.
-
-\subsection{Construction of reference basis}
-
-Constructing a representative, accurate reference basis for the data being analyzed is vital to the scTOP algorithm.
-This process is shown in figure \ref{FIG:1} (b).
-First, relevant single-cell RNA-seq atlases are gathered to be processed.
-For example, in analyzing mouse lung cells, we used the Mouse Cell Atlas \cite{noauthor_mapping_nodate}, which contains single-cell samples of mouse tissues across all major organs.
-Once the relevant atlases are identified, we take an average across each cell population that corresponds to a distinct cell type.
-To mitigate the potent effects of noise and avoid training data imbalances, a sufficient number of cells must be included in each population.
-We have empirically found that the minimum sample size to achieve reasonable results is 100-200 cells.
-Noise affects the algorithm most strongly when it appears in the basis since the decorrelation operation involved in the projection separates cells according to the canonical expression levels included in the reference basis.
-After each cell population is averaged to create the gene expression profiles for the desired cell types, each cell type profile is preprocessed separately using the same procedure as that used for processing single cells: normalization followed by z-scoring.
-This creates a modular reference basis where cell types can easily be replaced, removed, or added at will.
-
-\subsection{Calculation of scTOP scores}
-
-Given a reference basis, we can calculate the order parameters $a^\mu$ using Eq.~\ref{op_eqn}.
-Although these order parameters were originally inspired by attractor neural networks and spin glass physics, the $a^{\mu}$ can also be understood as non-orthogonal projections onto the subspace spanned by the reference basis (i.e., cell type space).
-This is illustrated in figure \ref{FIG:1} (c).
-A single-cell RNA-seq measurement of a sample can be thought of as a vector in $G$-dimensional gene expression space, where $G$ is the number of genes.
-The $C$-cell types in the reference basis are also each represented by a $G$-dimensional vector and form a non-orthogonal basis for the $C$-dimensional cell type space (we assume $C < G$).
-To find the coordinates $a^\mu$ in cell type space, we project the sample vector onto the reference basis using Eq.~\ref{op_eqn}.
-
-Mathematically, we can always write the sample vector as a sum of the projected components and a component $S^{\perp}_i$  perpendicular to the cell type subspace: 
-\be
-S_i = \sum_\mu a^{\mu} \xi^{\mu}_i + S^{\perp}_i.
-\ee
-Biologically, $S^{\perp}$ contains information about biological processes that introduce variation in transcriptional profile such as cell cycle as well as expression resulting from technical and biological noise.
-We refer to the $a^\mu$ as the scTOP scores.
-Each component of this vector measures the projection of the cell with gene expression profile $S_i$ onto the $\mu$-the cell type in the reference basis.
-These scores can be used to accurately classify cell identity and provide a natural visualization of gene expression in the space of possible cell fates (see figure \ref{FIG:1} (e)).
-
-In general, we are interested in calculating scTOP scores for individual cells.
-However, it is also sometimes useful to compute ``aggregate'' scTOP scores for cellular populations by averaging the mRNA counts \textit{first}, then preprocessing the data.
-This produces a gene expression profile and a set of scTOP aggregate scores for each population.
-Aggregate scores tend to be higher because averaging over the RNA counts mitigates the well-documented effects of scRNA-seq noise \cite{hicks2018missing, lahnemann_eleven_2020}; the averaged gene expression profile for the populations are not as sparse as individual profiles and thus are more similar to the averaged profiles that form the reference basis.
-
- In the appendix, we give a detailed discussion of scTOP, including the effects of basis choice, score distributions for correct and incorrect cell types, and other technical details.
-
-\subsection{Assumptions and computational complexity}
-
-scTOP does not have any stochasticity, require tuning of hyper-parameters, or statistical fitting procedures.
-The assumptions of the algorithm are explicit: that the gene expression profiles included in the reference basis are truly representative of the relevant cell types and that the reference cell types correspond to distinct cellular populations.
-We have found that constructing a reference basis for a cell type typically requires at least 100 cells.
-Additionally, if the cell types are not sufficiently different from one another, as in early embryonic development or certain stromal cells where the same cell type is compared between organs, the resulting scTOP scores may be unreliable and exhibit extreme sensitivity to small changes in the choice of reference basis.
-More information on the limitations of scTOP may be found in the appendix.
-Generally, outside these very limited edge cases, we have found that scTOP can reliably distinguish even extremely closely related cells (as shown in the example of pulmonary alveolar epithelial type 1 (AT1) and type 2 (AT2) cell lineages below).
-
-Finally, we note that scTOP is extremely computationally efficient.
-The core of the scTOP is equation \ref{op_eqn}.
-This requires a single matrix inversion involving the reference basis that can be precomputed.
-For this reason, the computational complexity of scTOP scales \emph{linearly} with the number of cells.
-The end result is that scTOP takes milliseconds to run, even for very large datasets.
-\section{Results}
-
-\subsection{Benchmarking and validation: Classification of cell fate}
-
-scTOP scores can be used to reliably and accurately predict cell identity.
-To show this, we applied the algorithm to several scRNA-seq datasets across species and laboratories.
-scTOP is organism-agnostic as long as an appropriate reference basis is used.
-Mice and humans are among the most common subjects of scRNA-seq measurements, and we restrict our analysis in this manuscript to data from these organisms.
-To do so, we make use of single-cell atlases, such as the Mouse Cell Atlas \cite{noauthor_mapping_nodate} and the Atlas of Lung Development \cite{negretti_single-cell_2021}.
-
-Table \ref{table:1} lists each of the datasets examined in this paper with their respective scTOP score accuracies.
-TopN refers to the percent of cells whose true cell types were in the top N scTOP scores, with a score greater than 0.1.
-Unspecified refers to the percent of cells for which the highest scTOP score was less than 0.1; this represents a failure of the algorithm to confidently identify the cell.
-The accuracies across species and tissues are high, and the unspecified percentages are very low, even in datasets exceeding a million cells.
-
-\begin{table}[h!]
-\centering
-\begin{tabular}{| c | c | c | c | c | c | c |}
+\begin{tabular}{ |p{3in}|p{3in}|} 
  \hline
-Query Data & Species & Reference Basis & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Total cells \\ 
-\hline
-Kotton lab & Mouse & Mouse Cell Atlas & 97.19 & 99.77 & 2.69 & 4,805 \\ 
-\hline
-Brain atlas & Mouse & Self & 86.28 & 96.89 & 1.78 & 1,161,041 \\
-\hline
-CellBench & Human & Self & 100 & 100 & 0 & 2,822 \\
-\hline
-Lung atlas & Human & Self & 79.78 & 85.16 & 3.32 & 2,952 \\  
+ scTOP space & PCA space\\ 
  \hline
  \hline
-Data set & Species & Reference Basis & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Number of cell types \\
+ Defined by an input reference basis of known cell types & Calculated by PCA by finding the orthogonal vectors (principal components) that best explain the variation in data  \\ 
  \hline
-Mouse Cell Atlas & Mouse & Tabula Muris & 79.17 & 87.5 & 0 & 48 \\
+ Can be non-orthogonal (e.g. can have highly-correlated cell types) & Orthogonal by definition  \\ 
  \hline
+ Dimensions have clear biological interpretation: alignment with cell types & Dimensions need to be closely analyzed to be interpreted \\
+ \hline
+\end{tabular}
 
-\end{tabular}
-\caption{Accuracy scores of included sample types for each of the described data sets.}
-\label{table:1}
-\end{table}
-
-scTOP performs comparably to state-of-the-art methods for cell identification.
-Abdelaal et al.
-compared the performance of automatic cell identification algorithms for various datasets, including the CellBench dataset and an older version of the Allen Mouse Brain atlas \cite{abdelaal_comparison_2019}.
-They found that the median F1-score for the CellBench dataset ranged from 0.9 to 1; scTOP has a median F1-score of 1 for the same data.
-The version of the Allen Mouse Brain atlas they analyzed contained three levels of cell population annotations, with 3, 16, and 92 populations for each annotation level.
-The median F1-score ranged from 0.64 to 1 for the 16-population case and 0 to 0.98 for the 92-population case.
-The version of the brain atlas we analyzed contained 40 populations, and scTOP had a median F1-score of 0.88 for this dataset.
-
-\subsubsection{Example 1: Lung lineages}
-To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from Herriges et al.
-\cite{herriges_durable_2022} onto the mouse cell atlas.
-The Herriges et al.
-dataset measures gene expression levels of six different specialized lung types from healthy mice: AT1, AT2, basal, ciliated, secretory, and neuroendocrine (See figure \ref{herriges} (a)).
-The reference basis used to find the individual scTOP scores for this data set was derived from the Mouse Cell Atlas.
-Even though the Herriges et al.
-dataset and the Mouse Cell Atlas were created in different laboratories using different sequencing methods, the resulting scTOP scores distinctly separate the Herriges et al.
-dataset into the expected cell types.
-Furthermore, the individual scTOP scores strongly correlate with the expression of lineage marker sets (figure \ref{herriges} (b)).
-Altogether this suggests that scTOP provides a reliable method to annotate scRNA-seq datasets without the need for the expertise required to generate lineage marker gene sets.
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.45]{figs/herriges colored by type.png}
-        \includegraphics[scale=0.5]{figs/herriges colored by signature.png}
-	\caption{scTOP identifies mouse lung cell types.
-Using the Mouse Cell Atlas as the reference basis, we show that mouse lung types separate clearly on the scTOP score axes.
-The data points correspond to individual cells, and in the first row of plots, the marker shape and color indicate the true cell type as determined by annotations from Herriges et al.
-The axes used are scTOP scores for similarity with lung ciliated cells and alveolar types 1 and 2.
-The second, third, and fourth rows of plots have individual cells colored by the marker genes for the type indicated by the color bars at the end of each row (see appendix \ref{alveolar markers} for details on marker genes used).}
-	\label{herriges}
-\end{figure}
-
-\subsubsection{Example 2: Mouse Brain Atlas}
-
-The mouse brain atlas \cite{yao_taxonomy_2021} sequenced over one million cells and clustered them into 42 subclasses and 101 supertypes.
-For a reference basis, we reserved a subset of 200 cells from each of the identified subclasses to use as training data.
-The test data, from which the accuracy score was calculated, was the remaining cells of each type not included in the training data.
-Ten different cell types are visualized in a grid in figure \ref{mouse_brain} (a).
-The color of each grid square corresponds to the average scTOP score of cells annotated as the type on the y-axis, with the cell type corresponding to the scTOP dimension on the x-axis.
-As shown by the distinct blue diagonal, the average scTOP score correctly matches the corresponding annotated cell type.
-This demonstrates that scTOP is able to distinguish cell fates even from closely related neural lineages.
-
-\begin{figure}
-	\centering
-    \includegraphics[scale=0.75]{figs/mouse brain.png}
-	\caption{scTOP identifies mouse brain cell types.
-(a) Heatmap showing the average scTOP score for individual cells of the type indicated on the y-axis, compared to the reference types on the x-axis.
-A subset of cells from the Mouse Brain Atlas is used as the reference basis to query other cells from the same data set.
-The diagonal indicates that scTOP accurately matches query cells to the true reference type.
-(b) Histograms showing the distribution of scTOP scores for individual cells of the type labeling the x-axis.
-The scores shown are of the same type as the sample data.
-The far left histogram shows the distribution of Astro scores for Astro cells, and the average of this distribution is indicated by the color of the upper left corner box in the heatmap shown in (a).}
-	\label{mouse_brain}
-\end{figure}
-
-\subsubsection{Example 3: Matching cell types to tissues}
-
-scTOP can also be used to match cells with tissues/organs.
-We illustrate this by comparing cells from the Mouse Cell Atlas with a reference basis consisting of Tabula Muris organs \cite{schaum_single-cell_2018}.
-To do so, we averaged RNA counts across cell populations from the Mouse Cell Atlas, then preprocessed the averaged data and calculated scTOP scores.
-These aggregate pseudo-bulk scores tend to be higher than individual cell scores because averaging over many cells compensates for the dropout effect.
-As shown in figure \ref{tabula_muris}, we found the aggregate scTOP scores for cells from the Mouse Cell Atlas were correctly and strongly identified with the annotated tissue of origin in the Tabula Muris atlas.
-Parenchymal cell types were the most associated with the correct organ, while stromal cells were sometimes misidentified -- for example, the mammary gland endothelial cells.
-This suggests that stromal cells are less specified to each organ.
-A discussion of the accuracy of scTOP, when applied to stromal types versus parenchymal types, may be found in appendix \ref{failure_cases}.
-
-\begin{figure}
-	\centering
-    \includegraphics[scale=0.5]{figs/tabula muris heatmap.pdf}
-	\caption{scTOP is able to classify tissues at different resolutions.
-Aggregate pseudo-bulk samples from the Mouse Cell Atlas are compared to a reference basis consisting of organs from Tabula Muris.
-The scTOP scores are significantly higher when comparing a cell type with the organ of origin.
-The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis.
-The y-axis lists the true types of the queried cells.
-The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types.}
-	\label{tabula_muris}
-\end{figure}
-
-\subsubsection{Example 4:  Human data}
-
-The CellBench dataset \cite{tian_benchmarking_2019} was designed specifically to benchmark scRNA-seq analysis methods.
-We used the subset of data pertaining to five cancer lines.
-200 cells from each line were used as training data to create the reference basis, then the rest of the cells were input as query data.
-As shown in figure \ref{cellbench} and table \ref{table:1}, scTOP classifies the cell line identity of the test data with 100 percent accuracy.
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.5]{figs/cellbench heatmap.pdf} 
-	\caption{scTOP correctly identifies human tissues.
-The CellBench data set contains human cell cancer lines specifically processed to benchmark scRNA-seq algorithms.
-We take a subset of cells to use as the reference basis in analyzing the rest of the CellBench data.
-The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis.
-The y-axis lists the true types of the queried cells.
-The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types.}
-	\label{cellbench}
-\end{figure}
-
-To illustrate the power of scTOP for classifying human samples, we analyzed data from the human lung atlas dataset \cite{travaglini_molecular_2020}.
-This dataset sequenced thousands of human lung cells and identified 58 phenotypic populations.
-For our analysis, we restricted the training and test data to epithelial and stromal types that had at least 200 cells in each type population.
-The reference basis was created using 80\% of the cells from each population, and the query data consisted of the remaining 20\% of the data.
-Figure \ref{human lung} shows that the scTOP scores were consistently high in cases where the score type matched the true type of the query.
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.5]{figs/human lung heatmap.pdf}
-	\caption{scTOP correctly identifies human tissues.
-Similar to figure \ref{cellbench}, cells from the Human Lung Atlas are compared to a reference basis constructed from cells from the Human Lung Atlas.
-The colors of the bins correspond to the average scTOP scores of individual cells.
-Again, the diagonal is the most prominent section, showing that cells of the y-axis types are correctly matched to the reference x-axis identities.}
-	\label{human lung}
-\end{figure}
-
-As demonstrated by the diverse data sources examined here, as long as a well-defined reference basis is available, scTOP works extremely well for cell type identification across different measurement conditions, species, organs, and resolutions.
-This performance is especially impressive since, as discussed earlier, the algorithm does not use any statistical fitting procedures or dimensional reduction methods such as PCA, tSNE, UMAP, or SPRING.
-
-\subsection{Visualizing developmental dynamics using scTOP }
-Having demonstrated the robust ability of scTOP to identify stable cell types, we next demonstrate how scTOP can be used in combination with scRNA-seq time series to visualize transient cell states in dynamic processes like development and differentiation.
-
-\subsubsection{Lung development}
-The alveolar epithelium of the lung undergoes dramatic maturation and morphological changes during embryonic and perinatal development, ultimately giving rise to AT1 cells and AT2 cells.
-To investigate this process with scTOP, we reanalyzed alveolar epithelial cells from a recent dataset by Zepp et al.
-\cite{zepp2021genomic}, which includes seven time points between embryonic day 12.5 (E12.5) and postnatal day 42 (P42).
-These cells were then compared against a reference basis developed from adult cell types in the Mouse Cell Atlas and an E12.5 epithelial progenitor cell type from the Atlas of Lung Development  \cite{negretti_single-cell_2021}.
-
-The resulting AT1 and AT2 scTOP scores were plotted to visualize maturation into these two lineages (Figure \ref{LungMAP}(a)).
-As expected, early embryonic progenitor cells (E12.5, E15.5), which express few of the mature lineage markers, align poorly with both lineages.
-In contrast, the E17.5 alveolar epithelial cells fall along a wide spectrum with cells aligning to AT1 cells, AT2 cells, or both lineages.
-In the postnatal time points, this continuous spectrum gives way to more distinct clusters, eventually resolving into AT2 and AT1 cell clusters at P42.
-Similar to our analysis of adult mouse lungs (Figure \ref{herriges}), the scTOP scores seen over development correlate well with our lineage marker gene sets (Figure \ref{LungMAP} (b)-(c)).
-
-In addition to the expected postnatal AT2 and AT1 cell clusters, this scTOP analysis illustrated a transient AT2/AT1 hybrid population that persisted through P7.
-The existence of these hybrid AT2/AT1 cells was already noticed in Zepp et al.
-using a combination of graph-based clustering and gene enrichment profiling (see Figure 1C, I in \cite{zepp2021genomic}).
-The authors conjectured that these AT2/AT1 cells were a transitional state of AT2 similar to Spock2+/Axin2+ AT2 cells previously identified in adult lungs \cite{frank2016emergence}.
-To assess whether this was the case, we displayed Axin2 expression on our scTOP plots (Figure \ref{LungMAP genes}(a)).
-These cells do not have significant upregulation of Axin2 relative to perinatal or adult AT2 cells, suggesting that the postnatal hybrid AT2/AT1 cells are distinct from adult Spock2+/Axin2+ AT2 cells \cite{frank2016emergence}.
-
-We then performed a similar analysis of markers for other published postnatal hybrid AT1/AT2 or transitional alveolar cell states, including alveolar differentiation intermediate (ADI, {\it Krt8} and {\it Trp53}), pre-alveolar type-1 transitional cell state (PATS, {\it Cldn4} and {\it Krt19}), and damage-associated transient progenitors (DATPs, {\it Ilr1} and {\it Hif1a}) \cite{verheyden2020transitional, strunz2020alveolar, kobayashi2020persistence, choi2020inflammatory}.
-In each case, markers were not uniquely upregulated in the perinatal hybrid AT1/AT2 cells relative to either the perinatal or adult committed lineages (Figure \ref{LungMAP genes} (b) - (d)).
-Furthermore, differential gene expression analysis failed to identify any genes that were uniquely differentially expressed in the hybrid AT1/AT2 cells relative to either timepoint-matched AT1 or AT2 cells.
-These results suggest that this population is most likely transcriptionally distinct from previously identified adult AT2-transitional cell types.
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.52]{figs/LungMAP.png}
-	\caption{Visualizing the developing mouse lung with scTOP.
-scTOP illustrates the specification of alveolar type I and type II cells in murine embryonic development.
-Alveolar cells from Zepp et al.
-are compared to a reference basis constructed from adult lung cells from Herriges et al.
-and early epithelial cells from the Atlas of Lung Development.
-(a) Each subplot corresponds to the age of the mouse from which the sample was extracted, from post-conception day 12,5 (E12.5) to 42 days post-birth (P42).
-(b), (c) Scatter plots of the same data as in (a), now with each cell colored by the average over that cell's expression of (b) AT1 marker genes and (c) AT2 marker genes.
-The scTOP scores of AT1 (AT2) cells are high when the average AT1 (AT2) marker gene expression is high.
-(d) The same data as above is displayed as a 2D histogram, with each bin colored corresponding to the average day of the data point(s) falling within that bin.
-The color trend shows that early cells generally have very low AT1 and AT2 scores, then the AT1 and AT2 scores increase over time.}
-	\label{LungMAP}
-\end{figure}
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.8]{figs/LungMAP genes.png}
-	\caption{The AT1/AT2 combination state (most apparent in days P3 and P7) do not uniquely express genes of AT1/AT2 transitional states described by previous papers.
-Alveolar cells from Zepp et al.
-are plotted on AT1 and AT2 axes, separated by day.
-Each cell is colored by the expression z-score of a particular gene, and the particular gene is different in each row.
-The gene used to color the cells is indicated by the color bar on the far right of each row.
-(a) Alveolar cells colored by Axin2, showing that the AT1/AT2 combination cells do not appear to belong to the Axin2+ sublineage described by Frank et al.
-\cite{frank2016emergence}.
-(b) Alveolar cells colored by Krt8 and Trp53 expression, which are genes corresponding to alveolar differentiation intermediate cells (ADI) described by Strunz et al.
-\cite{strunz2020alveolar}.
-AT1, AT2, and the combination AT1/AT2 state all express Krt8 and Trp53 at similar levels.
-(c) Alveolar cells colored by expression of pre-alveolar type-1 transitional cell state (PATS) genes Kobayashi et al.
-\cite{kobayashi2020persistence}.
-AT1, AT2, and the combination AT1/AT2 state all express Cldn4 at similar levels.
-AT1 and AT1/AT2 combination cells both express Krt19 at similar levels, although AT1 cells express the gene at higher levels.
-(d) Alveolar cells colored by expression of damage-associated transient progenitor (DATP) genes described by Choi et al.
-\cite{choi2020inflammatory}.
-AT1, AT2, and the combination AT1/AT2 state all express Il1r1 and HIF1a at similar levels.
-}
-	\label{LungMAP genes}
-\end{figure}
-
-\subsubsection{scTOP for hematopoiesis using lineage tracing data}
-
-In the previous section, we used scTOP to visualize differentiation into two closely related lung epithelial cell fates.
-In this section, we show how scTOP can also be used to gain insights into more complicated developmental processes such as hematopoiesis, in which hematopoietic stem cells (HSCs) give rise to multiple differentiated blood cell types.
-To do so, we make use of the unique dataset by Weinreb et al.
-\cite{weinreb_lineage_2020} on hematopoiesis in mice using a new technique, lineage and RNA recovery (LARRY), that combines scRNA-seq and lineage tracing data.
-LARRY uses heritable barcodes that can be detected using scRNA sequencing, allowing for the tracking of clone families during differentiation.
-Here, we focus on visualizing the ex-vivo experiments in \cite{weinreb_lineage_2020}, where cells were sequenced at a sufficient depth to infer lineage relationships.
-In these experiments, HSCs were extracted from mice, barcoded, and allowed to expand in primary culture.
-scRNA-seq measurements were performed on days 2, 4, and 6 post-barcoding.
-The cells were then annotated using marker genes into 9 differentiated blood cell fates.
-
-We restrict our analysis to a subset of these annotated cell types with at least 150 cells.
-We also excluded cells labeled as erythroids from our analysis since we found that this population was highly heterogeneous and gave rise to poor projections (likely due to the fact it was annotated using a single marker gene, \textit{Hbb-bs} \cite{weinreb_lineage_2020}, which did not sufficiently define a distinct population).
-Furthermore, we treated all day 2 undifferentiated cells as a single population we called progenitors.
-This resulted in a reference basis consisting of seven cell types: undifferentiated progenitors, neutrophils, monocytes, megakaryocytes, mast cells, eosinophils, and basophils.
-To construct the basis, we averaged the RNA counts for 150 random cells of each of these types, only including cells if they had no sister clones.
-Then we preprocessed the averaged gene expression profiles and inserted them into the reference basis.
-
-Figure \ref{hem clones} shows the differentiation dynamics of three different clonal families using scTOP.
-The histograms show the distribution of scTOP projection scores on each cell type for days 2, 4, and 6 post-barcoding, while the scatter plots show two-dimensional cross-sections of scTOP projection scores as a function of time for the indicated cell types.
-In all the plots, the scTOP projections for the undifferentiated hematopoietic progenitor stem cell decreased over time, with a high projection on day 2 and a low projection on day 6, while the scores for the mature types increased in time.
-
-Interestingly, the three clones show very different behaviors as they mature.
-Figure \ref{hem clones} (a) shows a clone where a single barcode detected at day 2 gives rise to multiple cells in the neutrophil lineage.
-This can be seen in the plots by noting that the day 6 cells have near zero scores on both progenitor cells and other differentiated lineages, such as monocytes, but project significantly on the neutrophils.
-Figure \ref{hem clones} (b) shows a barcode that was detected in three cells on day 2 that gives rise to a mix of neutrophils and monocytes on day 6.
-Interestingly, the day 4 cells move in the direction of both of these lineages before bifurcating at day 6.
-This data is consistent with Weinreb et al.
-\cite{weinreb_lineage_2020}, who previously highlighted the existence of such a bipotent neutrophil-monocyte progenitor population.
-We also find evidence that progenitor cells can even give rise to three distinct cellular populations.
-Figure \ref{hem clones} (c) shows a single barcode found on day 2 that results in day 6 cells with three distinct populations with non-zero scores on basophils, eosinophils, and neutrophils, suggesting a single clone can bifurcate into three different types of differentiated cell types.
-
-Figure \ref{hem clones} focuses on scTOP scores for individual clonal families, while Figure \ref{hem progenitor} shows the scTOP projection scores across all 1,816 multi-generational clonal families in the dataset.
-In each subplot, the horizontal axes show progenitor scores, and the vertical axes correspond to the six differentiated cell types included in the reference basis.
-Each bin of the 2-dimensional histogram is colored according to the {\it average} day of all the cells that fall within that bin.
-For this reason, although the cells were only sampled on days 2, 4, and 6, the color bar is continuous, with a range between 2 and 6 days.
-The histograms show that, as expected, the day 2 cells generally have high projections on progenitor cells, but as time passes, progenitor scores decrease while the scores on differentiated cell types either increase (indicating cells that differentiate into the cell type on the y-axis) or remain close to zero (indicating cells that differentiate into a different cell type than that on the y-axis).
-Somewhat surprisingly, the plots for different cell types look qualitatively different from each other (for example, compare the plots for megakaryocytes and monocytes).
-
-To try to better understand this, we made additional 2-dimensional histograms comparing scTOP scores between all differentiated cell types in our basis (see Figure \ref{hem mature}).
-What is striking about these plots is that they show two qualitatively different behaviors depending on which pairs of cells are being compared: the differentiation pathways for some pairs are mutually exclusive, while other pairs show a more graded behavior.
-For example, the histogram for megakaryocytes and monocytes exhibits a distinct L-shape, indicating that these developmental programs are mutually exclusive since cells never have significant projection scores of both cell types simultaneously.
-This is in stark contrast with the histogram for monocytes and neutrophils, where there exists a continuous spectrum of cells that are monocyte-like, neutrophil-like, and every step in between.
-
-Our scTOP-based visualizations of developmental decisions during hematopoiesis complement the analysis in \cite{weinreb_lineage_2020} using pseudo-time.
-Like in the original paper, we find strong evidence for the existence of progenitor cells that can give rise to multiple lineages (Figure \ref{hem clones}), including a neutrophil/monocyte bivalent progenitor and possibly a basophil/eosinophil/neutrophils trivalent progenitor.
-The existence of these multi-potent progenitors seems to give rise to graded developmental dynamics where intermediate cells have significant projections on multiple cell types.
-This is in stark contrast with developmental decisions between other cell lineages (megakaryocytes vs.
-monocytes, basophils vs.
-megakaryocytes, monocytes vs.
-basophils) that seem to be mutually exclusive.
-One compelling feature of these visualizations using scTOP is that they require no statistical fitting, dimensional reduction, or ordering of cells.
-Once we have chosen the relevant reference basis (in this case, progenitors and six differentiated lineages), the resulting plots give an interpretable way of assessing complex developmental dynamics.
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.75]{figs/hem. clone families.png}
-	\caption{Tracing hematopoietic cell differentiation with scTOP.
-(a), (b), (c) display scatter plots and distributions of scTOP scores for individual in vitro clone families from  \cite{weinreb_lineage_2020}.
-The scatter plots show the x-axis scTOP score compared to the y-axis scTOP score for individual cells, while the distributions show the scTOP scores of the type indicated on the x-axis of the corresponding column.
-The color and shape of each marker (and the color of each distribution) indicate the days in culture on which the cells were extracted.
-(a) A clone family where all of the cells end up becoming neutrophils.
-(b) A clone family where some clones end up becoming neutrophils and some become monocytes.
-At day 2, the 3 sister cells have low neutrophil and monocyte scores and high progenitor scores.
-At days 4 and 6, the progenitor scores decrease, and a portion of the cells increase in monocyte scores while others increase in neutrophil scores.
-(c) A clone family where the descendant cells become neutrophils, eosinophils, and basophils.}
-	\label{hem clones}
-\end{figure}
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.75]{figs/hem progenitor hists.png}
-	\caption{Tracing differentiation of hematopoietic cells from progenitor to mature type.
-The plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al., with the x-axis of each one indicating the scTOP score for hematopoietic progenitor type and the y-axis showing the score for a mature type.
-Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin.
-Early cells have high progenitor scores and low mature-type scores.
-As time passes, cells end up with high scores in mature types.}
-	\label{hem progenitor}
-\end{figure}
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.59]{figs/hem mature hists.png}
-	\caption{Comparing differentiation of hematopoietic cells between mature types.
-As in figure \ref{hem progenitor}, the plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al.
-The x-axis and y-axis of each histogram indicate the scTOP score of the labeled mature type.
-Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin.  Some pairs of types have cells with high scores of both types.
-This is apparent in the neutrophil-monocyte histogram, which presents a continuous spectrum of scores from the horizontal axis to the vertical.
-Other pairs of types only have cells expressing one type or the other, as in the L-shaped neutrophil-megakaryocyte histogram.}
-	\label{hem mature}
-\end{figure}
-
-\subsection{Using scTOP to compare endogenous and transplanted cells}
-
-\begin{figure}[h!]
-	\centering
-		\includegraphics[scale=0.83]{figs/engineered cells comparison.png}[h!]
-	\caption{Donor-derived and endogenous alveolar cells from Herriges et al.
-and Louie et al.
-The cell type annotations come from a separate Louvain clustering and are the same for all plots in this figure.
-(a) UMAP visualizations of the data, colored by cell type annotations.
-The three have different values for the UMAP parameter n\_neighbors, which changes whether the embedding preserves more local or global structure.
-(b) scTOP plot showing the AT1 score on the x-axis and the AT2 score on the y-axis.
-(c) Box-and-whisker plot showing the distributions of scTOP AT2 scores for the various cell populations.
-(d) Bar plot showing the top ten aggregate scores for the AT2 populations from different sources, sorted by the scores for the SCA1+ population.}
-	\label{engineered}
-\end{figure}
-Recent advancements in cell culture and cell transplantation techniques have allowed researchers to generate cells that were maintained or differentiated in vitro to approximate and eventually replace endogenous cells generated during in vivo development.
-However, even with scRNA-seq, it is difficult to consistently quantify the transcriptomic similarity between donor-derived and endogenous lineages from transplant recipients.
-Since it provides a quantitative measure of cell type similarity, scTOP can be useful for comparing cell populations.
-A relevant example can be found in pulmonary cell replacement therapy, where one goal is to differentiate or maintain alveolar cells in vitro and then transplant them into injured lungs.
-
-To compare two distinct pulmonary cell transplant protocols, we used scTOP to analyze the results of scRNA-seq from two recent publications featuring murine pulmonary cell transplantation \cite{herriges_durable_2022, louie_progenitor_2022}.
-In Louie et al., adult mouse AT2 cells were maintained in culture with support cells for 3 weeks to generate a donor cell population.
-In contrast, in Herriges et al.
-the authors developed a protocol for directed differentiation of mouse PSCs into lung distal tip-like cells, which mimic an embryonic progenitor of AT2 cells.
-In both papers, the donor cells were transplanted into injured mouse lungs, where they survived and gave rise to multiple cell lineages, including AT2-like cells.
-
-To highlight the usefulness of scTOP, we first consider another dimensional reduction method.
-Dimensional reduction methods like UMAP and t-SNE are often used to compare similar cells from different sources, such as transplanted and endogenous cells.
-Figure \ref{engineered} shows the AT1 and AT2 cell populations from Herriges et al.
-and Louie et al., both donor-derived and endogenous.
-In figure \ref{engineered} (a), three possible UMAP plots are shown with varying algorithmic parameters.
-The qualitative relations between the clusters are generally consistent, with AT1 cells clustering together apart from the AT2 cells.
-However, the distances between cells and the relative positions of clusters are not consistent when the UMAP parameter is varied.
-This illustrates how UMAP is useful for visualization and gaining qualitative intuition for a particular dataset but cannot be used to make a rigorous quantitative comparison between cells.
-
-On the other hand, figure \ref{engineered} (b) shows the scTOP scores for AT1 and AT2 for the same populations.
-The distances between cells will only change if the reference basis is changed, and even then will not change significantly (see \ref{robustness hists} for discussion).
-The AT1 and AT2 populations are separated along the relevant axes, and the highly-similar cell populations are largely overlapping.
-Because scTOP provides many relevant dimensions to choose from instead of trying to capture high-dimensional distances in two dimensions, it is possible to directly compare different populations to the same cell type.
-Figure \ref{engineered} (c) shows the distributions of AT2 scores for each of the populations, providing a clear quantitative way to compare the various engineered cells.
-As expected, the endogenous populations from both data sources are similar in distribution.
-The means and overall distributions of the AT2-derived (Louie et al.) cells are higher than those for the PSC-derived (Herriges et al.) cells, suggesting that the primary cell transplantation better emulates the endogenous AT2 cells.
-
-Figure \ref{engineered} (d) provides a comparison between cell populations using aggregate scTOP scores.
-Similar to figure \ref{engineered} (c), (d) provides a quantitative comparison between the populations.
-This figure confirms that while both AT2-derived and PSC-derived transplants generate cells that have high AT2 scores, the AT2-derived cells are quantitatively more similar to the reference AT2 profile.
-Altogether this analysis demonstrates how scTOP can be used to compare two distinct cell populations against a primary control.
-scTOP satisfies the need for quantitative assessment of transplantation protocols and can be used in combination with functional tests to identify protocols that most effectively replace endogenous cell lineages.
-
-\section{Discussion}
-
-In this paper, we have introduced a new physics-inspired method, scTOP, for analyzing scRNA-seq data.
-scTOP projection scores provide clear and meaningful axes to visualize differentiation and classify cells.
-Since scTOP requires no statistical fitting, clustering, or dimensional reduction techniques, plots can be made quickly and easily, even for an individual cell.
-The input to scTOP is a reference basis of cell fates of interest, allowing us to take advantage of the wealth of new scRNA-seq expression atlases now being generated.
-Importantly, scTOP scores are robust to the choice of basis and reproducible across experiments, labs, and datasets, allowing integration of disparate data sources.
-
-scTOP is especially suited for tasks where the cell types of interests are well characterized.
-For example, we have found in our own work that scTOP is an extremely sensitive measure for assessing the fidelity of engineered cells from directed differentiation protocols and reprogramming \cite{herriges_durable_2022}.
-Since scTOP does not require data harmonization, joint embeddings, or make use of cellular neighborhood information,  scTOP is able to distinguish between technical noise and biologically meaningful differences.
-scTOP also can be used to classify cells with minimal computational costs and thus represents a potentially simple, scalable, and reproducible way of dealing with the proliferation of scRNA-seq datasets.
-Since scTOP yields an alternative coordinate system for representing developmental dynamics in cell-fate space, scTOP-based visualizations are a powerful way of thinking about complex developmental processes.
-
-As an illustration, we analyzed the development of the lung epithelial AT1 and AT2 cell fates \cite{zepp2021genomic} and developmental dynamics during hematopoiesis  \cite{weinreb_lineage_2020}.
-Using scTOP, we showed that it was easy to identify a transient hybrid AT2/AT1 population 3 to 5 days post-birth whose gene expression profile is distinct from previously investigated transient AT2-to-AT1 states.
-By plotting scTOP scores as a function of time for AT1 and AT2, hybrid AT2/AT1 cells were easily identified since they formed a distinct cluster that was well separated from mature AT1 and AT2 cells.
-This example illustrates how given an accurate reference basis, scTOP is able to pick up on small but biologically meaningful differences in global gene expression profiles.
-
-In the context of hematopoietic development, where HSCs give rise to multiple downstream cell fates, we took advantage of the fact that scTOP maps each cell to a seven-dimensional coordinate (each direction in the coordinate systems corresponds to one of the seven cell types in the reference basis: undifferentiated progenitors, neutrophils, monocytes, megakaryocytes, mast cells, eosinophils, and basophils) to generate multiple two-dimensional visualizations using scTOP.
-We found that a single clonal family can often give rise to up to three different mature cell fates.
-Our visualizations also suggest that developmental decisions between different cell fates can be classified into two broad categories: mutually exclusive and graded.
-For cell fates such as megakaryocytes and monocytes, cells never have significant projection scores on both lineages simultaneously, suggesting that the underlying developmental decisions are mutually exclusive.
-In contrast, for other pairs of cell fates, such as monocytes and neutrophils, cells often exhibit significant projections on both members of the pair suggesting these developmental switches function in a more graded manner.
-It will be interesting to see if this general distinction is also present in other developmental systems or is specific to hematopoiesis.
-
-In the analysis here, we have limited ourselves to considering scRNA-seq data.
-However, in principle, our approach can be easily extended to include other data modalities, such as chromatic accessibility and histone modifications.
-Recently, several interesting works such as Dynamo have used RNA-velocity to directly learn cellular dynamics \cite{xing2022reconstructing, qiu2022mapping}.
-It will be interesting to see if these methods can be combined with scTOP to learn vector flows directly in cell-type space.
-Similarly, methods based on optimal transport, such as Waddington-OT \cite{schiebinger2019optimal}, or developmental directory reconstruction based on graph embedding, such as Monocle \cite{qiu2017reversed}, can also be performed in the cellular identity space instead of gene expression space.
-These represent promising directions for future research since the scTOP cell identity coordinates are often better suited for analyzing developmental dynamics than gene expression coordinates.
-
-We have found that scTOP can be used to quantitatively assess the fidelity of cultured cells, as in the case of donor-derived lung alveolar cells.
-However, we currently have no systematic manner within the scTOP framework for providing guidance on adjusting gene expression or designing differentiation protocols to steer cells toward desired fates.
-Developing such methods would provide powerful new computational tools for engineering cells via directed differentiation.
-
-We are also interested in extending scTOP to better understand the signaling pathways and genetic signatures underlying developmental dynamics.
-While scTOP provides a way to translate between gene expression space and cell type space, there is currently no direct interface with the biologically-essential signaling space.
-Further work in uniting the three relevant spaces, combined with studying the possible bifurcations in these spaces, may provide insight into the complex process of differentiation.
-
-\section*{Acknowledgments}
-We acknowledge useful discussions with Jason Rocks, Robert Marsland, and Alex Lang.
-We would like to thank the Mehta group and Kotton lab for their comments on the manuscript.
-The work was funded by a grant to the Boston University Kilachand Multicellular Design Program (to DK and PM) and NIH NIGMS 1R35GM119461 to PM.
-We also acknowledge programming support from Alena Yampolskaya and Sumner Hearth.
-
-\section*{Appendix}
-\beginsupplement
 \subsection{Preprocessing} \label{preprocessing}
@@
 	\label{FIG:supp1}
+\end{figure}
+
+\subsubsection{Preprocessing rationale}
+
+In order to mitigate batch effects, the preprocessing step in scTOP takes each cell and ranks each gene according to expression.
+These ranks are converted into percentiles, which are then transformed into z-scores assuming a normal distribution.
+
+By ranking the genes before converting to z-scores, we are ignoring the absolute values of RNA counts and instead focusing on the relative values.
+Some sequencing platforms may measure many more counts than others; although we can't recover the unmeasured counts that are lost via dropouts, we can remove the effect of differences in counts by ranking.
+We assume that each cell contains similar numbers of mRNA and that the actual number of RNA doesn't matter as much as the relative expression of different genes.
+For example, the highest-expressed gene in one cell should be comparable to the highest-expressed gene in another cell, regardless of RNA count.
+Ignoring the absolute values of RNA counts reduces the differences caused by different scRNA-seq technologies and experiments, as shown in section \ref{batch effects section}.
+
+The second preprocessing step is to convert the ranks to z-scores, assuming a normal distribution.
+This step results in a mean of zero and standard deviation of one.
+In data analysis, it is beneficial to standardize the data before comparing across datasets because this prevents a disproportionate bias towards features with large values.
+In the case of scTOP, having a mean of zero and a standard deviation of one is especially important because the underlying math relies on the dot product of two uncorrelated vectors to be zero and the magnitude of a vector to be one.
+
+By ranking and z-scoring, we remove undue weighting of genes according to magnitude of expression, instead prioritizing the relative expression.
+
+\subsection{Batch effects} \label{batch effects section}
+scRNA-seq data produced using different sequencing platforms from different experiments may introduce variation in the form of batch effects.
+Because the pre-processing step in scTOP prioritizes the relative rather than absolute expression of genes, batch effects are mitigated.
+
+Tran et al.
+\cite{tran2020benchmark} consolidated datasets from different experiments for the purposes of benchmarking various batch-effect correction methods.
+One of these is a dataset containing scRNA-seq human pancreatic data from five different sources \cite{baron2016single} \cite{muraro2016single} \cite{segerstolpe2016single} \cite{wang2016single} \cite{xin2016rna}, using various sequencing technologies.
+We used scTOP to analyze these datasets.
+Figure \ref{batch effects} illustrates that scTOP is not significantly affected by differences between batches.
+Because the Baron et al.
+dataset had the most cells per cell type, we used it to create a reference basis consisting of alpha, beta, ductal, acinar, delta, gamma, stellate, and endothelial gene expression profiles.
+We projected cells from the other four experiments onto the Baron reference basis.
+The top row shows kernel density estimate (KDE) plots for cells that match the reference cell type (e.g.
+alpha scores for alpha cells) with solid lines, and cells that do not match the reference type but have higher scores than the other non-matching types (e.g.
+alpha scores for beta cells) with dashed lines.
+The bottom row shows box-and-whisker plots for the cells that match the reference type.
+
+Despite coming from different sources, the distributions of scores are similar.
+There is also a clear bimodal distribution for cells that match the reference type and cells that don't.
+This bimodality is a clear indication that batch effects don't cause misidentification.
+
+According to Ziegenhain et al.
+\cite{ziegenhain2017comparative}, Smart-seq2 detects the most genes per cell when compared to other sequencing methods of that time.
+This decrease in dropouts may explain why the Segerstolpe data (which is the only experiment that used Smart-seq2) has slightly higher projections than the other datasets.
+
+\begin{figure}
+	\centering
+    \includegraphics[scale=0.95]{figs/batch effects.pdf}
+	\caption{Pancreatic data from different experiments using different scRNA-seq technologies.
+The reference basis uses data from Baron et al.
+\cite{baron2016single}, and the cells that are projected onto this basis are from Muraro et al.
+\cite{muraro2016single}, Segerstolpe et al.
+\cite{segerstolpe2016single}, Wang et al.
+\cite{wang2016single}, and Xin et al.
+\cite{xin2016rna} (A) Kernel density estimate (KDE) plot of alpha scTOP scores for alpha cells (solid) and beta cells (dashed), and a box-and-whisker plot of alpha scTOP scores for alpha cells.
+(B) KDE plot of beta scTOP scores for beta cells (solid) and delta cells (dashed), and a box-and-whisker plot of beta scTOP scores for beta cells.
+}
+	\label{batch effects}
 \end{figure}
@@
 This provides a useful metric for determining whether a query cell belongs to one of the cell types in the reference basis.
-If the highest scTOP score is lower than 0.1 for a query cell, this indicates that the reference basis lacks the relevant types or was poorly constructed (see appendix \ref{failure_cases} for a discussion of ill-defined bases).
+If the highest scTOP score is lower than 0.1 for a query cell, this indicates that the reference basis lacks the relevant types or was poorly constructed (see section \ref{failure_cases} for a discussion of ill-defined bases).
 
@@
 The other part of the assumption is that the reference types should be accurate profiles of the cell type of interest.
-As discussed in appendix \ref{basis construction}, reference types are created from averaging over cells, and the more cells included in this average, the better-sampled the type is.
+As discussed in section \ref{basis construction}, reference types are created from averaging over cells, and the more cells included in this average, the better-sampled the type is.
 As expected, scTOP gives unreliable results when the reference gene expression levels do not accurately reflect the true cell types' gene expression levels.
@@
 
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.8]{figs/LungMAP genes.png}
+	\caption{The AT1/AT2 combination state (most apparent in days P3 and P7) do not uniquely express genes of AT1/AT2 transitional states described by previous papers.
+Alveolar cells from Zepp et al.
+are plotted on AT1 and AT2 axes, separated by day.
+Each cell is colored by the expression z-score of a particular gene, and the particular gene is different in each row.
+The gene used to color the cells is indicated by the color bar on the far right of each row.
+(a) Alveolar cells colored by Axin2, showing that the AT1/AT2 combination cells do not appear to belong to the Axin2+ sublineage described by  \cite{frank2016emergence}.
+(b) Alveolar cells colored by Krt8 and Trp53 expression, which are genes corresponding to alveolar differentiation intermediate cells (ADI) described by \cite{strunz2020alveolar}.
+AT1, AT2, and the combination AT1/AT2 state all express Krt8 and Trp53 at similar levels.
+(c) Alveolar cells colored by expression of pre-alveolar type-1 transitional cell state (PATS) genes described by \cite{kobayashi2020persistence}.
+AT1, AT2, and the combination AT1/AT2 state all express Cldn4 at similar levels.
+AT1 and AT1/AT2 combination cells both express Krt19 at similar levels, although AT1 cells express the gene at higher levels.
+(d) Alveolar cells colored by expression of damage-associated transient progenitor (DATP) genes described by \cite{choi2020inflammatory}.
+AT1, AT2, and the combination AT1/AT2 state all express Il1r1 and HIF1a at similar levels.
+}
+	\label{LungMAP genes}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.4]{figs/lung barplot.pdf}
+	\caption{Bar plot showing the top ten aggregate scores for the AT2 populations from the various endogenous and transplanted sources shown in Figure 9 of the main text.
+The cell types are ranked according to the magnitude of scores for the SCA1+ population.}
+    \label{barplot}
+\end{figure}
+
 \bibliography{export-data.bib}

### 2023-07-21 manuscript/scTOP (Company of Biologists LaTeX template)/scTOP.tex
added revised manuscript

--- 2023-01-25 manuscript/scTOPv5.tex
@@
-	\documentclass[aps,superscriptaddress, notitlepage,longbibliography]{revtex4-1}
+\documentclass[vruler,JEB]{COB}%%%%where COB is the template name
+
+% \usepackage[nomarkers,figuresonly]{endfloat}
 \usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{physics}
-\usepackage[pdftex]{graphicx}
-\usepackage{xcolor}
-\usepackage{hyperref}
-\setlength {\marginparwidth }{2cm}
+% \usepackage{amsmath}
+% \usepackage{physics}
+\bibliographystyle{plainnat}
+% \usepackage[pdftex]{graphicx}
+% \usepackage{xcolor}
 \hypersetup{
@@
 }
-
+\usepackage{xr}
 \newcommand{\comment}[1]{{\color{blue}#1}}
@@
      }
-% Standardize command font styles and environments
-\newcommand{\doccmd}[1]{\texttt{\textbackslash#1}}% command name -- adds backslash automatically
-\newcommand{\docopt}[1]{\ensuremath{\langle}\textrm{\textit{#1}}\ensuremath{\rangle}}% optional command argument
-\newcommand{\docarg}[1]{\textrm{\textit{#1}}}% (required) command argument
-\newcommand{\docenv}[1]{\textsf{#1}}% environment name
-\newcommand{\docpkg}[1]{\texttt{#1}}% package name
-\newcommand{\doccls}[1]{\texttt{#1}}% document class name
-\newcommand{\docclsopt}[1]{\texttt{#1}}% document class option name
-\newenvironment{docspec}{\begin{quote}\noindent}{\end{quote}}% command specification environment
+
+\makeatletter
+\newcommand*{\addFileDependency}[1]{% argument=file name and extension
+\typeout{(#1)}% latexmk will find this if $recorder=0
+% however, in that case, it will ignore #1 if it is a .aux or 
+% .pdf file etc and it exists! If it doesn't exist, it will appear 
+% in the list of dependents regardless)
+%
+% Write the following if you want it to appear in \listfiles 
+% --- although not really necessary and latexmk doesn't use this
+%
+\@addtofilelist{#1}
+%
+% latexmk will find this message if #1 doesn't exist (yet)
+\IfFileExists{#1}{}{\typeout{No file #1.}}
+}\makeatother
+
+\newcommand*{\myexternaldocument}[1]{%
+\externaldocument{#1}%
+\addFileDependency{#1.tex}%
+\addFileDependency{#1.aux}%
+}
+
+\myexternaldocument{SI}
 
@@
 \newcommand{\mbbE} {\mathbb{E}}
+\usepackage{widetext-TI}
+
+\newtheorem{theorem}{Theorem}
+\newtheorem{condition}{Condition}
+
+%The author can find the documentation of the above style file and any additional
+%supporting files if required from "http://www.ctan.org"
+
+% *** Do not adjust lengths that control margins, column widths, etc. ***
+
 \begin{document}
 
-\title{ scTOP: physics-inspired order parameters for cellular identification and visualization}
+\supertitle{Research Article}
 
-\author{Maria Yampolskaya}
-\email{mariay@bu.edu}
-\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
-\author{Michael Herriges}
-\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
-\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
-\author{Laertis Ikonomou}
-\affiliation{Department of Oral Biology. University at Buffalo, The State University of New York, Buffalo, NY, USA}
-\affiliation{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
-\author{Darrell Kotton}
-\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
-\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
-\author{Pankaj Mehta}
-\email{pankajm@bu.edu}
-\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
-\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
-\affiliation{Faculty of Computing and Data Science, Boston University, Boston, MA 02215, USA}
-\affiliation{Biological Design Center, Boston University, Boston, MA 02215, USA}
+\title{scTOP: physics-inspired order parameters for cellular identification and visualization}
+
+\author{Maria Yampolskaya$^{1}$, Michael J Herriges$^{2, 3}$, Laertis Ikonomou$^{4, 5}$, Darrell N Kotton$^{2, 3}$, and Pankaj Mehta$^{1, 2, 6, 7}$}
+
+\address{\add{1}{Department of Physics, Boston University, Boston, MA 02215, USA}
+\add{2}{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
+\add{3}{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
+\add{4}{Department of Oral Biology. University at Buffalo, The State University of New York, Buffalo, NY, USA}
+\add{5}{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
+\add{6}{Faculty of Computing and Data Science, Boston University, Boston, MA 02215, USA}
+\add{7}{Biological Design Center, Boston University, Boston, MA 02215, USA}}
+
+\corres{(\email{mariay@bu.edu}; \email{pankajm@bu.edu})}
+
+\maketitle
 
@@
 Advances in single-cell RNA-sequencing (scRNA-seq) provide an unprecedented window into cellular identity.
-The increasing abundance of data requires new theoretical and computational frameworks for understanding cell fate determination, accurately classifying cell fates from expression data, and integrating knowledge from cell atlases.
-Here, we present single-cell Type Order Parameters (scTOP): a statistical-physics-inspired approach for constructing ``order parameters'' for cell fate given a reference basis of cell types.
-scTOP can quickly and accurately classify cells at a single-cell resolution, generate interpretable visualizations of developmental trajectories, and assess the fidelity of engineered cells.
-Importantly, scTOP does this without using feature selection, statistical fitting, or dimensional reduction (e.g., UMAP, PCA, etc.).
-We illustrate the power of scTOP utilizing a wide variety of human and mouse datasets (both {\it in vivo} and {\it in vitro}).
-By reanalyzing mouse lung alveolar development data, we characterize a transient perinatal hybrid alveolar type 1/alveolar type 2 (AT1/AT2) cell population that disappears by 15 days post-birth and show that it is transcriptionally distinct from previously identified adult AT2-to-AT1 transitional cell types.
-Visualizations of lineage tracing data on hematopoiesis using scTOP confirm that a single clone can give rise to as many as three distinct differentiated cell types.
-We also show how scTOP can quantitatively assess the transcriptional similarity between endogenous and transplanted cells in the context of murine pulmonary cell transplantation.
-Finally, we provide an easy-to-use Python implementation of scTOP.
-Our results suggest that physics-inspired order parameters can be an important tool for understanding development and characterizing engineered cells.
+The abundance of data requires new theoretical and computational frameworks to analyze the dynamics of differentiation and integrate knowledge from cell atlases.
+We present single-cell Type Order Parameters (scTOP): a statistical-physics-inspired approach for quantifying cell identity given a reference basis of cell types.
+scTOP can accurately classify cells, visualize developmental trajectories, and assess the fidelity of engineered cells.
+Importantly, scTOP does this without feature selection, statistical fitting, or dimensional reduction (e.g., UMAP, PCA, etc.).
+We illustrate the power of scTOP using human and mouse datasets.
+By reanalyzing mouse lung data, we characterize a transient hybrid alveolar type 1/alveolar type 2 cell population.
+Visualizations of lineage tracing hematopoiesis data using scTOP confirm that a single clone can give rise to multiple mature cell types.
+We assess the transcriptional similarity between endogenous and donor-derived cells in the context of murine pulmonary cell transplantation.
+Our results suggest physics-inspired order parameters can be an important tool for understanding differentiation and characterizing engineered cells.
+scTOP is available as an easy-to-use Python package.
 \end{abstract}
 
-\maketitle
+\keywords{Cell types, Differentiation, Developmental trajectory, Single-cell RNA-sequencing, Computational method}
+
+\section{Summary statement}
+A computational method was developed to quantify cell identity and visualize the dynamics of differentiation.
+Concepts from statistical physics provide meaningful coordinates for comparing single-cell RNA-sequencing data across experiments.
 
 \section{Introduction}
-Humans and other mammals have many specialized cells that arise through a complex developmental process  \cite{zeng_what_2022}.
-At the blastocyst/epiblast stages of embroynic development, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers \cite{Wolpert_2019}.
+Humans and other animals have many specialized cells that arise through a complex developmental process  \citep{zeng_what_2022}.
+At the blastocyst/epiblast stages of embroynic development, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers \citep{Wolpert_2019}.
 However, as embryos mature, cellular identity becomes less plastic and more specified through a complex spatiotemporal process involving signaling and other environmental cues.
-The questions of how and why cells end up in one fate or another are fundamental to developmental biology \cite{stephenson_intercellular_2012, zhu2020principles, rand2021geometry}.
-More practically, understanding the process of cell fate specification is crucial for developing directed differentiation protocols for engineering cells \emph{in vitro}, with important clinical applications for human health and disease \cite{rossant_impact_2011, rowe2019induced, basil2020cellular}.
+The questions of how and why cells end up in one fate or another are fundamental to developmental biology \citep{stephenson_intercellular_2012, zhu2020principles, rand2021geometry}.
+More practically, understanding the process of cell fate specification is crucial for developing directed differentiation protocols for engineering cells \emph{in vitro}, with important clinical applications for human health and disease \citep{rossant_impact_2011, rowe2019induced, basil2020cellular}.
 
@@
 Typical scRNA-seq experiments involve thousands of genes and dozens to millions of cells.
-For this reason, this data is intrinsically high-dimensional and requires specialized computational tools for analysis and interpretation \cite{zhu2020principles, setty2019characterization, schiebinger2019optimal}.
-This task is especially difficult because the technical challenges of sequencing RNA from individual cells result in data that is noisy and sparse \cite{hicks2018missing, lahnemann_eleven_2020, argelaguet2021computational}.
+For this reason, this data is intrinsically high-dimensional and requires specialized computational tools for analysis and interpretation \citep{zhu2020principles, setty2019characterization, schiebinger2019optimal}.
+This task is especially difficult because the technical challenges of sequencing RNA from individual cells result in data that is noisy and sparse \citep{hicks2018missing, lahnemann_eleven_2020, argelaguet2021computational}.
 Genes with relatively low expression are often recorded as not being expressed at all, resulting in zero counts called dropouts.
@@
 scRNA-seq is often used for cellular identification.
-The most frequently used workflow involves applying dimensional reduction (such as Uniform Manifold Approximation and Projection (UMAP) \cite{mcinnes_umap_2018} or SPRING \cite{weinreb_spring_2018}) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells.
+The most frequently used workflow involves applying dimensional reduction (such as Uniform Manifold Approximation and Projection (UMAP) \citep{mcinnes_umap_2018} or SPRING \citep{weinreb_spring_2018}) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells.
 These groups are then often manually annotated according to expression levels of known marker genes.
-Machine learning classifiers such as support vector machines are also commonly used to identify cells after training with labeled data \cite{abdelaal_comparison_2019}.
-Another common analysis task is trajectory inference, where the goal is to infer developmental trajectories from static time shots (reviewed in \cite{saelens2019comparison,wagner2020lineage}).
+Machine learning classifiers such as support vector machines are also commonly used to identify cells after training with labeled data \citep{abdelaal_comparison_2019}.
+Another common analysis task is trajectory inference, where the goal is to infer developmental trajectories from static time shots (reviewed in \citep{saelens2019comparison,wagner2020lineage}).
 Trajectory inference arranges cells on a low-dimensional manifold and assigns pseudo-times according to transcriptomic similarity.
-There are also methods such as partition-based graph abstraction \cite{wolf_paga_2019} that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions.
-RNA-velocity-based methods like Dynamo \cite{qiu2022mapping} use splicing information or time-resolved data to predict paths of differentiation from RNA data.
+There are also methods such as partition-based graph abstraction \citep{wolf_paga_2019} that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions.
+RNA-velocity-based methods like Dynamo \citep{qiu2022mapping} use splicing information or time-resolved data to predict paths of differentiation from RNA data.
 
 Despite the power and utility of these computational tools, most commonly employed methods for analyzing scRNA-seq data share certain limitations.
-Dimensional reduction is often one of the first steps in visualizing scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships \cite{chari_specious_2021}.
+Dimensional reduction is often one of the first steps in visualizing scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships \citep{chari_specious_2021}.
 Distorted embeddings inconsistently represent distances between individual data points and between cell types, making it difficult to interpret cell fate transitions.
 Similar issues arise in trajectory inference and RNA-velocity methods because local neighborhoods can also vary substantially with the choice of embedding.
-Another limitation of these methods is the lack of reproducibility inherent to stochastic machine learning methods like t-SNE and UMAP, which output different results on separate executions even when hyperparameters remain the same \cite{wattenberg_how_2016}.
+Another limitation of these methods is the lack of reproducibility inherent to methods like t-SNE and UMAP, which output different results on separate executions even when hyperparameters remain the same \citep{wattenberg_how_2016}.
 Many algorithms also perform feature selection (e.g., limiting their analysis to the most variable genes or highly-expressed genes) or require choosing parameters such as perplexity, making results even more sensitive to the inclusion or exclusion of different datasets in the analysis.
@@
 First, the algorithm should incorporate explicit assumptions that can be kept consistent between analyses.
-Second, the algorithm should be deterministic, further enhancing reproducibility.
+Second, the results of the algorithm should be reproducible, so that direct comparison between data sets is possible.
 Third, the axes in visualization should be interpretable, intuitive, and remain fixed across analysis and datasets.
 Fourth, the algorithm should assign a continuous value to cell identity instead of assigning discrete identity labels, thus enabling a more subtle and fine-grained analysis of cellular identity and developmental trajectories that reflect the continuous nature of biology.
-Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single-cell atlases now being generated \cite{quake2022decade}.
+Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single-cell atlases now being generated \citep{quake2022decade}.
 
@@
 The algorithm is based on the statistical physics concept of an order parameter -- a coarse-grained quantity that can distinguish between phases of matter.
-Arguably the most famous example of an order parameter is magnetization in a ferromagnetic system (i.e., a system that can spontaneously magnetize).
-In a ferromagnet, the magnetization measures the alignment of spins and can be used to distinguish magnetic and paramagnetic phases \cite{landau2013statistical,sethna2021statistical}.
-It is straightforward to generalize the idea of magnetization to more complex settings such as spin glasses and attractor neural networks \cite{amit1985spin}.
-In the context of epigenetic landscape models of cell identity, this takes the form of generalized magnetizations that measure how aligned a gene expression state is with each of the attractor basins for different cell fates \cite{lang_epigenetic_2014,pusuluri_cellular_2017}.
-These generalized magnetizations serve as natural order parameters for cellular identity and can be calculated directly from data \cite{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}.
+An order parameter takes on different values depending on the phase of the system.
+Typically, in a disordered phase, the corresponding order parameter is zero, and in an ordered phase it is nonzero.
+Arguably the most famous example of an order parameter is magnetization in a ferromagnetic system, which is a collection of spins (in the simplest case, spins can be ``up'' or ``down'') which align with one another.
+In a ferromagnet, the magnetization measures the alignment of spins and can be used to distinguish magnetic and paramagnetic phases \citep{landau2013statistical,sethna2021statistical}.
+Below a particular temperature (known as the ``critical temperature''), the spins in the ferromagnet align with one another.
+In a simple model, the order parameter (magnetization, which measures the average spin) is approximately one if all the spins are up and approximately negative one if all the spins are down.
+When the temperature exceeds the critical temperature, the system changes from the magnetic phase to the paramagnetic phase, and the spins do not align; instead, they are randomly up or down, and the order parameter is approximately zero.
+
+It is straightforward to generalize the idea of magnetization to more complex settings such as spin glasses and attractor neural networks \citep{amit1985spin}.
+In epigenetic landscape models of cell identity based on attractor networks, this takes the form of generalized magnetizations that measure how aligned a gene expression state is with each of the attractor basins for different cell fates \citep{lang_epigenetic_2014,pusuluri_cellular_2017}.
+Whereas magnetization for the ferromagnetic system measures whether the system is in a magnetic or paramagnetic phase, these generalized magnetizations measure which phase a gene regulatory network is in (where the different phases correspond to different cell types).
+These generalized magnetizations serve as natural order parameters for cellular identity and can be calculated directly from data \citep{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}.
 The scTOP algorithm builds on previous work by developing methods for integrating scRNA-seq data, greatly expanding the power and utility of this approach.
+See Supplementary Information section \ref{interpretation} for a further discussion on interpreting score values.
 
@@
 One appealing feature of scTOP is that the assumptions of the algorithm are explicit and are contained entirely in the choice of reference basis.
-This allows direct comparison across datasets and settings since the coordinates of and distances between data points do not change as long as one uses the same reference basis.
+This allows direct comparison across data sets and settings since the coordinates of and distances between data points do not change as long as one uses the same reference basis.
 One important limitation of scTOP worth noting is that it cannot be easily used to identify novel cell fates.
@@
 We have implemented scTOP in an easy-to-use Python package, available on the \href{https://pypi.org/project/scTOP/}{Python Package Index} and \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP}{Github}.
-The code written for the analyses in this paper is available on \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP-manuscript}{Github}.
+The code for the analyses in this paper is available in a separate \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP-manuscript}{Github repository}.
 
-\section{Methods}
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.8]{figs/scTOP manuscript fig1 - no aggregate.pdf}
-	\caption{Steps involved in the scTOP algorithm.
-(a) Each cell in the query data is preprocessed independently of the other cells in the dataset.
-(b) scRNA-seq atlases, such as the Mouse Cell Atlas, are used to define the reference basis in the algorithm.
-(c) scTOP scores are the projections of query data onto the non-orthogonal subspace of cell types.
-Since there is no statistical fitting, no tuning parameters are involved.
-(d) The process of finding scTOP scores is shown in the grey section, and the blue section illustrates cell-type space.
-This space has as many dimensions as there are cell types in the reference basis.
-We can visualize this concept in 1, 2, or 3 dimensions, depending on which cell types are most relevant.
-In the 3-dimensional case, the shadows on the bounds of the scatter plot are included to better illustrate the position of points in 3-dimensional space.}
-	\label{FIG:1}
-\end{figure}
-
-\subsection{Mathematical Details}
-To calculate cell type similarity scores, scTOP uses a linear algebra projection method inspired by the concept of cell types as attractors in a dynamical system \cite{huang_cell_2005}.
-This projection method has been applied previously to bulk RNA-seq samples, and we will summarize the theoretical background here \cite{lang_epigenetic_2014}.
-A cell type is a state of the gene regulatory network.
-A natural way to represent attractors in this system is with an attractor neural network, where attractors correspond to different cell types.
-If we represent each gene as a node in a network with an associated value measuring gene expression (positive value representing high gene expression, negative value indicating low gene expression), then a cell type may be denoted by a vector of gene expression values $S_i$, where $i=1,\ldots, G$ spans the $G$ genes being measured.
-The gene expression profiles
-corresponding to the $C$ cell type attractors are also $G$ dimensional vectors $\xi_i^\mu$, where $\mu=1,\ldots, C$ spans the $C$ cell types being stored in the network.
-In what follows, we assume that $G > C$; in other words, the number of genes being measured is greater than the number of cell types.
-
-Cell types are often highly similar in their patterns of gene regulation.
-Kanter and Sompolinsky \cite{kanter_associative_1987} defined a pattern storage method for spin-glass-like neural networks where even correlated patterns robustly act as attractors.
-For the same system, they also defined order parameters: generalized magnetizations $a^{\mu}$, where $\mu$ iterates through each of the stored cell type attractors.
-The $a^\mu$ can be understood as de-correlated versions of the conventional spin glass magnetization order parameter that measures the similarity between a given network state and the network state $\xi^\mu$.
-Explicitly, the $C$ generalized magnetizations $a^\mu$ can be calculated for a gene expression state $S_i$ via the expression
-\be \label{op_eqn}
-a^\mu= \sum_{\nu=1}^C\sum_{j=1}^G [A_{\mu \nu}]^{-1}\xi_j^\nu S_j.
-\ee
-where
-\be
-A_{\mu \nu}= \sum_{j}^G  \xi_j^\mu \xi_j^\nu,
-\ee
-is the overlap matrix of gene expression profiles of different cell types.
-Previous works have shown the order parameters $a^{\mu}$ provided an excellent similarity score for analyzing bulk RNA-seq data \cite{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}.
-However, bulk RNA-seq measurements average gene expression over many cells, rendering it impossible to measure cell fate transitions beyond an average tissue state.
-
-scTOP uses the same order parameters, $a^{\mu}$, to track cell type at the resolution of individual cells, making it possible to directly observe cells in different stages of differentiation.
-In both attractor neural networks and previous applications to bulk RNA-seq data, the expression vectors $S_i$ were binary variables.
-However, we have found that for working with scRNA-seq data, it is helpful to treat $S_i$ as continuous.
-
-\subsection{Preprocessing}
-
-As shown in figure \ref{FIG:1} (a), the first stage of the algorithm is preprocessing and normalization of the input data.
-In our algorithm, measurements from each cell are normalized independently and do not depend on the expression profiles of any other cells in the dataset.
-We assume that the scRNA-seq data has been processed, resulting in a gene-wise count matrix: for each cell, each gene has an integer value that counts the number of assigned RNA reads.
-Genes are then rank-ordered.
-The rank order is subsequently converted to a z-score representing the percentile rank of the expression of a gene within the cell relative to all other genes being measured (assuming a normal distribution with mean zero and standard deviation one, for the sake of simplicity).
-For example, a gene that has higher expression than $97.8\%$ of genes being measured is assigned a score of $z=2$, while a gene that has higher expression than half of the genes is assigned a score of $z=0$.
-The details of the process and its effects on data distributions are described in Appendix section \ref{preprocessing}.
-The preprocessing step is a key component of the algorithm.
-By processing each cell independently, the output for one cell is independent of other cells included in the analysis.
-This is in contrast with other algorithms which normalize gene expression across cells by, for example, selecting the most variable genes.
-
-\subsection{Construction of reference basis}
-
-Constructing a representative, accurate reference basis for the data being analyzed is vital to the scTOP algorithm.
-This process is shown in figure \ref{FIG:1} (b).
-First, relevant single-cell RNA-seq atlases are gathered to be processed.
-For example, in analyzing mouse lung cells, we used the Mouse Cell Atlas \cite{noauthor_mapping_nodate}, which contains single-cell samples of mouse tissues across all major organs.
-Once the relevant atlases are identified, we take an average across each cell population that corresponds to a distinct cell type.
-To mitigate the potent effects of noise and avoid training data imbalances, a sufficient number of cells must be included in each population.
-We have empirically found that the minimum sample size to achieve reasonable results is 100-200 cells.
-Noise affects the algorithm most strongly when it appears in the basis since the decorrelation operation involved in the projection separates cells according to the canonical expression levels included in the reference basis.
-After each cell population is averaged to create the gene expression profiles for the desired cell types, each cell type profile is preprocessed separately using the same procedure as that used for processing single cells: normalization followed by z-scoring.
-This creates a modular reference basis where cell types can easily be replaced, removed, or added at will.
-
-\subsection{Calculation of scTOP scores}
-
-Given a reference basis, we can calculate the order parameters $a^\mu$ using Eq.~\ref{op_eqn}.
-Although these order parameters were originally inspired by attractor neural networks and spin glass physics, the $a^{\mu}$ can also be understood as non-orthogonal projections onto the subspace spanned by the reference basis (i.e., cell type space).
-This is illustrated in figure \ref{FIG:1} (c).
-A single-cell RNA-seq measurement of a sample can be thought of as a vector in $G$-dimensional gene expression space, where $G$ is the number of genes.
-The $C$-cell types in the reference basis are also each represented by a $G$-dimensional vector and form a non-orthogonal basis for the $C$-dimensional cell type space (we assume $C < G$).
-To find the coordinates $a^\mu$ in cell type space, we project the sample vector onto the reference basis using Eq.~\ref{op_eqn}.
-
-Mathematically, we can always write the sample vector as a sum of the projected components and a component $S^{\perp}_i$  perpendicular to the cell type subspace: 
-\be
-S_i = \sum_\mu a^{\mu} \xi^{\mu}_i + S^{\perp}_i.
-\ee
-Biologically, $S^{\perp}$ contains information about biological processes that introduce variation in transcriptional profile such as cell cycle as well as expression resulting from technical and biological noise.
-We refer to the $a^\mu$ as the scTOP scores.
-Each component of this vector measures the projection of the cell with gene expression profile $S_i$ onto the $\mu$-the cell type in the reference basis.
-These scores can be used to accurately classify cell identity and provide a natural visualization of gene expression in the space of possible cell fates (see figure \ref{FIG:1} (e)).
-
-In general, we are interested in calculating scTOP scores for individual cells.
-However, it is also sometimes useful to compute ``aggregate'' scTOP scores for cellular populations by averaging the mRNA counts \textit{first}, then preprocessing the data.
-This produces a gene expression profile and a set of scTOP aggregate scores for each population.
-Aggregate scores tend to be higher because averaging over the RNA counts mitigates the well-documented effects of scRNA-seq noise \cite{hicks2018missing, lahnemann_eleven_2020}; the averaged gene expression profile for the populations are not as sparse as individual profiles and thus are more similar to the averaged profiles that form the reference basis.
-
- In the appendix, we give a detailed discussion of scTOP, including the effects of basis choice, score distributions for correct and incorrect cell types, and other technical details.
-
-\subsection{Assumptions and computational complexity}
-
-scTOP does not have any stochasticity, require tuning of hyper-parameters, or statistical fitting procedures.
-The assumptions of the algorithm are explicit: that the gene expression profiles included in the reference basis are truly representative of the relevant cell types and that the reference cell types correspond to distinct cellular populations.
-We have found that constructing a reference basis for a cell type typically requires at least 100 cells.
-Additionally, if the cell types are not sufficiently different from one another, as in early embryonic development or certain stromal cells where the same cell type is compared between organs, the resulting scTOP scores may be unreliable and exhibit extreme sensitivity to small changes in the choice of reference basis.
-More information on the limitations of scTOP may be found in the appendix.
-Generally, outside these very limited edge cases, we have found that scTOP can reliably distinguish even extremely closely related cells (as shown in the example of pulmonary alveolar epithelial type 1 (AT1) and type 2 (AT2) cell lineages below).
-
-Finally, we note that scTOP is extremely computationally efficient.
-The core of the scTOP is equation \ref{op_eqn}.
-This requires a single matrix inversion involving the reference basis that can be precomputed.
-For this reason, the computational complexity of scTOP scales \emph{linearly} with the number of cells.
-The end result is that scTOP takes milliseconds to run, even for very large datasets.
 \section{Results}
-
 \subsection{Benchmarking and validation: Classification of cell fate}
@@
 Mice and humans are among the most common subjects of scRNA-seq measurements, and we restrict our analysis in this manuscript to data from these organisms.
-To do so, we make use of single-cell atlases, such as the Mouse Cell Atlas \cite{noauthor_mapping_nodate} and the Atlas of Lung Development \cite{negretti_single-cell_2021}.
+We make use of single-cell atlases such as the Mouse Cell Atlas \citep{han2018mapping} and the Atlas of Lung Development \citep{negretti_single-cell_2021}.
 
 Table \ref{table:1} lists each of the datasets examined in this paper with their respective scTOP score accuracies.
+For cases where the training (or ``reference'') data source is the same as the test data source, we set aside a randomly-selected subset of cells for the training set and do not use them in the test data.
 TopN refers to the percent of cells whose true cell types were in the top N scTOP scores, with a score greater than 0.1.
@@
 
-\begin{table}[h!]
+\begin{table*}[h!]
 \centering
@@
 \hline
-Query Data & Species & Reference Basis & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Total cells \\ 
+Test data source & Species & Reference data source & Top1 (percent) & Top3 (percent) & Unspecified (percent) & Total cells \\ 
 \hline
@@
 \label{table:1}
-\end{table}
+\end{table*}
 
 scTOP performs comparably to state-of-the-art methods for cell identification.
-Abdelaal et al.
-compared the performance of automatic cell identification algorithms for various datasets, including the CellBench dataset and an older version of the Allen Mouse Brain atlas \cite{abdelaal_comparison_2019}.
+\cite{abdelaal_comparison_2019} compared the performance of automatic cell identification algorithms for various datasets, including the CellBench dataset and an older version of the Allen Mouse Brain atlas.
 They found that the median F1-score for the CellBench dataset ranged from 0.9 to 1; scTOP has a median F1-score of 1 for the same data.
@@
 \subsubsection{Example 1: Lung lineages}
-To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from Herriges et al.
-\cite{herriges_durable_2022} onto the mouse cell atlas.
+To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from \cite{herriges_durable_2022} onto the mouse cell atlas.
 The Herriges et al.
@@
 
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.45]{figs/herriges colored by type.png}
-        \includegraphics[scale=0.5]{figs/herriges colored by signature.png}
-	\caption{scTOP identifies mouse lung cell types.
-Using the Mouse Cell Atlas as the reference basis, we show that mouse lung types separate clearly on the scTOP score axes.
-The data points correspond to individual cells, and in the first row of plots, the marker shape and color indicate the true cell type as determined by annotations from Herriges et al.
-The axes used are scTOP scores for similarity with lung ciliated cells and alveolar types 1 and 2.
-The second, third, and fourth rows of plots have individual cells colored by the marker genes for the type indicated by the color bars at the end of each row (see appendix \ref{alveolar markers} for details on marker genes used).}
-	\label{herriges}
-\end{figure}
-
 \subsubsection{Example 2: Mouse Brain Atlas}
 
-The mouse brain atlas \cite{yao_taxonomy_2021} sequenced over one million cells and clustered them into 42 subclasses and 101 supertypes.
+The mouse brain atlas \citep{yao_taxonomy_2021} sequenced over one million cells and clustered them into 42 subclasses and 101 supertypes.
 For a reference basis, we reserved a subset of 200 cells from each of the identified subclasses to use as training data.
@@
 
-\begin{figure}
-	\centering
-    \includegraphics[scale=0.75]{figs/mouse brain.png}
-	\caption{scTOP identifies mouse brain cell types.
-(a) Heatmap showing the average scTOP score for individual cells of the type indicated on the y-axis, compared to the reference types on the x-axis.
-A subset of cells from the Mouse Brain Atlas is used as the reference basis to query other cells from the same data set.
-The diagonal indicates that scTOP accurately matches query cells to the true reference type.
-(b) Histograms showing the distribution of scTOP scores for individual cells of the type labeling the x-axis.
-The scores shown are of the same type as the sample data.
-The far left histogram shows the distribution of Astro scores for Astro cells, and the average of this distribution is indicated by the color of the upper left corner box in the heatmap shown in (a).}
-	\label{mouse_brain}
-\end{figure}
-
 \subsubsection{Example 3: Matching cell types to tissues}
@@
 scTOP can also be used to match cells with tissues/organs.
-We illustrate this by comparing cells from the Mouse Cell Atlas with a reference basis consisting of Tabula Muris organs \cite{schaum_single-cell_2018}.
+We illustrate this by comparing cells from the Mouse Cell Atlas with a reference basis consisting of Tabula Muris organs \citep{schaum_single-cell_2018}.
 To do so, we averaged RNA counts across cell populations from the Mouse Cell Atlas, then preprocessed the averaged data and calculated scTOP scores.
 These aggregate pseudo-bulk scores tend to be higher than individual cell scores because averaging over many cells compensates for the dropout effect.
-As shown in figure \ref{tabula_muris}, we found the aggregate scTOP scores for cells from the Mouse Cell Atlas were correctly and strongly identified with the annotated tissue of origin in the Tabula Muris atlas.
+As shown in figure \ref{heatmaps} (a), we found the aggregate scTOP scores for cells from the Mouse Cell Atlas were correctly and strongly identified with the annotated tissue of origin in the Tabula Muris atlas.
 Parenchymal cell types were the most associated with the correct organ, while stromal cells were sometimes misidentified -- for example, the mammary gland endothelial cells.
 This suggests that stromal cells are less specified to each organ.
-A discussion of the accuracy of scTOP, when applied to stromal types versus parenchymal types, may be found in appendix \ref{failure_cases}.
-
-\begin{figure}
-	\centering
-    \includegraphics[scale=0.5]{figs/tabula muris heatmap.pdf}
-	\caption{scTOP is able to classify tissues at different resolutions.
-Aggregate pseudo-bulk samples from the Mouse Cell Atlas are compared to a reference basis consisting of organs from Tabula Muris.
-The scTOP scores are significantly higher when comparing a cell type with the organ of origin.
-The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis.
-The y-axis lists the true types of the queried cells.
-The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types.}
-	\label{tabula_muris}
-\end{figure}
+A discussion of the accuracy of scTOP, when applied to stromal types versus parenchymal types, may be found in Supplementary Information \ref{failure_cases}.
 
@@
 
-The CellBench dataset \cite{tian_benchmarking_2019} was designed specifically to benchmark scRNA-seq analysis methods.
+The CellBench dataset \citep{tian_benchmarking_2019} was designed specifically to benchmark scRNA-seq analysis methods.
 We used the subset of data pertaining to five cancer lines.
 200 cells from each line were used as training data to create the reference basis, then the rest of the cells were input as query data.
-As shown in figure \ref{cellbench} and table \ref{table:1}, scTOP classifies the cell line identity of the test data with 100 percent accuracy.
+As shown in figure \ref{heatmaps} (b) and table \ref{table:1}, scTOP classifies the cell line identity of the test data with 100 percent accuracy.
 
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.5]{figs/cellbench heatmap.pdf} 
-	\caption{scTOP correctly identifies human tissues.
-The CellBench data set contains human cell cancer lines specifically processed to benchmark scRNA-seq algorithms.
-We take a subset of cells to use as the reference basis in analyzing the rest of the CellBench data.
-The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis.
-The y-axis lists the true types of the queried cells.
-The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types.}
-	\label{cellbench}
-\end{figure}
-
-To illustrate the power of scTOP for classifying human samples, we analyzed data from the human lung atlas dataset \cite{travaglini_molecular_2020}.
+To illustrate the power of scTOP for classifying human samples, we analyzed data from the human lung atlas dataset \citep{travaglini_molecular_2020}.
 This dataset sequenced thousands of human lung cells and identified 58 phenotypic populations.
@@
 The reference basis was created using 80\% of the cells from each population, and the query data consisted of the remaining 20\% of the data.
-Figure \ref{human lung} shows that the scTOP scores were consistently high in cases where the score type matched the true type of the query.
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.5]{figs/human lung heatmap.pdf}
-	\caption{scTOP correctly identifies human tissues.
-Similar to figure \ref{cellbench}, cells from the Human Lung Atlas are compared to a reference basis constructed from cells from the Human Lung Atlas.
-The colors of the bins correspond to the average scTOP scores of individual cells.
-Again, the diagonal is the most prominent section, showing that cells of the y-axis types are correctly matched to the reference x-axis identities.}
-	\label{human lung}
-\end{figure}
+Figure \ref{heatmaps}(c) shows that the scTOP scores were consistently high in cases where the score type matched the true type of the query.
 
@@
 
-\subsection{Visualizing developmental dynamics using scTOP }
+\subsection{Visualizing cell fate dynamics using scTOP }
 Having demonstrated the robust ability of scTOP to identify stable cell types, we next demonstrate how scTOP can be used in combination with scRNA-seq time series to visualize transient cell states in dynamic processes like development and differentiation.
@@
 The alveolar epithelium of the lung undergoes dramatic maturation and morphological changes during embryonic and perinatal development, ultimately giving rise to AT1 cells and AT2 cells.
-To investigate this process with scTOP, we reanalyzed alveolar epithelial cells from a recent dataset by Zepp et al.
-\cite{zepp2021genomic}, which includes seven time points between embryonic day 12.5 (E12.5) and postnatal day 42 (P42).
-These cells were then compared against a reference basis developed from adult cell types in the Mouse Cell Atlas and an E12.5 epithelial progenitor cell type from the Atlas of Lung Development  \cite{negretti_single-cell_2021}.
+To investigate this process with scTOP, we reanalyzed alveolar epithelial cells from a recent dataset by \cite{zepp2021genomic}, which includes seven time points between embryonic day 12.5 (E12.5) and postnatal day 42 (P42).
+These cells were then compared against a reference basis developed from adult cell types in the Mouse Cell Atlas and an E12.5 epithelial progenitor cell type from the Atlas of Lung Development  \citep{negretti_single-cell_2021}.
 
-The resulting AT1 and AT2 scTOP scores were plotted to visualize maturation into these two lineages (Figure \ref{LungMAP}(a)).
+The resulting AT1 and AT2 scTOP scores were plotted to visualize maturation into these two lineages (Figure \ref{LungMAP}).
 As expected, early embryonic progenitor cells (E12.5, E15.5), which express few of the mature lineage markers, align poorly with both lineages.
@@
 In the postnatal time points, this continuous spectrum gives way to more distinct clusters, eventually resolving into AT2 and AT1 cell clusters at P42.
-Similar to our analysis of adult mouse lungs (Figure \ref{herriges}), the scTOP scores seen over development correlate well with our lineage marker gene sets (Figure \ref{LungMAP} (b)-(c)).
+Similar to our analysis of adult mouse lungs (Figure \ref{herriges}), the scTOP scores changes over the course of development correlate well with our lineage marker gene sets.
 
@@
 using a combination of graph-based clustering and gene enrichment profiling (see Figure 1C, I in \cite{zepp2021genomic}).
-The authors conjectured that these AT2/AT1 cells were a transitional state of AT2 similar to Spock2+/Axin2+ AT2 cells previously identified in adult lungs \cite{frank2016emergence}.
-To assess whether this was the case, we displayed Axin2 expression on our scTOP plots (Figure \ref{LungMAP genes}(a)).
-These cells do not have significant upregulation of Axin2 relative to perinatal or adult AT2 cells, suggesting that the postnatal hybrid AT2/AT1 cells are distinct from adult Spock2+/Axin2+ AT2 cells \cite{frank2016emergence}.
+The authors conjectured that these AT2/AT1 cells were a transitional state of AT2 similar to Spock2+/Axin2+ AT2 cells previously identified in adult lungs \citep{frank2016emergence}.
+To assess whether this was the case, we displayed Axin2 expression on our scTOP plots (Supplementary Figure \ref{LungMAP genes}(a)).
+These cells do not have significant upregulation of Axin2 relative to perinatal or adult AT2 cells, suggesting that the postnatal hybrid AT2/AT1 cells are distinct from adult Spock2+/Axin2+ AT2 cells \citep{frank2016emergence}.
 
-We then performed a similar analysis of markers for other published postnatal hybrid AT1/AT2 or transitional alveolar cell states, including alveolar differentiation intermediate (ADI, {\it Krt8} and {\it Trp53}), pre-alveolar type-1 transitional cell state (PATS, {\it Cldn4} and {\it Krt19}), and damage-associated transient progenitors (DATPs, {\it Ilr1} and {\it Hif1a}) \cite{verheyden2020transitional, strunz2020alveolar, kobayashi2020persistence, choi2020inflammatory}.
-In each case, markers were not uniquely upregulated in the perinatal hybrid AT1/AT2 cells relative to either the perinatal or adult committed lineages (Figure \ref{LungMAP genes} (b) - (d)).
+We then performed a similar analysis of markers for other published postnatal hybrid AT1/AT2 or transitional alveolar cell states, including alveolar differentiation intermediate (ADI, {\it Krt8} and {\it Trp53}), pre-alveolar type-1 transitional cell state (PATS, {\it Cldn4} and {\it Krt19}), and damage-associated transient progenitors (DATPs, {\it Ilr1} and {\it Hif1a}) \citep{verheyden2020transitional, strunz2020alveolar, kobayashi2020persistence, choi2020inflammatory}.
+In each case, markers were not uniquely upregulated in the perinatal hybrid AT1/AT2 cells relative to either the perinatal or adult committed lineages (Supplementary Figure \ref{LungMAP genes} (b) - (d)).
 Furthermore, differential gene expression analysis failed to identify any genes that were uniquely differentially expressed in the hybrid AT1/AT2 cells relative to either timepoint-matched AT1 or AT2 cells.
 These results suggest that this population is most likely transcriptionally distinct from previously identified adult AT2-transitional cell types.
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.52]{figs/LungMAP.png}
-	\caption{Visualizing the developing mouse lung with scTOP.
-scTOP illustrates the specification of alveolar type I and type II cells in murine embryonic development.
-Alveolar cells from Zepp et al.
-are compared to a reference basis constructed from adult lung cells from Herriges et al.
-and early epithelial cells from the Atlas of Lung Development.
-(a) Each subplot corresponds to the age of the mouse from which the sample was extracted, from post-conception day 12,5 (E12.5) to 42 days post-birth (P42).
-(b), (c) Scatter plots of the same data as in (a), now with each cell colored by the average over that cell's expression of (b) AT1 marker genes and (c) AT2 marker genes.
-The scTOP scores of AT1 (AT2) cells are high when the average AT1 (AT2) marker gene expression is high.
-(d) The same data as above is displayed as a 2D histogram, with each bin colored corresponding to the average day of the data point(s) falling within that bin.
-The color trend shows that early cells generally have very low AT1 and AT2 scores, then the AT1 and AT2 scores increase over time.}
-	\label{LungMAP}
-\end{figure}
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.8]{figs/LungMAP genes.png}
-	\caption{The AT1/AT2 combination state (most apparent in days P3 and P7) do not uniquely express genes of AT1/AT2 transitional states described by previous papers.
-Alveolar cells from Zepp et al.
-are plotted on AT1 and AT2 axes, separated by day.
-Each cell is colored by the expression z-score of a particular gene, and the particular gene is different in each row.
-The gene used to color the cells is indicated by the color bar on the far right of each row.
-(a) Alveolar cells colored by Axin2, showing that the AT1/AT2 combination cells do not appear to belong to the Axin2+ sublineage described by Frank et al.
-\cite{frank2016emergence}.
-(b) Alveolar cells colored by Krt8 and Trp53 expression, which are genes corresponding to alveolar differentiation intermediate cells (ADI) described by Strunz et al.
-\cite{strunz2020alveolar}.
-AT1, AT2, and the combination AT1/AT2 state all express Krt8 and Trp53 at similar levels.
-(c) Alveolar cells colored by expression of pre-alveolar type-1 transitional cell state (PATS) genes Kobayashi et al.
-\cite{kobayashi2020persistence}.
-AT1, AT2, and the combination AT1/AT2 state all express Cldn4 at similar levels.
-AT1 and AT1/AT2 combination cells both express Krt19 at similar levels, although AT1 cells express the gene at higher levels.
-(d) Alveolar cells colored by expression of damage-associated transient progenitor (DATP) genes described by Choi et al.
-\cite{choi2020inflammatory}.
-AT1, AT2, and the combination AT1/AT2 state all express Il1r1 and HIF1a at similar levels.
-}
-	\label{LungMAP genes}
-\end{figure}
 
@@
 In this section, we show how scTOP can also be used to gain insights into more complicated developmental processes such as hematopoiesis, in which hematopoietic stem cells (HSCs) give rise to multiple differentiated blood cell types.
-To do so, we make use of the unique dataset by Weinreb et al.
-\cite{weinreb_lineage_2020} on hematopoiesis in mice using a new technique, lineage and RNA recovery (LARRY), that combines scRNA-seq and lineage tracing data.
+To do so, we make use of the unique dataset by \cite{weinreb_lineage_2020} on hematopoiesis in mice using a new technique, lineage and RNA recovery (LARRY), that combines scRNA-seq and lineage tracing data.
 LARRY uses heritable barcodes that can be detected using scRNA sequencing, allowing for the tracking of clone families during differentiation.
-Here, we focus on visualizing the ex-vivo experiments in \cite{weinreb_lineage_2020}, where cells were sequenced at a sufficient depth to infer lineage relationships.
+Here, we focus on visualizing the ex-vivo experiments from Weinreb et al., where cells were sequenced at a sufficient depth to infer lineage relationships.
 In these experiments, HSCs were extracted from mice, barcoded, and allowed to expand in primary culture.
@@
 We restrict our analysis to a subset of these annotated cell types with at least 150 cells.
-We also excluded cells labeled as erythroids from our analysis since we found that this population was highly heterogeneous and gave rise to poor projections (likely due to the fact it was annotated using a single marker gene, \textit{Hbb-bs} \cite{weinreb_lineage_2020}, which did not sufficiently define a distinct population).
+We also excluded cells labeled as erythroids from our analysis since we found that this population was highly heterogeneous and gave rise to poor projections (likely due to the fact it was annotated using a single marker gene, \textit{Hbb-bs} \citep{weinreb_lineage_2020}, which did not sufficiently define a distinct population).
 Furthermore, we treated all day 2 undifferentiated cells as a single population we called progenitors.
@@
 Interestingly, the day 4 cells move in the direction of both of these lineages before bifurcating at day 6.
-This data is consistent with Weinreb et al.
-\cite{weinreb_lineage_2020}, who previously highlighted the existence of such a bipotent neutrophil-monocyte progenitor population.
+This data is consistent with Weinreb et al., who previously highlighted the existence of such a bipotent neutrophil-monocyte progenitor population.
 We also find evidence that progenitor cells can even give rise to three distinct cellular populations.
@@
 
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.75]{figs/hem. clone families.png}
-	\caption{Tracing hematopoietic cell differentiation with scTOP.
-(a), (b), (c) display scatter plots and distributions of scTOP scores for individual in vitro clone families from  \cite{weinreb_lineage_2020}.
-The scatter plots show the x-axis scTOP score compared to the y-axis scTOP score for individual cells, while the distributions show the scTOP scores of the type indicated on the x-axis of the corresponding column.
-The color and shape of each marker (and the color of each distribution) indicate the days in culture on which the cells were extracted.
-(a) A clone family where all of the cells end up becoming neutrophils.
-(b) A clone family where some clones end up becoming neutrophils and some become monocytes.
-At day 2, the 3 sister cells have low neutrophil and monocyte scores and high progenitor scores.
-At days 4 and 6, the progenitor scores decrease, and a portion of the cells increase in monocyte scores while others increase in neutrophil scores.
-(c) A clone family where the descendant cells become neutrophils, eosinophils, and basophils.}
-	\label{hem clones}
-\end{figure}
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.75]{figs/hem progenitor hists.png}
-	\caption{Tracing differentiation of hematopoietic cells from progenitor to mature type.
-The plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al., with the x-axis of each one indicating the scTOP score for hematopoietic progenitor type and the y-axis showing the score for a mature type.
-Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin.
-Early cells have high progenitor scores and low mature-type scores.
-As time passes, cells end up with high scores in mature types.}
-	\label{hem progenitor}
-\end{figure}
-
-\begin{figure}
-	\centering
-		\includegraphics[scale=0.59]{figs/hem mature hists.png}
-	\caption{Comparing differentiation of hematopoietic cells between mature types.
-As in figure \ref{hem progenitor}, the plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al.
-The x-axis and y-axis of each histogram indicate the scTOP score of the labeled mature type.
-Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin.  Some pairs of types have cells with high scores of both types.
-This is apparent in the neutrophil-monocyte histogram, which presents a continuous spectrum of scores from the horizontal axis to the vertical.
-Other pairs of types only have cells expressing one type or the other, as in the L-shaped neutrophil-megakaryocyte histogram.}
-	\label{hem mature}
-\end{figure}
-
 \subsection{Using scTOP to compare endogenous and transplanted cells}
 
-\begin{figure}[h!]
-	\centering
-		\includegraphics[scale=0.83]{figs/engineered cells comparison.png}[h!]
-	\caption{Donor-derived and endogenous alveolar cells from Herriges et al.
-and Louie et al.
-The cell type annotations come from a separate Louvain clustering and are the same for all plots in this figure.
-(a) UMAP visualizations of the data, colored by cell type annotations.
-The three have different values for the UMAP parameter n\_neighbors, which changes whether the embedding preserves more local or global structure.
-(b) scTOP plot showing the AT1 score on the x-axis and the AT2 score on the y-axis.
-(c) Box-and-whisker plot showing the distributions of scTOP AT2 scores for the various cell populations.
-(d) Bar plot showing the top ten aggregate scores for the AT2 populations from different sources, sorted by the scores for the SCA1+ population.}
-	\label{engineered}
-\end{figure}
 Recent advancements in cell culture and cell transplantation techniques have allowed researchers to generate cells that were maintained or differentiated in vitro to approximate and eventually replace endogenous cells generated during in vivo development.
@@
 
-To compare two distinct pulmonary cell transplant protocols, we used scTOP to analyze the results of scRNA-seq from two recent publications featuring murine pulmonary cell transplantation \cite{herriges_durable_2022, louie_progenitor_2022}.
+To compare two distinct pulmonary cell transplant protocols, we used scTOP to analyze the results of scRNA-seq from two recent publications featuring murine pulmonary cell transplantation \citep{herriges_durable_2022, louie_progenitor_2022}.
 In Louie et al., adult mouse AT2 cells were maintained in culture with support cells for 3 weeks to generate a donor cell population.
@@
 
-Figure \ref{engineered} (d) provides a comparison between cell populations using aggregate scTOP scores.
-Similar to figure \ref{engineered} (c), (d) provides a quantitative comparison between the populations.
+Supplementary figure \ref{barplot} provides a comparison between cell populations using aggregate scTOP scores.
+Similar to figure \ref{engineered} (c), figure \ref{barplot} provides a quantitative comparison between the populations.
 This figure confirms that while both AT2-derived and PSC-derived transplants generate cells that have high AT2 scores, the AT2-derived cells are quantitatively more similar to the reference AT2 profile.
@@
 scTOP is especially suited for tasks where the cell types of interests are well characterized.
-For example, we have found in our own work that scTOP is an extremely sensitive measure for assessing the fidelity of engineered cells from directed differentiation protocols and reprogramming \cite{herriges_durable_2022}.
+For example, we have found in our own work that scTOP is a sensitive measure for assessing the fidelity of engineered cells from directed differentiation protocols and reprogramming \citep{herriges_durable_2022}.
 Since scTOP does not require data harmonization, joint embeddings, or make use of cellular neighborhood information,  scTOP is able to distinguish between technical noise and biologically meaningful differences.
@@
 
-As an illustration, we analyzed the development of the lung epithelial AT1 and AT2 cell fates \cite{zepp2021genomic} and developmental dynamics during hematopoiesis  \cite{weinreb_lineage_2020}.
+As an illustration, we analyzed the development of the lung epithelial AT1 and AT2 cell fates \citep{zepp2021genomic} and developmental dynamics during hematopoiesis  \citep{weinreb_lineage_2020}.
 Using scTOP, we showed that it was easy to identify a transient hybrid AT2/AT1 population 3 to 5 days post-birth whose gene expression profile is distinct from previously investigated transient AT2-to-AT1 states.
@@
 However, in principle, our approach can be easily extended to include other data modalities, such as chromatic accessibility and histone modifications.
-Recently, several interesting works such as Dynamo have used RNA-velocity to directly learn cellular dynamics \cite{xing2022reconstructing, qiu2022mapping}.
+Recently, several interesting works such as Dynamo have used RNA-velocity to directly learn cellular dynamics \citep{xing2022reconstructing, qiu2022mapping}.
 It will be interesting to see if these methods can be combined with scTOP to learn vector flows directly in cell-type space.
-Similarly, methods based on optimal transport, such as Waddington-OT \cite{schiebinger2019optimal}, or developmental directory reconstruction based on graph embedding, such as Monocle \cite{qiu2017reversed}, can also be performed in the cellular identity space instead of gene expression space.
+Similarly, methods based on optimal transport, such as Waddington-OT \citep{schiebinger2019optimal}, or developmental directory reconstruction based on graph embedding, such as Monocle \citep{qiu2017reversed}, can also be performed in the cellular identity space instead of gene expression space.
 These represent promising directions for future research since the scTOP cell identity coordinates are often better suited for analyzing developmental dynamics than gene expression coordinates.
@@
 
-\section*{Acknowledgments}
-We acknowledge useful discussions with Jason Rocks, Robert Marsland, and Alex Lang.
+\section{Materials and methods}
+
+\subsection{Mathematical Details}
+To calculate cell type similarity scores, scTOP uses a linear algebra projection method inspired by the concept of cell types as attractors in a dynamical system \citep{huang_cell_2005}.
+This projection method has been applied previously to bulk RNA-seq samples, and we will summarize the theoretical background here \citep{lang_epigenetic_2014}.
+A cell type is a state of the gene regulatory network.
+A natural way to represent attractors in this system is with an attractor neural network, where attractors correspond to different cell types.
+If we represent each gene as a node in a network with an associated value measuring gene expression (positive value representing high gene expression, negative value indicating low gene expression), then a cell type may be denoted by a vector of gene expression values $S_i$, where $i=1,\ldots, G$ spans the $G$ genes being measured.
+The gene expression profiles
+corresponding to the $C$ cell type attractors are also $G$ dimensional vectors $\xi_i^\mu$, where $\mu=1,\ldots, C$ spans the $C$ cell types being stored in the network.
+In what follows, we assume that $G > C$; in other words, the number of genes being measured is greater than the number of cell types.
+
+Cell types are often highly similar in their patterns of gene regulation.
+\cite{kanter_associative_1987} defined a pattern storage method for spin-glass-like neural networks where even correlated patterns robustly act as attractors.
+For the same system, they also defined order parameters: generalized magnetizations $a^{\mu}$, where $\mu$ iterates through each of the stored cell type attractors.
+The $a^\mu$ can be understood as de-correlated versions of the conventional spin glass magnetization order parameter that measures the similarity between a given network state and the network state $\xi^\mu$.
+We refer to $\xi$ as the reference basis.
+Explicitly, the $C$ generalized magnetizations $a^\mu$ can be calculated for a gene expression state $S_i$ via the expression
+\be \label{op_eqn}
+a^\mu= \sum_{\nu=1}^C\sum_{j=1}^G [A_{\mu \nu}]^{-1}\xi_j^\nu S_j.
+\ee
+where
+\be
+A_{\mu \nu}= \sum_{j}^G  \xi_j^\mu \xi_j^\nu,
+\ee
+is the overlap matrix of gene expression profiles of different cell types.
+Previous works have shown the order parameters $a^{\mu}$ provided an excellent similarity score for analyzing bulk RNA-seq data \citep{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}.
+However, bulk RNA-seq measurements average gene expression over many cells, rendering it impossible to measure cell fate transitions beyond an average tissue state.
+
+scTOP uses the same order parameters, $a^{\mu}$, to track cell type at the resolution of individual cells, making it possible to directly observe cells in different stages of differentiation.
+In both attractor neural networks and previous applications to bulk RNA-seq data, the expression vectors $S_i$ were binary variables.
+However, we have found that for working with scRNA-seq data, it is helpful to treat $S_i$ as continuous.
+
+\subsection{Preprocessing}
+
+As shown in figure \ref{FIG:1} (a), the first stage of the algorithm is preprocessing and normalization of the input data.
+In our algorithm, measurements from each cell are normalized independently and do not depend on the expression profiles of any other cells in the dataset.
+We assume that the scRNA-seq data has been processed, resulting in a gene-wise count matrix: for each cell, each gene has an integer value that counts the number of assigned RNA reads.
+Genes are then rank-ordered.
+The rank order is subsequently converted to a z-score representing the percentile rank of the expression of a gene within the cell relative to all other genes being measured (assuming a normal distribution with mean zero and standard deviation one, for the sake of simplicity).
+For example, a gene that has higher expression than $97.8\%$ of genes being measured is assigned a score of $z=2$, while a gene that has higher expression than half of the genes is assigned a score of $z=0$.
+The details of the process and its effects on data distributions are described in Supplementary Information section \ref{preprocessing}.
+The preprocessing step is a key component of the algorithm.
+By processing each cell independently, the output for one cell is independent of other cells included in the analysis.
+This is in contrast with other algorithms which normalize gene expression across cells by, for example, selecting the most variable genes.
+
+\subsection{Construction of reference basis}
+
+Constructing a representative, accurate reference basis for the data being analyzed is vital to the scTOP algorithm.
+This process is shown in figure \ref{FIG:1} (b).
+First, relevant single-cell RNA-seq atlases are gathered to be processed.
+For example, in analyzing mouse lung cells, we used the Mouse Cell Atlas \citep{han2018mapping}, which contains single-cell samples of mouse tissues across all major organs.
+Once the relevant atlases are identified, we take an average across each cell population that corresponds to a distinct cell type.
+To mitigate the potent effects of noise and avoid training data imbalances, a sufficient number of cells must be included in each population.
+We have empirically found that the minimum sample size to achieve reasonable results is 100-200 cells.
+Noise affects the algorithm most strongly when it appears in the basis since the decorrelation operation involved in the projection separates cells according to the canonical expression levels included in the reference basis.
+After each cell population is averaged to create the gene expression profiles for the desired cell types, each cell type profile is preprocessed separately using the same procedure as that used for processing single cells: normalization followed by z-scoring.
+This creates a modular reference basis where cell types can easily be replaced, removed, or added at will.
+
+\subsection{Calculation of scTOP scores}
+
+Given a reference basis, we can calculate the order parameters $a^\mu$ using Eq.~\ref{op_eqn}.
+Although these order parameters were originally inspired by attractor neural networks and spin glass physics, the $a^{\mu}$ can also be understood as non-orthogonal projections onto the subspace spanned by the reference basis (i.e., cell type space).
+This is illustrated in figure \ref{FIG:1} (c).
+A single-cell RNA-seq measurement of a sample can be thought of as a vector in $G$-dimensional gene expression space, where $G$ is the number of genes.
+The $C$-cell types in the reference basis are also each represented by a $G$-dimensional vector and form a non-orthogonal basis for the $C$-dimensional cell type space (we assume $C < G$).
+To find the coordinates $a^\mu$ in cell type space, we project the sample vector onto the reference basis using Eq.~\ref{op_eqn}.
+
+Mathematically, we can write the sample vector as a sum of the projected components and a component $S^{\perp}_i$  perpendicular to the cell type subspace: 
+\be
+S_i = \sum_\mu a^{\mu} \xi^{\mu}_i + S^{\perp}_i.
+\ee
+Biologically, $S^{\perp}$ contains information about processes that affect gene expression but are not associated with cell identity, such as the cell cycle.
+We refer to the $a^{\mu}$ as the scTOP scores.
+Each component of this vector measures the projection of the cell with gene expression profile $S_i$ onto the $\mu$ - the cell type in the reference basis.
+These scores can be used to accurately classify cell identity and provide a natural visualization of gene expression in the space of possible cell fates (see figure \ref{FIG:1} (e)).
+Gene regulatory dynamics that are unrelated to transitions between cell types in the reference basis would not be captured by $a^{\mu}$; instead, these would contribute to $S^{\perp}$.
+As a result, these fate-unrelated dynamics cannot be visualized using scTOP.
+
+In general, we are interested in calculating scTOP scores for individual cells.
+However, it is also sometimes useful to compute ``aggregate'' scTOP scores for cellular populations by averaging the mRNA counts \textit{first}, then preprocessing the data.
+This produces a gene expression profile and a set of scTOP aggregate scores for each population.
+Aggregate scores tend to be higher because averaging over the RNA counts mitigates the well-documented effects of scRNA-seq noise \citep{hicks2018missing, lahnemann_eleven_2020}; the averaged gene expression profile for the populations are not as sparse as individual profiles and thus are more similar to the averaged profiles that form the reference basis.
+
+ In the Supplementary Information, we give a detailed discussion of scTOP, including the effects of basis choice, score distributions for correct and incorrect cell types, and other technical details.
+
+\subsection{Assumptions and computational complexity}
+
+scTOP does not require tuning of hyper-parameters or statistical fitting procedures.
+The assumptions of the algorithm are explicit: that the gene expression profiles included in the reference basis are truly representative of the relevant cell types and that the reference cell types correspond to distinct cellular populations.
+We have found that constructing a reference basis for a cell type typically requires at least 100 cells.
+Additionally, if the cell types are not sufficiently different from one another, as in early embryonic development or certain stromal cells where the same cell type is compared between organs, the resulting scTOP scores may be unreliable and exhibit extreme sensitivity to small changes in the choice of reference basis.
+More information on the limitations of scTOP may be found in the Supplementary Information.
+Generally, outside these very limited edge cases, we have found that scTOP can reliably distinguish even extremely closely related cells (as shown in the example of pulmonary alveolar epithelial type 1 (AT1) and type 2 (AT2) cell lineages below).
+
+Finally, we note that scTOP is extremely computationally efficient.
+The core of the scTOP is equation \ref{op_eqn}.
+This requires a single matrix inversion involving the reference basis that can be precomputed.
+For this reason, the computational complexity of scTOP scales \emph{linearly} with the number of cells.
+The end result is that scTOP takes milliseconds to run, even for very large datasets.
+
+\vfill\eject
+
+%%%%%%%%%%% Please use the respective coding for Back matter section %%%%%%%%
+
+\ack{We acknowledge useful discussions with Jason Rocks, Robert Marsland, and Alex Lang.
 We would like to thank the Mehta group and Kotton lab for their comments on the manuscript.
 The work was funded by a grant to the Boston University Kilachand Multicellular Design Program (to DK and PM) and NIH NIGMS 1R35GM119461 to PM.
-We also acknowledge programming support from Alena Yampolskaya and Sumner Hearth.
+We also acknowledge programming support from Alena Yampolskaya and Sumner Hearth.}
 
-\section*{Appendix}
-\beginsupplement
-\subsection{Preprocessing} \label{preprocessing}
-The raw scRNA-seq count data for each dataset was downloaded as a matrix of mRNA counts corresponding to genes and cells.
-An example of such a matrix is shown in step 0 of figure \ref{FIG:supp1}, where each row is a different gene, and each column is a different cell.
-Throughout this figure, the histogram to the right of each step shows the distribution of entry values for the first column (cell) in the matrix, and the histogram below the step shows the distribution of entry values for the second row (gene Rpl7) of the matrix.
-The histograms are presented as examples of how the distributions (for each cell and gene) of gene expressions change as the data is preprocessed.
+\competing{Insert the Competing interests text here.}
 
-In the first step of preprocessing, each cell is normalized independently.
-This step is taken because scRNA-seq generally measures relative rather than absolute expression levels within a cell.
-In order to facilitate the comparison of cells across sequencing conditions, we transform the normalized RNA counts into quantities that are centered around the mean expression level.
-Z-scores are exactly such a quantity; they measure the number of standard deviations away from the mean.
-To calculate the z-scores after step 1, we convert the normalized gene expressions into percentiles by ranking each gene in a given cell from lowest to highest, then dividing by the number of genes (plus one so that the highest rank is not assigned the 100th percentile).
-For example, if a cell has 1000 genes sampled and Rpl7 is the highest-expressing gene, it will be given rank 1000 and labeled as the 99th percentile.
-If two or more genes are tied for rank, as is the case for zero-values, they are assigned the average of the ranks they would have been assigned if they were not tied (see rankdata function from the Python package Scipy).
-To convert from percentiles to z-scores, we assumed a normal distribution with mean 0 and standard deviation 1 and applied the corresponding quantile function.
+\data{scTOP is available as a package on the Python Package Index (https://pypi.org/project/scTOP/) and the code is accessible via Github (https://github.com/Emergent-Behaviors-in-Biology/scTOP).
+The scripts with the analyses for this paper are also available on Github (https://github.com/Emergent-Behaviors-in-Biology/scTOP-manuscript).
+The scRNA-seq data analyzed can be downloaded from a range of sources.
+The endogenous and transplanted mouse lung data from the Kotton lab is available on NCBI's Gene Expression Omnibus under accession codes GSE200886, GSE200883, GSE200884, and GSE200885, as well as the Kotton Lab's Bioinformatics Portal (http://www.kottonlab.com) \citep{herriges_durable_2022}.
+The Mouse Cell Atlas data is available on FigShare (https://figshare.com/articles/dataset/MCA\_DGE\_Data/5435866) \citep{han2018mapping}.
+The mouse brain atlas 10x data is available on the Neuroscience Multi-omic Data Archive (https://assets.nemoarchive.org/dat-jb2f34y) \citep{yao_taxonomy_2021}.
+The CellBench data is available under accession GSE118767 and on Github (https://github.com/LuyiTian/sc\_mixology/blob/master/data/) \citep{tian_benchmarking_2019}.
+The human lung atlas data is available on Synapse (https://www.synapse.org/\#!Synapse:syn21041850) as well as the European Genome-phenome Archive under accession EGAS00001004344 \citep{travaglini_molecular_2020}.
+Tabula Muris data is available through FigShare (https://figshare.com/projects/Tabula\_Muris\_Transcriptomic\_characterization\_of\_20\_organs\_and\_tissues\_from\_Mus\_musculus\_at\_single\_cell\_resolution/27733) and accession GSE109774 \citep{schaum_single-cell_2018}.
+The developing mouse lung data is available under accession GSE149563 \citep{zepp2021genomic}.
+The hematopoietic lineage tracing data is available under accession GSE140802, and the metadata can be found on Github (https://github.com/AllonKleinLab/paper-data/tree/master/Lineage\_tracing\_on\_transcriptional\_landscapes\_links\_state\_to\_fate\_during\_differentiation) \citep{weinreb_lineage_2020}.
+The endogenous and engineered mouse lung data from the Kim lab is available under accession GSE190565 \citep{louie_progenitor_2022}.
+The human pancreatic data used for the batch effects figure in the SI was downloaded from Docker (https://hub.docker.com/repository/docker/jinmiaochenlab/batch-effect-removal-benchmarking) \citep{tran2020benchmark}, and it's available under accessions GSE85241 \citep{baron2016single}, E-MTAB-5061 \citep{muraro2016single}, GSE84133
+\citep{segerstolpe2016single}, GSE83139 \citep{wang2016single}, and GSE81608 \citep{xin2016rna}.}
 
-\begin{figure}
+%%%%%%%%% References %%%%%%%%%%%%%%%%%
+\newcommand{\newblock}{}
+\bibliography{export-data.bib}
+
+\begin{figure}[h!]
 	\centering
-		\includegraphics[scale=0.6]{figs/fig1a supplement.pdf}
-	\caption{Data distributions at each preprocessing step.
-Step 0 shows the original RNA count matrix before preprocessing, where each entry corresponds to the number of RNA detected for a particular gene in a particular cell.
-The rows of each matrix correspond to genes, and the columns correspond to cells.
-The histograms to the right of each matrix show the distribution of an example cell, while the histograms below each matrix show the distribution of an example gene.
-Step 1 of preprocessing is to normalize each cell.
-Steps 2 and 3 convert the normalized gene expressions to z-scores.
-Step 2 ranks the genes in each cell according to magnitude and converts the rank into a percentile.
-Step 3 uses a quantile function, assuming a normal distribution with mean zero and standard deviation one, to convert from percentiles to z-scores.}
-	\label{FIG:supp1}
+	\caption{\textbf{Steps involved in the scTOP algorithm.} (a) Each cell in the query data is preprocessed independently of the other cells in the dataset.
+(b) scRNA-seq atlases, such as the Mouse Cell Atlas, are used to define the reference basis in the algorithm.
+(c) scTOP scores are the projections of query data onto the non-orthogonal subspace of cell types.
+Since there is no statistical fitting, no tuning parameters are involved.
+(d) The process of finding scTOP scores is shown in the grey section, and the blue section illustrates cell-type space.
+This space has as many dimensions as there are cell types in the reference basis.
+We can visualize this concept in 1, 2, or 3 dimensions, depending on which cell types are most relevant.
+In the 3-dimensional case, the shadows on the bounds of the scatter plot are included to better illustrate the position of points in 3-dimensional space.}
+	\label{FIG:1}
 \end{figure}
 
-\subsection{Reference basis construction}\label{basis construction}
-Creating an appropriate reference basis is vital to the accuracy of scTOP.
-There are many existing scRNA-seq atlases, such as the Mouse Cell Atlas; using them with scTOP requires some amount of curation.
-For each reference dataset, we took an average over each population corresponding to a given cell type.
-The average gene expression profiles were preprocessed and then used as the reference profiles for each cell type.
-Often, it was necessary to perform additional curation of the reference basis.
-Cell types were dropped from an atlas if they did not include enough cells (i.e., if they contained $<100$ cells) or were found to be poorly defined (i.e., not enough marker genes to distinguish that cell type from others).
-Common instances of the latter were cell types that differed only in the expression of a single gene.
-
-Ensuring that each reference gene expression profile sampled enough cells was essential.
-This is because the number of cells determines how well-sampled the reference cell type is.
-Although cell types have distinct transcriptomic patterns, they are not all transcriptomically identical.
-There are variations in gene expression levels between cells of the same type due to differences in cell-level processes such as cell division or environmental stimuli.
-Averaging over a population of cells which are some cell type X creates an approximate gene expression profile of the archetypal X cell.
-
-\subsubsection{Mouse Cell Atlas}
-The Mouse Cell Atlas (MCA) contains hundreds of thousands of cells from tissues across the mouse body at various stages of development.
-For the reference basis created from the Mouse Cell Atlas (MCA), we dropped cell types that had fewer than 100 cells.
-For cell types that had greater than or equal to 100 cells, we averaged across the entire population for each one, then preprocessed.
-For cell types that were divided according to high expressions of various genes, we combined them into one cell type.
-For example, the original MCA contained dozens of clusters of mammary gland secretory alveoli cells such as ``secretory alveoli cell, Hes1 high" or ``secretory alveoli cell, Gpx3 high." These were all combined into one cell type, ``mammary gland secretory alveoli." This is because we wanted each reference basis type to act as an archetype rather than a perfect representation of every possible iteration of a cell type; each reference type represents a basin of attraction, and the basins were defined by the overall cell type, not whether individual genes were particularly high within that cell type.
-We cleaned each of the MCA type labels and ended up with cell types defined by the organ of origin, specific cell type, and time point of collection (e.g., ``lung ciliated cell week 6-10").
-The time points of cell types in the MCA were either E14.5 or somewhere between week 6 and week 10 post-birth.
-The resulting reference basis contained 221 cell types.
-
-\subsection{Accuracy measures}
-For table \ref{table:1}, the top1 accuracy was determined by applying scTOP to a cell, finding the type with the highest score, and checking whether that type matched the cell's true type.
-Top3 was similarly determined, except instead of just checking the highest-scoring type, we checked whether the true type was included in the top 3 highest-scoring types.
-
-By looking at the scTOP scores for cells whose true type did not match the reference type, we set a score of 0.1 as the lower bound for identifying a type.
-As shown in figure \ref{robustness hists} (c) and (f), when the reference type does not match the sample type, the scores for individual cells fall well within [-0.1, 0.1], even when most of the reference basis is missing.
-
-In table \ref{table:1}, we labeled cells as unspecified if the highest scTOP score was less than 0.1.
-This provides a useful metric for determining whether a query cell belongs to one of the cell types in the reference basis.
-If the highest scTOP score is lower than 0.1 for a query cell, this indicates that the reference basis lacks the relevant types or was poorly constructed (see appendix \ref{failure_cases} for a discussion of ill-defined bases).
-
-scTOP can only identify cells that exist within the reference basis.
-As such, the accuracy scores were calculated only for the cells whose true type (as determined by the annotations from the authors) was included in the reference basis for the corresponding analysis.
-For cells whose true types were not represented in the basis, the unspecified rate was very high.
-This indicates that the rate of false positives is very low.
-
-\subsection{Robustness of scTOP}
-scTOP is consistent between iterations of the algorithm, and it is robust to changes in the reference basis.
-As shown in figure \ref{robustness hists}, many scTOP scores do not change significantly even when most of the reference basis is removed.
-Generally, scores for cells that match the reference type do not change much even when only 25\% of the original basis is retained (figure \ref{robustness hists} (d), (e)).
-The largest difference occurs for cells that are similar to the reference type, such as the AT1 score for AT2 cells (figure \ref{robustness hists} (g)).
-These scores tend to increase because removing other types from the basis reduces the ability to de-correlate effectively.
-In other words, scTOP is more likely to confuse similar types when it has fewer reference types to compare.
-This effect is most apparent when relevant types are removed from the basis.
-Figures \ref{robustness hists} (j) and (k) show that all lung types score higher for AT1 and AT2 when only lung alveolar types are included in the basis.
-Figures \ref{robustness hists} (c) and (f) show that irrelevant scores, such as bladder and thymus cells, do not change much even when large portions of the basis are removed.
-scTOP does not confuse sample cells for types that are dissimilar even with a small reference basis because there is no need to de-correlate between types that are naturally not correlated.
-
-\subsubsection{Cases where it does not work}\label{failure_cases}
-The performance of scTOP relies on the quality of the reference basis.
-scTOP is expected to give inaccurate results in cases where the reference basis is ill-defined.
-The underlying assumption of the algorithm is that the reference basis consists of accurate gene expression profiles of cell types that are attractor basins.
-In other words, the reference types are assumed to represent stable cell types rather than cell states that are on the path to the true cell type basins.
-For example, scTOP performs poorly when using a reference basis consisting of early embryonic identities because the cell types are not sufficiently differentiated from one another.
-Adapting scTOP to such early lineage data is an important and interesting problem.
-
-The other part of the assumption is that the reference types should be accurate profiles of the cell type of interest.
-As discussed in appendix \ref{basis construction}, reference types are created from averaging over cells, and the more cells included in this average, the better-sampled the type is.
-As expected, scTOP gives unreliable results when the reference gene expression levels do not accurately reflect the true cell types' gene expression levels.
-An important factor in the accuracy of scRNA-seq is the amplification step: after RNA is extracted, it is amplified to improve the sensitivity of sequencing, but not all RNA is amplified at the same rate.
-This results in inflated counts for some genes but not others.
-Unique molecular identifiers (UMIs) are vital for reducing the inflating effect of amplification \cite{islam2014quantitative}.
-UMIs tag the RNA before amplification and are present post-amplification; this makes it possible to see which amplified RNA came from the same original molecule and prevents uneven double-counting of RNA.
-We have found that scTOP gives more accurate results when used with reference and query datasets that use UMIs, compared to scTOP results using data that did not use UMIs.
-
-\begin{figure}
+\begin{figure}[h!]
 	\centering
-		\includegraphics[scale=0.72]{figs/hem genes.png}
-	\caption{Scatter plots of in vitro hematopoietic clone families from Weinreb et al., with scTOP scores for progenitor, neutrophil, and monocyte types, and colored by normalized gene expression.
-In Weinreb et al.
-figure 4 (a), cells visualized using pseudo-time inference were separated into branches.
-The monocyte branch appeared to have half of its cells expressing dendritic cell genes and the other half expressing neutrophil genes.
-This figure plots those same cells on progenitor, neutrophil, and monocyte axes.
-The colors correspond to neutrophil, dendritic cell, and monocyte marker genes.
-(a) Hematopoietic cells colored according to the expression of neutrophil marker genes S100a8 and Elane.
-(b) Hematopoietic cells colored according to the expression of dendritic cell marker genes Cd74 and H2-Aa.
-(c) Hematopoietic cells colored according to the expression of monocyte marker genes Csf1r and Klf4. }
-	\label{hem genes}
+	\caption{\textbf{scTOP identifies mouse lung cell types.} Using the Mouse Cell Atlas as the reference basis, we show that mouse lung types separate clearly on the scTOP score axes.
+The data points correspond to individual cells, and in the first row of plots, the marker shape and color indicate the true cell type as determined by annotations from Herriges et al.
+The axes used are scTOP scores for similarity with lung ciliated cells and alveolar types 1 and 2.
+The second, third, and fourth rows of plots have individual cells colored by the marker genes for the type indicated by the color bars at the end of each row (see Supplementary Information \ref{alveolar markers} for details on marker genes used).}
+	\label{herriges}
 \end{figure}
 
-\begin{figure}
+\begin{figure}[h!]
 	\centering
-		\includegraphics[scale=0.82]{figs/robustness hists.png}
-	\caption{Distributions of scTOP scores for lung cells from Herriges et al., using varying portions of the Mouse Cell Atlas as the basis.
-The label above each plot indicates the Herriges et al.
-sample subset, while the x-axis label indicates the scTOP reference type.
-"All lung types" means that all of the lung types from the data set were included.
-(a) - (i) show how scTOP scores change when 25-100\% of the basis is used.
-(j) - (k) show how scTOP scores change when the basis is changed from the full basis (which includes many tissues and organs) to just the lung types and to just the two alveolar types.
-(c) and (f) show scTOP scores for random irrelevant types, Bladder Epithelium and Thymus B Cell, to show that the scTOP scores remain low for types that do not match the sample.}
-	\label{robustness hists}
+	\caption{\textbf{scTOP identifies mouse brain cell types.} (a) Heatmap showing the average scTOP score for individual cells of the type indicated on the y-axis, compared to the reference types on the x-axis.
+A subset of cells from the Mouse Brain Atlas is used as the reference basis to query other cells from the same data set.
+The diagonal indicates that scTOP accurately matches query cells to the true reference type.
+(b) Histograms showing the distribution of scTOP scores for individual cells of the type labeling the x-axis.
+The scores shown are of the same type as the sample data.
+The far left histogram shows the distribution of Astro scores for Astro cells, and the average of this distribution is indicated by the color of the upper left corner box in the heatmap shown in (a).}
+	\label{mouse_brain}
 \end{figure}
 
-\subsection{AT1/AT2 marker genes}\label{alveolar markers}
-In figure \ref{LungMAP}, plots in (b) and (c) were colored according to AT1 and AT2 marker gene expression.
-For each cell, the average was taken over the expressions of a set of marker genes.
-The AT1 marker gene set was:
+\begin{figure}[h!]
+	\centering
+	\caption{\textbf{scTOP correctly classifies tissues across biological contexts.} The shade of each bin indicates the average scTOP score for individually-queried cells projected on the reference cell type indicated on the x-axis.
+The y-axis lists the true types of the queried cells.
+The dark diagonals demonstrate that scTOP scores correctly match the queried data to the corresponding true types.
+(a) Aggregate pseudo-bulk samples from the Mouse Cell Atlas are compared to a reference basis consisting of organs from Tabula Muris.
+The scTOP scores are significantly higher when comparing a cell type with the organ of origin.
+(b) The CellBench data set contains human cell cancer lines specifically processed to benchmark scRNA-seq algorithms.
+We take a subset of cells to use as the reference basis in analyzing the rest of the CellBench data.
+(c) Cells from the Human Lung Atlas are compared to a reference basis constructed from cells from the Human Lung Atlas.
+}
+	\label{heatmaps}
+\end{figure}
 
-\begin{itemize}
-    \item Ager
-    \item Akap5
-    \item Aqp5
-    \item Cldn18
-    \item Clic3
-    \item Clic5
-    \item Col4a3
-    \item Col4a4
-    \item Cryab
-    \item Cyp2b10
-    \item Emp2
-    \item Fam189a2
-    \item Gprc5a
-    \item Hopx
-    \item Hs2st1
-    \item Igfbp2
-    \item Krt7
-    \item Lmo7
-    \item Mal2
-    \item Pdpn
-    \item Prdx6
-    \item Pxdc1
-    \item Rtkn2
-    \item Scnn1g
-    \item Spock2
-    \item Tspan8
-    \item Vegfa
-\end{itemize}
+\begin{figure}[h!]
+	\centering
+	\caption{\textbf{Visualizing the developing mouse lung with scTOP.
+scTOP illustrates the specification of alveolar type I and type II cells in murine embryonic development.} Alveolar cells from Zepp et al.
+are compared to a reference basis constructed from adult lung cells from Herriges et al.
+and early epithelial cells from the Atlas of Lung Development.
+Each subplot corresponds to the age of the mouse from which the sample was extracted, from post-conception day 12.5 (E12.5) to 42 days post-birth (P42).
+Each cell is colored by the average of that cell's expression of (b) AT1 marker genes and (c) AT2 marker genes.
+The scTOP scores of AT1 (AT2) cells are high when the average AT1 (AT2) marker gene expression is high.}
+	\label{LungMAP}
+\end{figure}
 
-The AT2 marker gene set was:
+\begin{figure}[h!]
+	\centering
+	\caption{\textbf{Visualizing hematopoietic lineage tracing with scTOP.} (a), (b), (c) display scatter plots and distributions of scTOP scores for individual in vitro clone families from  \cite{weinreb_lineage_2020}.
+The scatter plots show the x-axis scTOP score compared to the y-axis scTOP score for individual cells, while the distributions show the scTOP scores of the type indicated on the x-axis of the corresponding column.
+The color and shape of each marker (and the color of each distribution) indicate the days in culture on which the cells were extracted.
+(a) A clone family where all of the cells end up becoming neutrophils.
+(b) A clone family where some clones end up becoming neutrophils and some become monocytes.
+At day 2, the 3 sister cells have low neutrophil and monocyte scores and high progenitor scores.
+At days 4 and 6, the progenitor scores decrease, and a portion of the cells increase in monocyte scores while others increase in neutrophil scores.
+(c) A clone family where the descendant cells become neutrophils, eosinophils, and basophils.}
+	\label{hem clones}
+\end{figure}
 
-\begin{itemize}
-    \item Abca3
-    \item Acot7
-    \item Acoxl
-    \item Acsl4
-    \item Ank3
-    \item Atp8a1
-    \item Cd74
-    \item Cebpa
-    \item Cpm
-    \item Ctsh
-    \item Cxcl15
-    \item Dram1
-    \item Egfl6
-    \item Elovl1
-    \item Etv5
-    \item Fabp5
-    \item Fasn
-    \item Hc
-    \item Lamp3
-    \item Lgi3
-    \item Lpcat1
-    \item Lyz2
-    \item Mlc1
-    \item Muc1
-    \item Napsa
-    \item Npc2
-    \item Ppp1r14c
-    \item S100g
-    \item Scd1
-    \item Sfta2
-    \item Sftpa1
-    \item Sftpb
-    \item Sftpc
-    \item Sftpd
-    \item Slc34a2
-    \item Zdhhc3
-\end{itemize}
+\begin{figure}[h!]
+	\centering
+	\caption{\textbf{Tracing differentiation of hematopoietic cells from progenitor to mature type.} The plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al., with the x-axis of each one indicating the scTOP score for hematopoietic progenitor type and the y-axis showing the score for a mature type.
+Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin.
+Early cells have high progenitor scores and low mature-type scores.
+As time passes, cells end up with high scores in mature types.}
+	\label{hem progenitor}
+\end{figure}
 
-\bibliography{export-data.bib}
+\begin{figure}[h!]
+	\centering
+	\caption{\textbf{Comparing differentiation of hematopoietic cells between mature types.} As in figure \ref{hem progenitor}, the plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al.
+The x-axis and y-axis of each histogram indicate the scTOP score of the labeled mature type.
+Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin.
+Some pairs of types have cells with high scores of both types.
+This is apparent in the neutrophil-monocyte histogram, which presents a continuous spectrum of scores from the horizontal axis to the vertical.
+Other pairs of types only have cells expressing one type or the other, as in the L-shaped neutrophil-megakaryocyte histogram.}
+	\label{hem mature}
+\end{figure}
+
+\begin{figure}[h!]
+	\centering
+	\caption{\textbf{Comparing donor-derived and endogenous alveolar cells from Herriges et al.
+and Louie et al.} The cell type annotations come from a separate Louvain clustering and are the same for all plots in this figure.
+(a) UMAP visualizations of the data, colored by cell type annotations.
+The three have different values for the UMAP parameter n\_neighbors, which changes whether the embedding preserves more local or global structure.
+(b) scTOP plot showing the AT1 score on the x-axis and the AT2 score on the y-axis.
+(c) Box-and-whisker plot showing the distributions of scTOP AT2 scores for the various cell populations.
+Supplementary figure \ref{barplot} shows the scores for other types are consistently low.}
+	\label{engineered}
+\end{figure}
 
